---
title: "AhR-RNAseq-patient-analysis"
author: "Adam Sorbie"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of RNAseq AHR 

Load libraries with PACMAN
```{r}
pacman::p_load("DESeq2", "GenomicFeatures", "org.Hs.eg.db", "AnnotationDbi", "tidyverse",
               "ComplexHeatmap", "apeglm", "RColorBrewer", "EnhancedVolcano", "viridis",
               "readxl", "xlsx", "circlize", "ggpubr")
source("../scripts/RNAseq_functions.R")
```

### Data Loading and Preprocessing

First, we load the metadata and expression data, then filter for the samples of interest.

```{r}
# Load metadata
metadata <- read.csv("../data/meta.csv")

# Display the metadata structure
str(metadata)

# Filter metadata for LPS stimulated and male samples
metadata_all <- metadata %>% 
   filter(stimulated == "LPS" & gender == "m")
metadata_24h <- metadata %>% 
  filter(stimulated == "LPS" & gender == "m" & time != "72h")

metadata_72h <- metadata %>% 
  filter(stimulated == "LPS" & gender == "m" & time != "24h")

# Load count data - assuming this is already processed and normalized
counts <- read_csv("../data/rawCounts.csv") %>% 
  select(-gene_name) %>% 
  column_to_rownames("gene_id")

# Subset counts to only include samples from filtered metadata
counts_24h <- counts[, colnames(counts) %in% metadata_24h$sample_id]
counts_72h <- counts[, colnames(counts) %in% metadata_72h$sample_id]
counts_all <- counts[, colnames(counts) %in% metadata_all$sample_id]
```

### Create DESeq2 Objects

Now we create DESeq2 objects for both time points.

```{r}
# Create colData dataframes with matching sample order
coldata_24h <- metadata_24h %>%
  column_to_rownames("sample_id") %>%
  .[colnames(counts_24h), ] %>% 
  mutate(group = factor(group, levels = c("control", "stroke")))

coldata_72h <- metadata_72h %>%
  column_to_rownames("sample_id") %>%
  .[colnames(counts_72h), ] %>% 
  mutate(group = factor(group, levels = c("control", "stroke")))

# Create DESeq2 objects
dds_24h <- DESeqDataSetFromMatrix(
  countData = counts_24h,
  colData = coldata_24h,
  design = ~ group
)

dds_72h <- DESeqDataSetFromMatrix(
  countData = counts_72h,
  colData = coldata_72h,
  design = ~ group
)

# Set reference level for group variable
dds_24h$group <- relevel(dds_24h$group, ref = "control")
dds_72h$group <- relevel(dds_72h$group, ref = "control")

# Filter low count genes
keep_24h <- rowSums(counts(dds_24h)) >= 10
dds_24h <- dds_24h[keep_24h,]

keep_72h <- rowSums(counts(dds_72h)) >= 10
dds_72h <- dds_72h[keep_72h,]

## all timepoints 
coldata_all <- metadata_all %>%
  column_to_rownames("sample_id") %>%
  .[colnames(counts_all), ] %>% 
  mutate(group = factor(group, levels = c("control", "stroke")))

# Create DESeq2 objects
dds_all <- DESeqDataSetFromMatrix(
  countData = counts_all,
  colData = coldata_all,
  design = ~ group
)

# Set reference level for group variable
dds_24h$group <- relevel(dds_24h$group, ref = "control")
dds_72h$group <- relevel(dds_72h$group, ref = "control")
dds_all$group <- relevel(dds_all$group, ref = "control")

# Filter low count genes
keep_all <- rowSums(counts(dds_all)) >= 10
dds_all <- dds_all[keep_24h,]
```

## Principal Component Analysis

Let's examine the overall variation in our datasets using PCA.

### PCA of all male samples by group (24h)

```{r}
# Variance stabilizing transformation
vst_24h <- vst(dds_24h, blind = TRUE)

# Extract PCA data
pca_data_24h <- plotPCA(vst_24h, intgroup = "group", returnData = TRUE)

# Create improved PCA plot
ggplot(pca_data_24h, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 3) +
  labs(x = paste0("PC1 [", round(100 * attr(pca_data_24h, "percentVar")[1], 1), "%]"),
       y = paste0("PC2 [", round(100 * attr(pca_data_24h, "percentVar")[2], 1), "%]"),
       color = "Group") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

### PCA of all samples by group (72h)

```{r}
# Variance stabilizing transformation
vst_72h <- vst(dds_72h, blind = TRUE)

# Extract PCA data
pca_data_72h <- plotPCA(vst_72h, intgroup = "group", returnData = TRUE)

# Create improved PCA plot
ggplot(pca_data_72h, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 3) +
  labs(x = paste0("PC1 [", round(100 * attr(pca_data_72h, "percentVar")[1], 1), "%]"),
       y = paste0("PC2 [", round(100 * attr(pca_data_72h, "percentVar")[2], 1), "%]"),
       color = "Group") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

### PCA of stroke samples colored by NIHSS score (24h)

```{r}
# Filter for stroke samples only
stroke_samples_24h <- colData(vst_24h) %>%
  as.data.frame() %>%
  filter(group == "stroke")

# Subset VST data to include only stroke samples
vst_stroke_24h <- vst_24h[, rownames(stroke_samples_24h)]

# PCA on stroke samples only
pca_stroke_24h <- prcomp(t(assay(vst_stroke_24h)))

# Create data frame for plotting
pca_df_stroke_24h <- as.data.frame(pca_stroke_24h$x) %>%
  rownames_to_column("SampleID") %>%
  left_join(stroke_samples_24h %>% rownames_to_column("SampleID"), by = "SampleID")

# Plot PCA colored by NIHSS score
ggplot(pca_df_stroke_24h, aes(x = PC1, y = PC2, color = NIHSS)) +
  geom_point(size = 4) +
  scale_color_viridis_c(direction = -1) +  # Yellow (high) to purple (low)
  labs(x = paste0("PC1 [", round(100 * attr(pca_df_stroke_24h, "percentVar")[1], 1), "%]"),
       y = paste0("PC2 [", round(100 * attr(pca_df_stroke_24h, "percentVar")[2], 1), "%]"),
       color = "NIHSS score") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

### PCA of stroke samples colored by NIHSS score (72h)

```{r}
# Filter for stroke samples only
stroke_samples_72h <- colData(vst_72h) %>%
  as.data.frame() %>%
  filter(group == "stroke")

# Subset VST data to include only stroke samples
vst_stroke_72h <- vst_72h[, rownames(stroke_samples_72h)]

# PCA on stroke samples only
pca_stroke_72h <- prcomp(t(assay(vst_stroke_72h)))

# Create data frame for plotting
pca_df_stroke_72h <- as.data.frame(pca_stroke_72h$x) %>%
  rownames_to_column("SampleID") %>%
  left_join(stroke_samples_72h %>% rownames_to_column("SampleID"), by = "SampleID")

# Plot PCA colored by NIHSS score
ggplot(pca_df_stroke_72h, aes(x = PC1, y = PC2, color = NIHSS)) +
  geom_point(size = 4) +
  scale_color_viridis_c(direction = -1) +  # Yellow (high) to purple (low)
  labs(x = paste0("PC1 [", round(100 * attr(pca_df_stroke_72h, "percentVar")[1], 1), "%]"),
       y = paste0("PC2 [", round(100 * attr(pca_df_stroke_72h, "percentVar")[2], 1), "%]"),
       color = "NIHSS score") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

## PCA all timepoints 

```{r}
# Variance stabilizing transformation
vst_all <- vst(dds_all, blind = TRUE)
# Extract PCA data
pca_data_all <- plotPCA(vst_all, intgroup = c("group","time"), returnData = TRUE)

# Create improved PCA plot
ggplot(pca_data_all, aes(x = PC1, y = PC2, color = group.1, shape=time)) +
  geom_point(size = 3) +
  labs(x = paste0("PC1 [", round(100 * attr(pca_data_all, "percentVar")[1], 1), "%]"),
       y = paste0("PC2 [", round(100 * attr(pca_data_all, "percentVar")[2], 1), "%]"),
       color = "Group") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```


```{r}
# Filter for stroke samples only
stroke_samples <- colData(vst_all) %>%
  as.data.frame() %>%
  filter(group == "stroke")

# Subset VST data to include only stroke samples
vst_stroke_all <- vst_all[, rownames(stroke_samples)]

# PCA on stroke samples only
pca_stroke_all <- prcomp(t(assay(vst_stroke_all)))

# Create data frame for plotting
pca_stroke_all_df <- as.data.frame(pca_stroke_all$x) %>%
  rownames_to_column("SampleID") %>%
  left_join(stroke_samples %>% rownames_to_column("SampleID"), by = "SampleID")

percentVar <- pca_stroke_all$sdev^2 / sum(pca_stroke_all$sdev^2)

# Plot PCA colored by NIHSS score
ggplot(pca_stroke_all_df, aes(x = PC1, y = PC2, color = NIHSS, shape=time)) +
  geom_point(size = 4) +
  scale_color_viridis_c(direction = 1) +  # Yellow (high) to purple (low)
  labs(x = paste0("PC1 [", round(100 * percentVar[1], 1), "%]"),
       y = paste0("PC2 [", round(100 * percentVar[2], 1), "%]"),
       color = "NIHSS score") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
ggsave("PCA_stroke_samples_all-2026.pdf", width = 5, height = 3.5, dpi = 300)
```


## Differential Expression Analysis

Now we perform differential expression analysis to identify genes that are differentially expressed between stroke and control samples.

```{r}
# Run DESeq2 analysis
dds_24h <- DESeq(dds_24h)
res_24h <- results(dds_24h, contrast = c("group", "stroke", "control"), 
                   independentFiltering = TRUE, alpha = 0.05)

dds_72h <- DESeq(dds_72h)
res_72h <- results(dds_72h, contrast = c("group", "stroke", "control"), 
                   independentFiltering = TRUE, alpha = 0.05)

dds_all <- DESeq(dds_all)
res_all <- results(dds_all, contrast = c("group", "stroke", "control"), 
                   independentFiltering = TRUE, alpha = 0.05)

# Annotate results
res_24h_annot <- annotate_degs(org.Hs.eg.db, res_24h, keys = rownames(res_24h), 
                               keytype = "ENSEMBL", multivals = "first")
res_24h_df <- as.data.frame(res_24h_annot)

res_72h_annot <- annotate_degs(org.Hs.eg.db, res_72h, keys = rownames(res_72h), 
                               keytype = "ENSEMBL", multivals = "first")
res_72h_df <- as.data.frame(res_72h_annot)

res_all_annot <- annotate_degs(org.Hs.eg.db, res_all, keys = rownames(res_all), 
                               keytype = "ENSEMBL", multivals = "first")
res_all_df <- as.data.frame(res_all_annot)

# Summary of results
summary(res_24h, alpha = 0.05)
summary(res_72h, alpha = 0.05)
summary(res_all, alpha = 0.05)
```

### Volcano Plot (24h)

```{r}
prettyVolcano(
  res_24h_df,
  lfc_thresh = 1,
  pval_thresh = 0.1,
  contrast = c("control", "stroke"),
  gene_names = "gene_symbol",
  xlim = c(-10, 10),
  ylim = c(0, 4),
  downreg_col = "#6D6D6F",
  upreg_col = "#FFA845",
  pointSize = 3,
  drawConnectors = TRUE,
  max.overlaps = 20
)
```

### Volcano Plot (72h)

```{r}
prettyVolcano(
  res_72h_df,
  lfc_thresh = 1,
  pval_thresh = 0.1,
  contrast = c("control", "stroke"),
  gene_names = "gene_symbol",
  xlim = c(-10, 10),
  ylim = c(0, 4),
  downreg_col = "#6D6D6F",
  upreg_col = "#FFA845",
  pointSize = 3,
  drawConnectors = TRUE,
  max.overlaps = 20
)
```

```{r}
# write.xlsx(res_24h_df, file="log2FC_stroke_monocytes.xlsx", sheetName="controlvsstroke24h", row.names=FALSE)
# write.xlsx(res_72h_df, file="log2FC_stroke_monocytes_72.xlsx", sheetName="controlvsstroke72h", append=TRUE, row.names=FALSE)
# write.xlsx(res_all_df, file="log2FC_stroke_monocytes_all.xlsx", sheetName="controlvsstrokealltp", append=TRUE, row.names=FALSE)
```


## Heatmap of Gene Panel

Now we'll create a heatmap using the gene panel from the provided Excel file, focused only on stroke samples and ordered by NIHSS score.

```{r}
# Load gene panel
ahr_gene_panel <- read_excel("../data/mmc2 (4).xlsx")
ahr_gene_panel$Gene <- str_to_sentence(ahr_gene_panel$Gene)  # Convert to sentence case

# metadata
stroke_samples <- colData(vst_stroke_all) %>%
  as.data.frame() %>%
  filter(group == "stroke") %>%
  arrange(NIHSS)

# Convert Entrez to Ensembl
ensembl_ids <- mapIds(org.Hs.eg.db,
                      keys = as.vector(as.character(ahr_gene_panel$`Entrez ID`)),
                      column = "ENSEMBL",
                      keytype = "ENTREZID",
                      multiVals = "first")

detected_ahr <- ensembl_ids[ensembl_ids %in%  rownames(vst_stroke_all)]

# Extract data for panel genes
vst_ahr_panel <- vst_stroke_all[detected_ahr, ]

# Set gene symbols as row names
rownames(vst_ahr_panel) <- res_all_df %>%
  rownames_to_column("ensembl") %>% 
  filter(ensembl %in% detected_ahr) %>%
  pull(gene_symbol)

# Calculate Z-scores
vst_panel_z <- t(scale(t(assay(vst_ahr_panel))))

# Create annotation data
ha <- HeatmapAnnotation(
  Time = stroke_samples$time,
  NIHSS = stroke_samples$NIHSS,
  col = list(
    Time = c("24h" = "#3498db", "72h" = "#e74c3c"),
    NIHSS = colorRamp2(c(min(stroke_samples$NIHSS), 
                        mean(stroke_samples$NIHSS), 
                        max(stroke_samples$NIHSS)), 
                      viridis(3, direction = -1))
  ),
  show_legend = TRUE
)

# Create the heatmap with k-means clustering
Heatmap(vst_panel_z, 
        name = "Z-score",
        top_annotation = ha,
        column_split = stroke_samples$time,
        cluster_rows = TRUE,
        cluster_columns = FALSE,  # Columns already ordered by NIHSS
        column_title = "Stroke Patients",
        row_title = "Gene Panel",
        km = 5,  # k-means with 3 clusters
        column_order = rownames(stroke_samples),
        row_names_gp = gpar(fontsize = 5.5),
        heatmap_legend_param = list(title = "Z-score\nexpression"))
```
Correlations 

```{r}
key_ahr_genes <- c("AHR", "CYP1A1", "CYP1B1", "AHRR")
df_ahr_genes <- assay(vst_ahr_panel)[key_ahr_genes, ]
df_ahr_genes <- df_ahr_genes %>% 
  as.data.frame() %>% 
  rownames_to_column("gene_symbol") %>%
  pivot_longer(-(gene_symbol), names_to = "sample_id", values_to = "expression") 

# add metadata 
df_ahr_genes <- df_ahr_genes %>%
  left_join(metadata_all %>% select(sample_id, NIHSS, time), by = "sample_id")
```

Plot gene expression vs NIHSS score stratified by time

```{r}
for (i in key_ahr_genes){
  sub <- df_ahr_genes %>% 
    filter(gene_symbol == i)
  p <- ggscatter(sub, x = "NIHSS", y = "expression", 
                            color = "time", facet.by = "time",
                            add = "reg.line", conf.int = T,
                            add.params = list(color = "black", fill = "lightgray"),
                            xlab = "NIHSS Score", ylab = paste0(i, " Expression")) +
    stat_cor(method = "pearson") +
    ggtitle(paste0("Correlation of ", i, " with NIHSS Score"))
  print(p)
}
```


# Heatmap of AhR target genes

```{r}
ahr_target_genes <-readxl::read_excel("../data/mmc2 (4).xlsx")
ahr_target_genes$Gene <- str_to_sentence(ahr_target_genes$Gene)
```

```{r}
ahr_target_genes_ensembl <- res_df %>% 
  filter(gene_symbol %in% ahr_target_genes$Gene) %>% 
  rownames()
rld_mat_ahr <- rld_mat[ahr_target_genes_ensembl, ]
rownames(rld_mat_ahr) <- res_df %>% 
  filter(gene_symbol %in% ahr_target_genes$Gene) %>% 
  pull(gene_symbol)
```

```{r}
# scale matrix and plot
rld_mat_ahr_z <- calc_z_score(rld_mat_ahr)
# some have zero-variance so replace with 0 (need to check why this is)
rld_mat_ahr_z[is.na(rld_mat_ahr_z)] <- 0
Heatmap(rld_mat_ahr_z, top_annotation = hmap_annot, column_split = groups_hmap, border = "black", 
        cluster_column_slices = F)
```

## Deconvolution 


```{r}
library(immunedeconv)

```

```{r}
counts_deconv <- annotate_degs(org.Hs.eg.db, counts_all, keys = rownames(counts_all), 
                               keytype = "ENSEMBL", multivals = "first")

```


```{r}
counts_deconv <- counts_deconv %>% 
  drop_na(gene_symbol) %>%
  rownames_to_column("ensembl") %>%
  distinct(gene_symbol, .keep_all = TRUE) %>%
  select(-c(ensembl, entrez, `gene name`)) %>%
  column_to_rownames("gene_symbol")

```


```{r}
deconv_results <- deconvolute(as.matrix(counts_deconv), method = "quantiseq")
```


```{r}
deconv_results_long <- pivot_longer(deconv_results, cols = -cell_type, names_to = "sample_id", values_to = "proportion") %>%
  left_join(metadata_all %>% select(sample_id, group, time), by = "sample_id") %>% 
  filter(group == "stroke") %>% 
  group_by(cell_type, time) %>%
  mutate(mean_prop = mean(proportion) * 100) %>% 
  ungroup
write.csv(deconv_results_long, "deconvolution_results.csv", row.names = FALSE)
```


