---
title: "AHR paper"
author: "Adam Sorbie"
date: "07/11/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# AHR paper figures

This document contains the analysis of ileal and plasma untargeted
metabolomics from Sham and stroke mice using the tMCAO model (citation).

Methods:

"Untargeted metabolite profiling from ileum fecal content and portal
vein (plasma) was performed by Metabolon (Durham, NC). Briefly, samples
were analyzed with a Q-Exactive high resolution/accuratemass
spectrometry (MS) interfaced with a heated ESIsource and Orbitrap mass
analyzer (Thermo Scientific) coupled with UPLC and consisted of 2
separated (RP)/UPLC-MS/MS methods with positive ESI mode, one with
negative ESI mode, and hydrophilic interaction liquid chromatography
[HILIC])/UPLC-MS/MS with negative ESI mode. Curation was performed by
Metabolon (Evans et al., 2014)."

## Libraries

Load required libraries for analysis and source functions.

```{r, include=FALSE}
pacman::p_load(ropls,
               ggforce,
               caret,
               cowplot,
               ggpubr,
               POMA,
               gghighlight,
               ggside)
# load data as metabo list 
source("../scripts/data_processing.R")
source("../scripts/analysis_functions.R")
conflicted::conflict_prefer_all("dplyr")
conflicted::conflicts_prefer(purrr::map)
```

## Functions

Functions, specifically for use in this document.

```{r}
scatter_theme <- theme(
  axis.title=element_text(size=26),
  axis.text=element_text(size=22),
  legend.title=element_text(size=26),
  legend.text=element_text(size=22),
  strip.text = element_text(size=22)
)
```

## Load data

Load unscaled features and metadata. Compound ID is used as rowname
here.

```{r}
il_feat <- read_csv("../data/metabolites_unscaled_ileal_content.csv") %>% 
  column_to_rownames(var="COMP ID")
il_meta <- read_csv("../data/metadata_ileal_content.csv")
pl_feat <- read_csv("../data/metabolites_unscaled_plasma.csv") %>% 
  column_to_rownames(var="COMP ID")
pl_meta <- read_csv("../data/metadata_plasma.csv")
```

## Data inspection

Convert to metabo list object. Imported function creates a list of
features, sample metadata as well as feature metadata. This keeps the
environment from getting to cluttered and keepy everything associated
with a particular experiment in one place.

```{r}
il_metabo <- metabo_list(il_feat, il_meta, "SampleID")
pl_metabo <- metabo_list(pl_feat, pl_meta, "SampleID")
```

NOTE - currently issue with I3S - up in fMCAO in untargeted but down in targeted (ileum). Check normalisation and imputation 

```{r}
i3s_raw <- il_metabo$features["27672", ]
i3s_raw %>%
  pivot_longer(everything(), values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram()
```

Plot by group 
```{r}
t(i3s_raw) %>% 
  as.data.frame() %>% 
  merge(il_meta %>% column_to_rownames("SampleID"), by=0) %>% 
  #mutate(`27672`)
  ggboxplot(y="`27672`", x="group", add="jitter")
```

```{r}
il_meta %>% 
  ggboxplot(y="sample_weight", x="group", add="jitter")
```

## Filtering, imputation and normalization

Here metabolite profiles are filtered removing those present in less
than 75% of samples (in this case 14). Missing values are then
subsequently imputed using half minimum imputation.

Test how different filtering parameters effect resulting metabolite
matrix

```{r}
# ILEUM
n_samples <- seq(1, 18, 1)
map(n_samples, function(x) il_metabo %>% 
      filter_prevalence(filt_type = "min_samples", max_na = x)) %>% 
  map(function(x) dim(x$features)[[1]])
```

Remove those which are missing in 75% or more sameples

```{r}
il_metabo <- il_metabo %>% 
  filter_prevalence(filt_type = "min_samples", max_na = 14) %>% 
  half_min_imp()
print( paste0( dim(il_feat)[[1]] - dim(il_metabo$features)[[1]], " features removed after filtering") ) 
```

```{r}
pl_metabo <- pl_metabo %>% 
  filter_prevalence(filt_type = "min_samples", max_na = 14) %>% 
  half_min_imp()
print( paste0( dim(pl_feat)[[1]] - dim(pl_metabo$features)[[1]], " features removed after filtering") ) 
```

Normalisation using quantile normalisation

Normalisation is performed using probabilistic quantile normalisation
(Dieterle et al 2006, Analytical Chemistry).

Sample weight normalisation also necessary? This needs to be checked, not clearly an issue based on i3s plots above but some of the 
highest values do tend to have higher weight

```{r}
il_metabo_norm <- pqn(il_metabo)
pl_metabo_norm <- pqn(pl_metabo)
```

Check normalisation

Ileum:

```{r}
unnorm <- il_metabo$features %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("SampleID") %>% 
  pivot_longer(-SampleID) %>% 
  ggboxplot(x="SampleID", y="value")

norm <- il_metabo_norm$norm.features %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("SampleID") %>% 
  pivot_longer(-SampleID) %>% 
  ggboxplot(x="SampleID", y="value")

cowplot::plot_grid(unnorm, norm, ncol = 2, nrow = 1)
ggsave("figures/ileum_normalisation.pdf", device = "pdf", dpi=300)
```

Plasma:

```{r}
unnorm <- pl_metabo$features %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("SampleID") %>% 
  pivot_longer(-SampleID) %>% 
  ggboxplot(x="SampleID", y="value")

norm <- pl_metabo_norm$norm.features %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("SampleID") %>% 
  pivot_longer(-SampleID) %>% 
  ggboxplot(x="SampleID", y="value")

cowplot::plot_grid(unnorm, norm, ncol = 2, nrow = 1)
ggsave("figures/plasma_normalisation.pdf", device = "pdf", dpi=300)
```

### Scaling

Finally, data are transformed to log2 (+ pseudocount) and scaled using
Z-score scaling.

```{r}
il_metabo_norm$log.features <- log2(il_metabo_norm$norm.features + 1e-10)
il_metabo_norm$scaled.features <- scale_z_score(il_metabo_norm$log.features)
```

```{r}
pl_metabo_norm$log.features <- log2(pl_metabo_norm$norm.features + 1e-10)
pl_metabo_norm$scaled.features <- scale_z_score(pl_metabo_norm$log.features)
```

Save processed data as Rdata object.

```{r}
save(il_metabo_norm, pl_metabo_norm, file="Processed_metabolome_data.Rdata")
```

## Outlier detection {.tabset}

Outlier detection here is performed using POMA (Castellano-Escuder et
al, 2021, Plos Comp. Bio) which employs a method based on euclidean
distance and Q3+1.5 ∗ IQR by default.

```{r, warning=FALSE}
outliers_ileum <- detect_outliers(il_metabo_norm)
rmarkdown::paged_table(outliers_ileum)
```

```{r, warning=FALSE}
outliers_plasma <- detect_outliers(pl_metabo_norm)
rmarkdown::paged_table(outliers_plasma)
```

## Analysis

```{r}
# load preprocessed data to save time
#load("Processed_metabolome_data.Rdata")
```

### PCA

PCA is performed using prcomp function and ggplot2 for visualisation.

#### Ileum PCA

```{r}
il_pca <- prcomp(t(il_metabo_norm$scaled.features)) 
il_ev <- il_pca$sdev^2/sum(il_pca$sdev^2)
```

Stat test PC1 PC2 

```{r}
stroke_samples <- il_metabo$sample.metadata %>% filter(group == "Stroke") %>% pull(SampleID)
sham_samples <- il_metabo$sample.metadata %>% filter(group == "Sham") %>%  pull(SampleID)

pc1_wilcox <- wilcox.test(il_pca$x[stroke_samples,"PC1"], il_pca$x[sham_samples,"PC1"])
pc2_wilcox <- wilcox.test(il_pca$x[stroke_samples,"PC2"], il_pca$x[sham_samples,"PC2"])
```


```{r}
ileum_cols <- c("#414042","#FF9300")
as.data.frame(il_pca$x)%>%
  add_column(Group=il_metabo_norm$sample.metadata$group,
             Infarct_volume=il_metabo_norm$sample.metadata$indir_inf_vol) %>%
  ggplot(aes(x=PC1, y=PC2, colour=Group, fill=Group, shape=Group)) +
  
  theme_bw() +
  theme(panel.grid = element_blank()) + 
  scatter_theme +
  geom_point(size=4) +
  scale_color_manual(values = ileum_cols) +
  scale_fill_manual(values= ileum_cols) + 
  xlab(paste0("PC", 1, " (", round(il_ev[1], 3)*100, "%)")) +
  ylab(paste0("PC", 2, " (", round(il_ev[2], 3)*100, "%)")) + 
  stat_ellipse(geom = "polygon", alpha=0.1)  + 
  geom_ysideboxplot(
    aes(fill = Group, y = PC2),
    color = "black",
    orientation = "x",
  ) +
  geom_xsideboxplot(
    aes(fill = Group, y = PC1),
    color = "black",
    orientation = "y"
  )
ggsave("figures/PCA_ileum.pdf", device = "pdf", dpi=300, height = 8, width = 10)
```

PCA Biplot

```{r}
factoextra::fviz_pca_biplot(il_pca, repel = TRUE, select.var = list(contrib = 25)) + 
  scatter_theme
```

#### Plasma PCA

```{r}
pl_pca <- prcomp(t(pl_metabo_norm$scaled.features)) 
pl_ev <- pl_pca$sdev^2/sum(pl_pca$sdev^2)
```

Stat test PC1 PC2 

```{r}
pc1_wilcox_pl <- wilcox.test(pl_pca$x[stroke_samples,"PC1"], pl_pca$x[sham_samples,"PC1"])
pc2_wilcox_pl <- wilcox.test(pl_pca$x[stroke_samples,"PC2"], pl_pca$x[sham_samples,"PC2"])
```



```{r}
plasma_cols <- c("#9C9C9C","#D52D2D")
as.data.frame(pl_pca$x)%>%
  add_column(Group=pl_metabo_norm$sample.metadata$group,
             Infarct_volume=pl_metabo_norm$sample.metadata$indir_inf_vol) %>%
  ggplot(aes(x=PC1, y=PC2, colour=Group, fill=Group, shape=Group)) +
  geom_point(size=4) +
  theme_bw() +
  theme(panel.grid = element_blank()) + 
  scatter_theme +
  scale_color_manual(values = plasma_cols) +
  scale_fill_manual(values= plasma_cols) + 
  xlab(paste0("PC", 1, " (", round(pl_ev[1], 3)*100, "%)")) +
  ylab(paste0("PC", 2, " (", round(pl_ev[2], 3)*100, "%)")) + 
  stat_ellipse(geom = "polygon", alpha=0.1) + 
  geom_ysideboxplot(
    aes(fill = Group, y = PC2),
    color = "black",
    orientation = "x",
  ) +
  geom_xsideboxplot(
    aes(fill = Group, y = PC1),
    color = "black",
    orientation = "y"
  )
ggsave("figures/PCA_plasma.pdf", device = "pdf", dpi=300, height = 8, width = 10)
```

statistically detected outliers removed: excl CB61 & CB63

```{r}
pl_outliers <-unique(outliers_plasma$sample)
pl_filt <- colnames(pl_metabo_norm$features)[!colnames(pl_metabo_norm$features) %in% pl_outliers]
pl_filt_meta <- pl_metabo_norm$sample.metadata %>% 
  filter(!SampleID %in% pl_outliers) 


as.data.frame(pl_pca$x)%>%
  filter_rownames(pl_filt) %>% 
  add_column(Group=pl_filt_meta$group,
             Infarct_volume=pl_filt_meta$indir_inf_vol) %>%
  ggplot(aes(x=PC1, y=PC2, colour=Group, fill=Group, shape=Group)) +
  geom_point(size=4) +
  theme_pubr() +
  scatter_theme +
  scale_color_manual(values = plasma_cols) +
  scale_fill_manual(values= plasma_cols) + 
  xlab(paste0("PC", 1, " (", round(pl_ev[1], 3)*100, "%)")) +
  ylab(paste0("PC", 2, " (", round(pl_ev[2], 3)*100, "%)")) + 
  stat_ellipse(geom = "polygon", alpha=0.1)
ggsave("figures/PCA_plasma_excl_outliers.pdf", device = "pdf", dpi=300, height = 8, width = 10)
```

Visually detected outlier(s) excl CB52

```{r}
pl_filt <- colnames(pl_metabo_norm$features)[colnames(pl_metabo_norm$features) != "CB52"]
pl_filt_meta <- pl_metabo_norm$sample.metadata %>% 
  filter(SampleID != "CB52") 


as.data.frame(pl_pca$x)%>%
  filter_rownames(pl_filt) %>% 
  add_column(Group=pl_filt_meta$group,
             Infarct_volume=pl_filt_meta$indir_inf_vol) %>%
  ggplot(aes(x=PC1, y=PC2, colour=Group, fill=Group, shape=Group)) +
  geom_point(size=4) +
  theme_pubr() +
  scatter_theme +
  scale_color_manual(values = plasma_cols) +
  scale_fill_manual(values= plasma_cols) + 
  xlab(paste0("PC", 1, " (", round(pl_ev[1], 3)*100, "%)")) +
  ylab(paste0("PC", 2, " (", round(pl_ev[2], 3)*100, "%)")) + 
  stat_ellipse(geom = "polygon", alpha=0.1)
ggsave("figures/PCA_plasma_excl_CB52.pdf", device = "pdf", dpi=300, height = 8, width = 10)
```

### PCA colored by metabolite classes

```{r}
il_loadings <- merge(il_pca$rotation, il_metabo_norm$feature.metadata, by=0)
```

```{r}
as.data.frame(il_loadings)%>%
  ggplot(aes(x=PC1, y=PC2, colour=`SUPER PATHWAY`)) +
  geom_point(size=4) +
  theme_pubr() +
  ggsci::scale_color_npg() +
  scatter_theme +
  xlab(paste0("PC", 1, " (", round(il_ev[1], 3)*100, "%)")) +
  ylab(paste0("PC", 2, " (", round(il_ev[2], 3)*100, "%)")) +
  theme(legend.title = element_text(size=14),
    legend.text = element_text(size=12),
    legend.title.align=-30)
ggsave("figures/PCA_loadings_ileum.pdf", device = "pdf", dpi=300, height = 10, width = 12)
```

```{r}
as.data.frame(il_loadings)%>%
  ggplot(aes(x=PC1, y=PC2, colour=`SUB PATHWAY`)) +
  geom_point(size=4) +
  theme_pubr() +
  ggsci::scale_color_npg() +
  scatter_theme +
  xlab(paste0("PC", 1, " (", round(il_ev[1], 3)*100, "%)")) +
  ylab(paste0("PC", 2, " (", round(il_ev[2], 3)*100, "%)")) +
  gghighlight(`SUB PATHWAY` == "Tryptophan Metabolism", use_direct_label = F) + 
  theme(legend.position = "None")
ggsave("figures/PCA_loadings_ileum_trp_only.pdf", device = "pdf", dpi=300, height = 10, width = 12)
```

## Univariate analysis

Univariate analysis is performed by calculating fold change on log data
(geometric mean) and performing pairwise t-tests on scaled data (with
FDR adjustment for multiple comparisons).

### FC

```{r}
ileum_fc <- fold.change(il_metabo_norm, group_col = "group", group.order = c("Stroke", "Sham"))
plasma_fc <- fold.change(pl_metabo_norm, group_col = "group", group.order = c("Stroke", "Sham"))
```

```{r}
sig_ileum_str_0.25 <- ileum_fc %>% 
  filter(Log2FC > 0.58 & p.adj < 0.25)
sig_ileum_sham_0.25 <- ileum_fc %>% 
  filter(Log2FC < -0.58 & p.adj < 0.25)

kegg_id_str <- sig_ileum_str_0.25 %>% pull(KEGG)
kegg_id_sham <- sig_ileum_sham_0.25 %>% pull(KEGG)
```

Write out kegg ids for AMON

```{r}
fileConn <-file("results/KEGG_ids_stroke_0.25.txt")
writeLines(kegg_id_str, fileConn)
close(fileConn)

fileConn<-file("results/KEGG_ids_sham_0.25.txt")
writeLines(kegg_id_sham, fileConn)
close(fileConn)
```

Since we are interested in AHR ligands here we will look more
specifically at these.

### AHR ligands

```{r}
ahr_ligands <- c("xanthurenate", "2-oxindole-3-acetate", "kynurenate", "indoleacetylglycine",
                 "indolepropionylglycine", "3-indoxyl sulfate", "3-formylindole", "kynurenine", 
                 "5-hydroxyindole sulfate", "serotonin", "anthranilate", "quinolinate")
indole_derived <- c("2-oxindole-3-acetate",  "indoleacetylglycine",
                 "indolepropionylglycine", "3-indoxyl sulfate", "3-formylindole", "5-hydroxyindole sulfate",  
                 "indolelactate")
kynurenine_derived <- c("xanthurenate", "kynurenate","kynurenine", "anthranilate", "quinolinate")
serotonin_derived <- c("serotonin")
tryptophan_metabolites <- il_metabo_norm$feature.metadata %>% filter(`SUB PATHWAY`== "Tryptophan Metabolism") %>% 
  pull(`Metabolite`)
```

```{r}
host_derived <- c("xanthurenate", "kynurenate","kynurenine", "biliverdin", "bilirubin")
microbiome_derived <- c("3-indoxyl sulfate", "indolelactate", "3-formylindole")
```


#### Ileum AHR ligands

```{r, fig.height=12, fig.width=5}
ileum_fc_ahr <- ileum_fc %>% 
  select(Metabolite, Log2FC, p.adj, `SUPER PATHWAY`, `SUB PATHWAY`) %>% 
  mutate(AHR = case_when(Metabolite %in% ahr_ligands ~ "AHR", 
                         TRUE ~ "other")) %>% 
  mutate(Trp_pathway = case_when(Metabolite %in% indole_derived ~ "Indole Pyruvate Pathway",
                                 Metabolite %in% kynurenine_derived ~ "Kynurenine Pathway",
                                 Metabolite %in% serotonin_derived ~ "Serotonin Pathway",
                                 TRUE ~ "Other")) %>% 
  mutate(sig = case_when(p.adj <= 0.05 ~ "*",
                         p.adj <= 0.1 ~ "+",
                         p.adj <= 0.25 ~ "++",
                         TRUE ~ ""))

hm.palette <- cartography::carto.pal("orange.pal", n1 = 13)


ahr_ligands_il <-
  ggplot(filter(ileum_fc_ahr, Trp_pathway != "Other"), aes(x = 1,y = Metabolite, fill = Log2FC)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = alpha(hm.palette, 0.87)) +
  xlab("")  +
  facet_col(vars(Trp_pathway), scales = "free_y", space = "free",
            labeller = label_wrap_gen(width=10), strip.position = "right") + 
  #facet_wrap(~Trp_pathway,nrow = 3, ncol = 1, scales = "free", strip.position = "right") +  
  theme(plot.margin = unit(c(0,0,0,0), "cm"),
        strip.background = element_blank(),
        strip.text.y = element_text(size=16),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size=16),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        panel.background = element_blank(),
        legend.justification = "top",
        legend.title = element_blank(),
        legend.text = element_text(size=11),
        plot.title = element_text(size=20, hjust = 0.5)) +
  geom_text(aes(label = sig), size=6) + 
  guides(fill = guide_colourbar(barwidth = 3, ticks.colour="black"))  + 
 ggtitle("AHR ligands FC Stroke/Sham")
ahr_ligands_il
ggsave("figures/ahr_ligands_ileum.pdf", ahr_ligands_il, device = "pdf", dpi=300, height = 11, width = 5.5)
```
Host derived vs microbiome-derived

```{r}
ileum_fc_ahr_source <- ileum_fc %>% 
  select(Metabolite, Log2FC, p.adj, `SUPER PATHWAY`, `SUB PATHWAY`) %>% 
  mutate(AHR = case_when(Metabolite %in% ahr_ligands ~ "AHR", 
                         TRUE ~ "other")) %>% 
  mutate(Source = case_when(Metabolite %in% host_derived ~ "Host-derived AHR ligands",
                                 Metabolite %in% microbiome_derived ~ "Microbiome-derived AHR ligands",
                                 TRUE ~ "Other")) %>% 
  mutate(sig = case_when(p.adj <= 0.05 ~ "*",
                         p.adj <= 0.1 ~ "+",
                         p.adj <= 0.25 ~ "++",
                         TRUE ~ "")) 

ahr_ligands_source_il <-
  ggplot(filter(ileum_fc_ahr_source, Source != "Other"), aes(x = 1,y = Metabolite, fill = Log2FC)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = alpha(hm.palette, 0.87)) +
  xlab("")  +
  facet_col(.~factor(Source, rev(stringr::str_sort(unique(Source)))), scales = "free_y", space = "free",
            labeller = label_wrap_gen(width=10), strip.position = "right") + 
  #facet_wrap(~Trp_pathway,nrow = 3, ncol = 1, scales = "free", strip.position = "right") +  
  theme(plot.margin = unit(c(0,0,0,0), "cm"),
        strip.background = element_blank(),
        strip.text.y = element_text(size=16),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size=16),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        panel.background = element_blank(),
        legend.justification = "top",
        legend.title = element_blank(),
        legend.text = element_text(size=11),
        plot.title = element_text(size=20, hjust = 0.5)) +
  geom_text(aes(label = sig), size=6) + 
  guides(fill = guide_colourbar(barwidth = 3, ticks.colour="black"))  + 
 ggtitle("Ileum AHR ligands (Stroke/Sham)")
ahr_ligands_source_il
ggsave("figures/ahr_ligands_source_ileum.pdf", ahr_ligands_source_il, device = "pdf", dpi=300, height = 11, width = 5.5)
```


#### Plasma AHR ligands 

```{r}
plasma_fc_ahr <- plasma_fc %>% 
  select(BIOCHEMICAL, Log2FC, p.adj, `SUPER PATHWAY`, `SUB PATHWAY`) %>% 
  mutate(AHR = case_when(BIOCHEMICAL %in% ahr_ligands ~ "AHR", 
                         TRUE ~ "other")) %>% 
  mutate(Trp_pathway = case_when(BIOCHEMICAL %in% indole_derived ~ "Indole Pyruvate Pathway",
                                 BIOCHEMICAL %in% kynurenine_derived ~ "Kynurenine Pathway",
                                 BIOCHEMICAL %in% serotonin_derived ~ "Serotonin Pathway",
                                 TRUE ~ "Other")) %>% 
  mutate(sig = case_when(p.adj <= 0.05 ~ "*",
                         p.adj <= 0.1 ~ "+",
                         p.adj <= 0.25 ~ "++",
                         TRUE ~ ""))


ahr_ligands_pl <-
  ggplot(filter(plasma_fc_ahr, Trp_pathway != "Other"), aes(x = 1,y = BIOCHEMICAL, fill = Log2FC)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = alpha(hm.palette, 0.87)) +
  xlab("")  +
  facet_col(vars(Trp_pathway), scales = "free_y", space = "free",
            labeller = label_wrap_gen(width=10), strip.position = "right") + 
  #facet_wrap(~Trp_pathway,nrow = 3, ncol = 1, scales = "free", strip.position = "right") +  
  theme(plot.margin = unit(c(0,0,0,0), "cm"),
        strip.background = element_blank(),
        strip.text.y = element_text(size=16),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size=16),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        panel.background = element_blank(),
        legend.justification = "top",
        legend.title = element_blank(),
        legend.text = element_text(size=11),
        plot.title = element_text(size=20, hjust = 0.5)) +
  geom_text(aes(label = sig), size=6) + 
  guides(fill = guide_colourbar(barwidth = 3, ticks.colour="black"))  + 
 ggtitle("AHR ligands FC Stroke/Sham")
ahr_ligands_pl
ggsave("figures/ahr_ligands_plasma.pdf", ahr_ligands_pl, device = "pdf", dpi=300, height = 11, width = 5.5)
```
Host derived vs microbiome-derived

```{r}
plasma_fc_ahr_source <- plasma_fc %>% 
  select(BIOCHEMICAL, Log2FC, p.adj, `SUPER PATHWAY`, `SUB PATHWAY`) %>% 
  mutate(AHR = case_when(BIOCHEMICAL %in% ahr_ligands ~ "AHR", 
                         TRUE ~ "other")) %>% 
  mutate(Source = case_when(BIOCHEMICAL %in% host_derived ~ "Host-derived AHR ligands",
                                 BIOCHEMICAL %in% microbiome_derived ~ "Microbiome-derived AHR ligands",
                                 TRUE ~ "Other")) %>% 
  mutate(sig = case_when(p.adj <= 0.05 ~ "*",
                         p.adj <= 0.1 ~ "+",
                         p.adj <= 0.25 ~ "++",
                         TRUE ~ ""))

ahr_ligands_source_pl <-
  ggplot(filter(plasma_fc_ahr_source, Source != "Other"), aes(x = 1,y = BIOCHEMICAL, fill = Log2FC)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = alpha(hm.palette, 0.87)) +
  xlab("")  +
  facet_col(.~factor(Source, rev(stringr::str_sort(unique(Source)))), scales = "free_y", space = "free",
            labeller = label_wrap_gen(width=10), strip.position = "right") + 
  #facet_wrap(~Trp_pathway,nrow = 3, ncol = 1, scales = "free", strip.position = "right") +  
  theme(plot.margin = unit(c(0,0,0,0), "cm"),
        strip.background = element_blank(),
        strip.text.y = element_text(size=16),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size=16),
        axis.ticks = element_blank(),
        axis.title.y = element_blank(),
        panel.background = element_blank(),
        legend.justification = "top",
        legend.title = element_blank(),
        legend.text = element_text(size=11),
        plot.title = element_text(size=20, hjust = 0.5)) +
  geom_text(aes(label = sig), size=6) + 
  guides(fill = guide_colourbar(barwidth = 3, ticks.colour="black"))  + 
 ggtitle("Plasma AHR ligands (Stroke/Sham)")
ahr_ligands_source_pl
ggsave("figures/ahr_ligands_source_plasma.pdf", ahr_ligands_source_pl, device = "pdf", dpi=300, height = 11, width = 5.5)
```


### Volcano plots

In both cases (ileum and plasma) a log2 FC cutoff of 0.58 (equal to 1.5x
actual FC) and adjusted p-value threshold of 0.05 is used.

#### Ileum volcano

Coloured by group

```{r}
plot_volcano(ileum_fc, groups = c("Sham", "Stroke"), names = ileum_fc$Metabolite, pointSize = 3) + 
  xlim(c(-5, 3)) +
  ylim(c(0, 3))
ggsave("figures/ileum_volcano.pdf", device = "pdf", dpi=300, height = 9, width = 8)
```


Coloured by metabolite class

```{r}
res_ileum <- ileum_fc %>% 
  mutate(`SUPER PATHWAY`= recode(`SUPER PATHWAY`, Lipid = "Lipids",
                                 `Amino Acid` = "Amino Acids",
                                 Carbohydrate = "Carbohydrates",
                                 Nucleotide = "Nucleotides",
                                 Peptide = "Peptides"))


keyvals <- case_when(res_ileum$`SUPER PATHWAY` == "Amino Acids" ~ "#3074B4",
                     res_ileum$`SUPER PATHWAY` == "Carbohydrates" ~ "#FF9200",
                     res_ileum$`SUPER PATHWAY` == "Cofactors and Vitamins" ~ "#FFD379",
                     res_ileum$`SUPER PATHWAY` == "Energy" ~ "#75D179",
                     res_ileum$`SUPER PATHWAY` == "Lipids"  ~ "#009091",
                     res_ileum$`SUPER PATHWAY` == "Nucleotides" ~ "#FF2600",
                     res_ileum$`SUPER PATHWAY` == "Partially Characterized Molecules" ~ "#BFBFBF",
                     res_ileum$`SUPER PATHWAY` == "Peptides"  ~ "#76D5FF",
                     res_ileum$`SUPER PATHWAY` == "Xenobiotics"  ~ "#FF7E79",
                     TRUE ~ "")

names(keyvals)[keyvals == '#3074B4'] <- 'Amino Acids'
names(keyvals)[keyvals == '#FF9200'] <- 'Carbohydrates'
names(keyvals)[keyvals == '#FFD379'] <- 'Cofactors and Vitamins'
names(keyvals)[keyvals == '#75D179'] <- 'Energy'
names(keyvals)[keyvals == '#009091'] <- 'Lipids'
names(keyvals)[keyvals == '#FF2600'] <- 'Nucleotides'
names(keyvals)[keyvals == '#BFBFBF'] <- 'Partially Characterized Molecules'
names(keyvals)[keyvals == '#76D5FF'] <- 'Peptides'
names(keyvals)[keyvals == '#FF7E79'] <- 'Xenobiotics'


EnhancedVolcano(res_ileum, lab = NA,x="Log2FC", y="p.adj",pointSize = 4,
                colCustom = keyvals, FCcutoff = 0.58, pCutoff = 0.05, xlim = c(-5, 3), 
                ylim = c(0,3), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL, legendLabSize = 9.5)
ggsave("figures/ileum_volcano_classes.pdf", device = "pdf", dpi=300, height = 9.5, width = 9)
```



#### Plasma volcano

Coloured by group

```{r}
plot_volcano(plasma_fc, groups = c("Sham", "Stroke"), names = plasma_fc$BIOCHEMICAL, 
             pointSize = 3) + 
  xlim(c(-4, 4)) +
  ylim(c(0, 4))
ggsave("figures/plasma_volcano_groups.pdf", device = "pdf", dpi=300, height = 9, width = 8)
```

Coloured by metabolite class

```{r}
res_plasma <- plasma_fc %>% 
  mutate(`SUPER PATHWAY`= recode(`SUPER PATHWAY`, Lipid = "Lipids",
                                 `Amino Acid` = "Amino Acids",
                                 Carbohydrate = "Carbohydrates",
                                 Nucleotide = "Nucleotides",
                                 Peptide = "Peptides"))


keyvals <- case_when(res_plasma$`SUPER PATHWAY` == "Amino Acids" ~ "#3074B4",
                     res_plasma$`SUPER PATHWAY` == "Carbohydrates" ~ "#FF9200",
                     res_plasma$`SUPER PATHWAY` == "Cofactors and Vitamins" ~ "#FFD379",
                     res_plasma$`SUPER PATHWAY` == "Energy" ~ "#75D179",
                     res_plasma$`SUPER PATHWAY` == "Lipids"  ~ "#009091",
                     res_plasma$`SUPER PATHWAY` == "Nucleotides" ~ "#FF2600",
                     res_plasma$`SUPER PATHWAY` == "Partially Characterized Molecules" ~ "#BFBFBF",
                     res_plasma$`SUPER PATHWAY` == "Peptides"  ~ "#76D5FF",
                     res_plasma$`SUPER PATHWAY` == "Xenobiotics"  ~ "#FF7E79",
                     TRUE ~ "")

names(keyvals)[keyvals == '#3074B4'] <- 'Amino Acids'
names(keyvals)[keyvals == '#FF9200'] <- 'Carbohydrates'
names(keyvals)[keyvals == '#FFD379'] <- 'Cofactors and Vitamins'
names(keyvals)[keyvals == '#75D179'] <- 'Energy'
names(keyvals)[keyvals == '#009091'] <- 'Lipids'
names(keyvals)[keyvals == '#FF2600'] <- 'Nucleotides'
names(keyvals)[keyvals == '#BFBFBF'] <- 'Partially Characterized Molecules'
names(keyvals)[keyvals == '#76D5FF'] <- 'Peptides'
names(keyvals)[keyvals == '#FF7E79'] <- 'Xenobiotics'


EnhancedVolcano(res_plasma, lab = NA,x="Log2FC", y="p.adj",pointSize = 4,
                colCustom = keyvals, FCcutoff = 0.58, pCutoff = 0.05, xlim = c(-4, 4), 
                ylim = c(0,4), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL, legendLabSize = 9.5)
ggsave("figures/plasma_volcano_classes.pdf", device = "pdf", dpi=300, height = 9.5, width = 9)
```

### Individual plots - AHR ligands

#### Ileum ahr ligands

```{r}
feat_long_il <- il_metabo_norm$log.features %>% 
  as.data.frame()
rownames(feat_long_il) <- il_metabo_norm$feature.metadata$Metabolite

feat_long_il <- feat_long_il %>% 
  rownames_to_column("Metabolite") %>% 
  pivot_longer(-Metabolite, names_to = "SampleID")

```

```{r}
sham_samples <- il_metabo_norm$sample.metadata %>% filter(group == "Sham") %>% pull(SampleID)
stroke_samples <-il_metabo_norm$sample.metadata %>% filter(group == "Stroke") %>% pull(SampleID)

feat_long_meta_il <- feat_long_il %>% 
  mutate(group = case_when(SampleID %in% sham_samples ~ "Sham",
                           SampleID %in% stroke_samples ~ "Stroke",
                           TRUE ~ "other"))
```

```{r}
dir.create("figures/AHR_ileum")
ahr_ligands2 <- c("tryptophan", "indolelactate", ahr_ligands)
for (i in ahr_ligands2){
  sub <- filter(feat_long_meta_il, Metabolite == i)
  print(sub)

  p <- ggstripchart(sub, x="group", y="value", color = "group", palette = ileum_cols,
                    xlab = "", ylab = "LogInt", size = 4, jitter = 0.05, 
                    add = "mean_sd", add.params = list(color="black"), order = c("Sham", "Stroke"),
                    error.plot = c("errorbar"), title = i)  + 
    stat_summary(geom = "crossbar", fun = mean, colour="black",linewidth=0.4, width=0.15) +
  stat_compare_means(comparisons = list(c("Sham", "Stroke")), method = "t.test", label = "p.format") + 
    theme_pubr() +
    theme(legend.position = "None",
          axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          plot.title = element_text(size=16, hjust = 0.75))
  ggsave(filename = paste0("figures/AHR_ileum/", i, "_ileum.pdf"), p, device="pdf", dpi=300, height=4.25, width=2.5)
  print(p)
}
```

```{r}
dir.create("figures/Trp_ileum")
for (i in tryptophan_metabolites){
  sub <- filter(feat_long_meta_il, Metabolite == i)

  p <- ggstripchart(sub, x="group", y="value", color = "group", palette = ileum_cols,
                    xlab = "", ylab = "LogInt", size = 4, jitter = 0.05, 
                    add = "mean_sd", add.params = list(color="black"), order = c("Sham", "Stroke"),
                    error.plot = c("errorbar"), title = i) +
    stat_summary(geom = "crossbar", fun = mean, colour="black",linewidth=0.4, width=0.15) +
  stat_compare_means(comparisons = list(c("Sham", "Stroke")), method = "t.test", label = "p.signif") + 
    theme_pubr() +   
    theme(legend.position = "None",
          axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          plot.title = element_text(size=16, hjust = 0.75)) 
  print(p)
  ggsave(filename = paste0("figures/Trp_ileum/", i, "_ileum.pdf"), p, device="pdf", dpi=300, height=4.25, width=2.5)
}
```

##### Other metabolites 

```{r}
other_metabolites <- c("stachyose", "orotate", "glucose", "pyruvate","lactate", "glycerol", "malate", "oxaloacetate")
```

```{r}
dir.create("figures/other_metabolites_ileum")

for (i in other_metabolites){
  sub <- filter(feat_long_meta_il, Metabolite == i)
  print(sub)

  p <- ggstripchart(sub, x="group", y="value", color = "group", palette = ileum_cols,
                    xlab = "", ylab = "LogInt", size = 4, jitter = 0.05, 
                    add = "mean_sd", add.params = list(color="black"), order = c("Sham", "Stroke"),
                    error.plot = c("errorbar"), title = i)  + 
    stat_summary(geom = "crossbar", fun = mean, colour="black",linewidth=0.4, width=0.15) +
  stat_compare_means(comparisons = list(c("Sham", "Stroke")), method = "t.test", label = "p.format") + 
    theme_pubr() +
    theme(legend.position = "None",
          axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          plot.title = element_text(size=16, hjust = 0.75))
  ggsave(filename = paste0("figures/other_metabolites_ileum/", i, "_ileum.pdf"), p, device="pdf", dpi=300, height=4.25, width=2.5)
  print(p)
}
```

#### Plasma ahr ligands

```{r}
feat_long_pl <- pl_metabo_norm$log.features %>% 
  as.data.frame()
rownames(feat_long_pl) <- pl_metabo_norm$feature.metadata$BIOCHEMICAL

feat_long_pl <- feat_long_pl %>% 
  rownames_to_column("Metabolite") %>% 
  pivot_longer(-Metabolite, names_to = "SampleID")

```

```{r}
feat_long_meta_pl <- feat_long_pl %>% 
  mutate(group = case_when(SampleID %in% sham_samples ~ "Sham",
                           SampleID %in% stroke_samples ~ "Stroke",
                           TRUE ~ "other"))
```

```{r}
dir.create("figures/AHR_plasma")
for (i in ahr_ligands){
  sub <- filter(feat_long_meta_pl, Metabolite == i)

  p <- ggstripchart(sub, x="group", y="value", color = "group", palette = plasma_cols,
                    xlab = "", ylab = "LogInt", size = 4, jitter = 0.05, 
                    add = "mean_sd", add.params = list(color="black"), , order = c("Sham", "Stroke"),
                    error.plot = c("errorbar"), title = i) +
    stat_summary(geom = "crossbar", fun = mean, colour="black",linewidth=0.4, width=0.15)  + 
  stat_compare_means(comparisons = list(c("Sham", "Stroke")), method = "t.test", label = "p.signif") + 
    theme_pubr() + 
    theme(legend.position = "None",
          axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          plot.title = element_text(size=16, hjust = 0.75))
  print(p)
  ggsave(filename = paste0("figures/AHR_plasma/", i, "_plasma.pdf"), p, device="pdf", dpi=300, height=4.25, width=2.5)
}
```

```{r}
dir.create("figures/Trp_plasma")
for (i in tryptophan_metabolites){
  sub <- filter(feat_long_meta_pl, Metabolite == i)

  p <- ggstripchart(sub, x="group", y="value", color = "group", palette = plasma_cols,
                    xlab = "", ylab = "LogInt", size = 4, jitter = 0.05, 
                    add = "mean_sd", add.params = list(color="black"), , order = c("Sham", "Stroke"),
                    error.plot = c("errorbar"), title = i) +
    stat_summary(geom = "crossbar", fun = mean, colour="black",linewidth=0.4, width=0.15)  + 
  stat_compare_means(comparisons = list(c("Sham", "Stroke")), method = "t.test", label = "p.signif") + 
    theme_pubr() + 
    theme(legend.position = "None",
          axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          plot.title = element_text(size=16, hjust = 0.75))
  print(p)
  ggsave(filename = paste0("figures/Trp_plasma/", i, "_plasma.pdf"), p, device="pdf", dpi=300, height=4.25, width=2.5)
}
```
##### Other metabolites 

```{r}
dir.create("figures/other_metabolites_plasma")

for (i in other_metabolites){
  sub <- filter(feat_long_meta_pl, Metabolite == i)
  print(sub)

  p <- ggstripchart(sub, x="group", y="value", color = "group", palette = plasma_cols,
                    xlab = "", ylab = "LogInt", size = 4, jitter = 0.05, 
                    add = "mean_sd", add.params = list(color="black"), order = c("Sham", "Stroke"),
                    error.plot = c("errorbar"), title = i)  + 
    stat_summary(geom = "crossbar", fun = mean, colour="black",linewidth=0.4, width=0.15) +
  stat_compare_means(comparisons = list(c("Sham", "Stroke")), method = "t.test", label = "p.format") + 
    theme_pubr() +
    theme(legend.position = "None",
          axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          plot.title = element_text(size=16, hjust = 0.75))
  ggsave(filename = paste0("figures/other_metabolites_plasma/", i, "_plasma.pdf"), p, device="pdf", dpi=300, height=4.25, width=2.5)
  print(p)
}
```

### Trp/Kyn ratio

```{r}
keep_cols <- c("tryptophan", "group", "perc_weight_loss", "indir_inf_vol")
ahr_corr_ileum <- il_metabo_norm$log.features %>% 
  merge(il_metabo_norm$feature.metadata, by=0) %>% 
  column_to_rownames("BIOCHEMICAL") %>%
  select(contains("CB")) %>% 
  t() %>% 
  merge(il_metabo_norm$sample.metadata %>% column_to_rownames("SampleID"), by=0) %>% 
  column_to_rownames("Row.names") %>% 
  select(all_of(c(ahr_ligands, keep_cols)), indir_inf_vol, dir_inf_vol)
```

```{r}
ahr_corr_ileum <- ahr_corr_ileum %>% 
  mutate(Trp_Kyn_ratio = kynurenine / tryptophan)
```

```{r}
ratio_dat <- ahr_corr_ileum %>% 
  select(c(group, Trp_Kyn_ratio, indir_inf_vol, perc_weight_loss))

ggbarplot(
  ahr_corr_ileum,
  x = "group",
  y = "Trp_Kyn_ratio",
  fill = "group",
  alpha=0.9,
  add = c("mean_se","jitter"),
  order = c("Sham", "Stroke"),
  palette = ileum_cols,
  width = 0.5
) + 
scale_y_continuous(expand=c(0,0), limits = c(0, 0.8)) + 
  ylab("Kynurenine/Tryptophan Ratio") + 
  stat_compare_means(method = "t.test", label.y = 0.75, label.x = 1.45, label = "p.signif")
```

Does the ratio correlate with infarct volume in stroke mice?

```{r}
ratio_dat %>% 
  filter(group == "Stroke") %>% 
  ggscatter(
    x = "indir_inf_vol",
    y = "Trp_Kyn_ratio",
    add = "reg.line",
    conf.int = T,
    color = ileum_cols[[2]], 
    cor.coef = T,
    cor.method = "pearson"
  ) + 
  ylab("Kynurenine/Tryptophan Ratio")
```

Does the ratio correlate with weight loss in stroke mice?

```{r}
ratio_dat %>% 
  filter(group == "Stroke") %>% 
  ggscatter(
    x = "perc_weight_loss",
    y = "Trp_Kyn_ratio",
    add = "reg.line",
    conf.int = T,
    color = ileum_cols[[2]], 
    cor.coef = T,
    cor.method = "pearson"
  ) + 
  ylab("Kynurenine/Tryptophan Ratio")
```

### Tryptophan metabolite correlation with stroke severity/weight loss

```{r}
infarct_vol <- il_metabo_norm$sample.metadata$indir_inf_vol
weight_loss <- il_metabo_norm$sample.metadata$perc_weight_loss
il.plsda.infarct <- opls(t(il_metabo_norm$log.features), infarct_vol, predI = 2)
il.plsda.sev <- opls(t(il_metabo_norm$log.features), weight_loss, predI = 2)
```

```{r}
VIP_il.infarct <- data.frame("Metabolite" = names(il.plsda.infarct@vipVn), "VIP"= il.plsda.infarct@vipVn) %>% 
  slice_max(VIP, n = 100)

metabolite_names.infarc <- il_metabo_norm$feature.metadata %>% 
  filter_rownames(VIP_il.infarct$Metabolite) %>% 
  pull(BIOCHEMICAL)
```

```{r}
VIP_il.sev <- data.frame("Metabolite" = names(il.plsda.sev@vipVn), "VIP"= il.plsda.sev@vipVn) %>% 
  slice_max(VIP, n = 100)

metabolite_names.sev <- il_metabo_norm$feature.metadata %>% 
  filter_rownames(VIP_il.sev$Metabolite) %>% 
  pull(BIOCHEMICAL)
```

Are any of the ahr ligands correlated with infarct volume?

```{r}
ahr_corr_ileum_stroke <- ahr_corr_ileum %>% 
  filter(group == "Stroke")

colnames(ahr_corr_ileum_stroke)[1:14] <- paste("Trp", colnames(ahr_corr_ileum_stroke)[1:14], sep="_")
colnames(ahr_corr_ileum_stroke)[1:14] <- gsub("-","_",colnames(ahr_corr_ileum_stroke)[1:14])

ahr_ligands2 <- paste("Trp", c(ahr_ligands, "tryptophan"), sep="_")
ahr_ligands2 <- gsub("-","_",ahr_ligands2)
```

```{r}
for (i in ahr_ligands2){
  sub <- select(ahr_corr_ileum_stroke, .data[[ i ]], indir_inf_vol)
  featname <- gsub("Trp_", "", i)
  p <- ggscatter(sub, x="indir_inf_vol", y=i, color = ileum_cols[[2]], 
                    xlab = "Infarct volume mm3", ylab = paste("LogInt",featname, sep=" "), size = 4,
                    add = "reg.line",conf.int = T) +
  stat_cor(color = ileum_cols[[2]], method = "spearman")
    theme_pubr() +
    theme(legend.position = "None",
          axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          plot.title = element_text(size=16, hjust = 0.75))
    print(p)
  # ggsave(filename = paste0("figures/Trp_ileum/", i, "_ileum.pdf"), p, device="pdf", dpi=300, height=4.25, width=2.5)
}
```

With weight loss?

```{r}
for (i in ahr_ligands2){
  sub <- select(ahr_corr_ileum_stroke, .data[[ i ]], perc_weight_loss)
  featname <- gsub("Trp_", "", i)
  
  p <- ggscatter(sub, x="perc_weight_loss", y=i, color = ileum_cols[[2]], 
                    xlab = "Weight loss d3 post-stroke [%]", ylab = paste("LogInt",featname, sep=" "), size = 4,
                    add = "reg.line",conf.int = T) +
  stat_cor(color = ileum_cols[[2]], method = "spearman")
    theme_pubr() +
    theme(legend.position = "None",
          axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          plot.title = element_text(size=16, hjust = 0.75))
    print(p)
  # ggsave(filename = paste0("figures/Trp_ileum/", i, "_ileum.pdf"), p, device="pdf", dpi=300, height=4.25, width=2.5)
}
```

## Pathway enrichment

Subpathway enrichment based on approach used by Granados et al 2021, JBC

"Enrichment was determined using Equation 1, where ***k*** is the number of
significantly altered metabolites in a Subpathway, ***m*** is the number
of metabolites in a Subpathway, ***n*** is the number of significantly
altered metabolites in the total data set, and ***N*** is the number of
measured metabolites in the total data set."

$$ 
enrichment = \frac{\frac{k}{m}}{\frac{n-k}{N-m}}
$$

```{r}
subp_enrichment <- subpathway_enrichment(ileum_fc, thresh = 0.25)
```

```{r}
subp_enrichment %>%
  slice_max(Enrichment, n=15) %>%
  ggplot(aes(x = fct_reorder(SUB_PATHWAY, Enrichment), y = Enrichment))+
  geom_col(fill = "#0FA3B1", width = 0.7) + 
  scale_y_continuous(expand=c(0,0)) + 
  theme_cowplot() + 
  coord_flip() + 
  xlab("Subpathwy") + 
  ylab("Enrichment Score")
ggsave("figures/subpathway_enrichmment.pdf",device = "pdf", dpi=300, height = 7, width = 9)
```


Pathway enrichment is conducted by considering enriched metabolites
(qval \<0.1) and using Metabolon pathway annotations.

```{r}
stroke_enriched_il <- ileum_fc %>% 
  filter(p.adj < 0.1 & Log2FC > 0) %>% 
  pull(BIOCHEMICAL)


```

## Bile acids

```{r}
primary <- res_ileum %>% filter(`SUB PATHWAY` == "Primary Bile Acid Metabolism") %>% 
  pull(Metabolite)
secondary <- res_ileum %>% filter(`SUB PATHWAY` == "Secondary Bile Acid Metabolism")  %>% 
  pull(Metabolite)
```

"Animated plot"

```{r}
base_plot <- EnhancedVolcano(res_ileum, lab = NA,x="Log2FC", y="p.adj",pointSize = 4,
                col = "lightgrey", FCcutoff = 0.58, pCutoff = 0.05, xlim = c(-5, 3), 
                ylim = c(0,3), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL, legendLabSize = 9.5)
base_plot
ggsave("../../../Metagenomics/mouse_stool/analysis/Rosa_thesis//base_plot_ileum.pdf", device = "pdf", dpi=300, height = 8, width = 10)
keyvals_lipids <- keyvals 
keyvals_lipids[keyvals_lipids != "#009091"] <- "lightgrey"
lipids <- EnhancedVolcano(res_ileum, lab = NA,x="Log2FC", y="p.adj",pointSize = 4,
                colCustom  = keyvals_lipids, FCcutoff = 0.58, pCutoff = 0.05, xlim = c(-5, 3), 
                ylim = c(0,3), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL, legendLabSize = 9.5)
lipids
ggsave("../../../Metagenomics/mouse_stool/analysis/Rosa_thesis//lipids_plot_ileum.pdf", device = "pdf", dpi=300, height = 8, width = 10)

lipids_primary <- EnhancedVolcano(res_ileum, lab = res_ileum$Metabolite,x="Log2FC", y="p.adj",pointSize = 4,
                colCustom  = keyvals_lipids, FCcutoff = 0.58, pCutoff = 0.05, xlim = c(-5, 3), 
                ylim = c(0,3), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL, legendLabSize = 9.5,
                selectLab = primary,drawConnectors = TRUE, labSize = 4)
lipids_primary
ggsave("../../../Metagenomics/mouse_stool/analysis/Rosa_thesis//lipids_plot_ileum_primary.pdf", device = "pdf", dpi=300, height = 8, width = 10)
lipids_secondary <- EnhancedVolcano(res_ileum, lab = res_ileum$Metabolite,x="Log2FC", y="p.adj",pointSize = 4,
                colCustom  = keyvals_lipids, FCcutoff = 0.58, pCutoff = 0.05, xlim = c(-5, 3), 
                ylim = c(0,3), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL, legendLabSize = 9.5,
                selectLab = secondary,drawConnectors = TRUE, labSize = 4)
lipids_secondary
ggsave("../../../Metagenomics/mouse_stool/analysis/Rosa_thesis//lipids_plot_ileum_secondary.pdf", device = "pdf", dpi=300, height = 8, width = 10)
```

```{r}

```

```{r}
base_plot <- EnhancedVolcano(res_plasma, lab = NA,x="Log2FC", y="p.adj",pointSize = 4,
                col = "lightgrey", FCcutoff = 0.58, pCutoff = 0.05, xlim = c(-5, 3), 
                ylim = c(0,4), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL, legendLabSize = 9.5)
base_plot
ggsave("figures/base_plot_plasma.pdf", device = "pdf", dpi=300, height = 8, width = 10)
keyvals_lipids <- keyvals 
keyvals_lipids[keyvals_lipids != "#009091"] <- "lightgrey"
lipids <- EnhancedVolcano(res_plasma, lab = NA,x="Log2FC", y="p.adj",pointSize = 4,
                colCustom  = keyvals_lipids, FCcutoff = 0.58, pCutoff = 0.05, xlim = c(-5, 3), 
                ylim = c(0,4), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL, legendLabSize = 9.5)
lipids
ggsave("figures/lipids_plot_plasma.pdf", device = "pdf", dpi=300, height = 8, width = 10)

lipids_primary <- EnhancedVolcano(res_plasma, lab = res_plasma$BIOCHEMICAL,x="Log2FC", y="p.adj",pointSize = 4,
                colCustom  = keyvals_lipids, FCcutoff = 0.58, pCutoff = 0.05, xlim = c(-5, 3), 
                ylim = c(0,4), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL, legendLabSize = 9.5,
                selectLab = primary,drawConnectors = TRUE,
                labSize = 4)
lipids_primary
ggsave("figures/lipids_plot_plasma_primary.pdf", device = "pdf", dpi=300, height = 8, width = 10)
lipids_secondary <- EnhancedVolcano(res_plasma, lab = res_plasma$BIOCHEMICAL,x="Log2FC", y="p.adj",pointSize = 4,
                colCustom  = keyvals_lipids, FCcutoff = 0.58, pCutoff = 0.05, xlim = c(-5, 3), 
                ylim = c(0,4), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL, legendLabSize = 9.5,
                selectLab = secondary,drawConnectors = TRUE, labSize = 4)
lipids_secondary
ggsave("figures/lipids_plot_plasma_secondary.pdf", device = "pdf", dpi=300, height = 8, width = 10)
```

## Other metabolites 

Colored by p-cresol metabolites 

```{r}
p_cresol_metabolites <- c("p-cresol sulfate", "2-amino-p-cresol sulfate","tyrosine")
res_ileum <- ileum_fc %>% 
  mutate(`p-cresol_metabolism`= case_when(BIOCHEMICAL %in% p_cresol_metabolites ~ "p-cresol metabolism",
                                          TRUE ~ "other"))


keyvals <- case_when(res_ileum$`p-cresol_metabolism` == "p-cresol metabolism" ~ "#009091",
                     TRUE ~ "lightgrey")

names(keyvals)[keyvals == '#009091'] <- 'p-cresol metabolisms'
names(keyvals)[keyvals == 'lightgrey'] <- 'other'



EnhancedVolcano(res_ileum, lab = NA,x="Log2FC", y="p.adj",pointSize = 4,
                colCustom = keyvals, FCcutoff = 0.58, pCutoff = 0.1, xlim = c(-5, 3), 
                ylim = c(0,3), caption = NULL, subtitle = NULL, gridlines.major = F, 
                gridlines.minor = F, title = NULL,colAlpha = 0.15, legendLabSize = 9.5)
ggsave("figures/ileum_volcano_samuele_colored.pdf", device = "pdf", dpi=300, height = 9.5, width = 9)
```

### Xenobiotics heatmap 


```{r}
xenobiotics <- il_metabo_norm$feature.metadata %>% 
  filter(`SUPER PATHWAY` == "Xenobiotics") 
  
xenobiotic_ids <- xenobiotics %>% 
  rownames()
```


```{r}
il_xenobiotics <- il_metabo_norm$scaled.features %>% 
  as.data.frame() %>% 
  rownames_to_column("Metabolite") %>% 
  filter(Metabolite %in% xenobiotic_ids) %>% 
  column_to_rownames("Metabolite")
```

```{r}
rownames(il_xenobiotics) == rownames(xenobiotics)
```

```{r}
rownames(il_xenobiotics) <- xenobiotics$Metabolite
```

```{r}
pdf(file="Xenobiotic_heatmap.pdf", width=14, height=11)
ht <- ComplexHeatmap::Heatmap(il_xenobiotics, show_row_names = T, show_column_names = F, 
                    cluster_rows = T, cluster_columns = T, name = "Scaled Intensity",
                    column_split = il_metabo_norm$sample.metadata$group)
ComplexHeatmap::draw(ht)
dev.off()
```
