---
title: "16S analysis template"
author: "Adam Sorbie"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    df_print: paged
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Microbiota analysis

This document contains a template for conducting routine analysis of microbiota data (16S rRNA sequencing).

## Set project directories

```{r set-directories}
scripts_dir <- "../scripts/"
data_dir <- "../data/"
```

## Load libraries

```{r load-libraries, warning=FALSE, message=FALSE}
# download latest version of analysis script and source 
if (!file.exists(paste0(scripts_dir,"/analysis_functions.R"))){
  download.file("https://raw.githubusercontent.com/adamsorbie/Stroke_Microbiota_reproducibility/main/scripts/analysis_functions.R", destfile = paste0(scripts_dir,"/analysis_functions.R"))
}

source( paste0(scripts_dir,"/analysis_functions.R"))
# put other functions in benakis lab repo
devtools::source_url("https://raw.githubusercontent.com/adamsorbie/BenakisLab/main/16S/analysis/helper_functions_microbiome.R")
## additional libraries
pacman::p_load(microViz, microbiome, microbiomeutilities, cowplot, shiny, conflicted, plotly)
conflict_prefer_all("rstatix")
conflict_prefer_all("purrr")
```

## Set project attributes

Fill the details accordingly.

**Project title** : Analysis of 16S sequencing from AhR KO mice and controls\
**Main author** : Adam Sorbie\
**Principal Investigator (s)** : Corinne Benakis

```{r set-seed}
set.seed(42)
```

```{r set-attributes}
# palette for categories 
cat.palette <- c("#FF3030", "#1E90FF", "#2C6E49", "#F0C808", "#EEF0F2")
stacked_bar.palette <- c("#E64B35FF", "#4DBBD5FF", "#00A087FF", "#3C5488FF", "#F39B7FFF", 
                         "#8491B4FF", "#91D1C2FF", "#DC0000FF", "#7E6148FF", "#B09C85FF", 
                         "#E4E9B2", "#F9A620", "#054A29", "#52414C")

hm.palette <-  RColorBrewer::brewer.pal(9, "RdBu")
```

```{r setup-2, warning=FALSE, message=FALSE}
dir.create("QC", showWarnings = F)
dir.create("Processed-Data", showWarnings = F)
dir.create("Alpha-Diversity", showWarnings = F)
dir.create("Beta-Diversity", showWarnings = F)
dir.create("Taxonomic-Composition", showWarnings = F)
dir.create("Differential-Abundance", showWarnings = F)
dir.create("Other", showWarnings = F)
```

## Load data

This code chunk reads the input files and creates a phyloseq object.

```{r load-data, warning=FALSE, message=FALSE, echo=FALSE}
ps <- import_as_pseq(asvtab = paste0(data_dir, "dada2_out/ASV_seqtab_tax.tab"),
                mapping = paste0(data_dir, "metadata_AhR_KO_16S.txt"),
                tree = paste0(data_dir, "dada2_out/ASV_tree.tre"))
```

## QC {.tabset}

### Distribution of reads

```{r, warning=FALSE, message=FALSE, echo=FALSE}
qc_plot <- plot_read_distribution(ps, groups ="group", 
    plot.type = "density") + theme_cowplot()
qc_plot
ggsave("./QC/ReadDistribution.pdf")
```

### Library sizes (histogram)

```{r, warning=FALSE, message=FALSE, echo=FALSE}
pdf("QC/Library_size.pdf")
barplot(sort(sample_sums(ps)), horiz = TRUE, las = 2, xlab=NULL, main="Library sizes")
hist(sort(sample_sums(ps)), las = 2)
dev.off()
```

### Rarefaction curve

```{r,warning=FALSE, message=FALSE, echo=FALSE}
pdf("QC/rarefaction_curve.pdf")
mat <- otu_table(ps)
class(mat) <- "matrix"
rcurve <- vegan::rarecurve(t(mat), step=50, cex=0.5)
dev.off()
```

### ASV variance

```{r, warning=FALSE, message=FALSE, echo=FALSE}
Variance.plot <- qplot(log10(apply(otu_table(ps), 1, var)), 
                         xlab = "log10(variance)", 
                         main = "Variance in OTUs") + 
  ggtitle("Before filtering") + theme_cowplot()

print(Variance.plot)

ggsave("QC/Variance.pdf")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
ps_filt <- prune_taxa(taxa_sums(ps) > 5000, ps)

Variance.plot <- qplot(log10(apply(otu_table(ps_filt), 1, var)), 
                         xlab = "log10(variance)", 
                         main = "Variance in OTUs") + 
  ggtitle("After filtering") + theme_cowplot()

print(Variance.plot)

ggsave("QC/Variance_filtered.pdf")
```

Remember to adjust normalisation with the filtered phyloseq object here if necessary!

## Normalisation {.tabset}

```{r, warning=FALSE, message=FALSE, echo=FALSE}
ps_norm <- transform(ps, transform = "mss")
ps_rel <- transform(ps, transform = "relative")
save(ps_norm, ps_rel, file = "Processed-Data/normalised_ps.Rdata")
```

```{r}
ps_naive_wt <- subset_samples(ps, group != "KO")
ps_naive_ko <- subset_samples(ps, group != "WT")
ps_wt_ko <- subset_samples(ps, group != "naive")
```

## Alpha diversity {.tabset}

### All groups

```{r fig.height=6, fig.width=4, warning=FALSE, message=FALSE, echo=FALSE}
alpha_div <- calc_alpha(ps_norm)
meta <-meta_to_df(ps_norm)
# merge alpha diversity df and metadata by rownames (0)
alpha_div_meta <- merge(alpha_div, meta, by=0)
```

```{r}
comps <- microbiomeutilities::make_pairs(alpha_div_meta$group)
colours <- cat.palette[c(2,3,4)]
names(colours) <- c("KO", "naive", "WT")
metrics <- c("Richness", "Shannon.Effective", "Faiths.PD")
```

```{r}
for (i in metrics){
  p <- plot_boxplot(df = alpha_div_meta, variable_col = "group", value_col = i, 
             fill_var = "group", comparisons_list = comps, xlab = "group", 
             ylab = i, p_title = "", multiple_groups = T,
             cols = colours, group.order = c("naive","WT", "KO")) +
  theme(axis.text = element_text(size=16))
  name <- paste0("Alpha-Diversity/", gsub("\\.", "_",i), ".pdf")
  ggsave(filename = name,plot=p, device = "pdf",dpi = 300, height=4, width=4.25)
  print(p)
}

```

### KO vs WT

AhR KO and WT mice were from the same litter, however naive were housed separately.

```{r}
ps_wt_ko <- subset_samples(ps_norm, group != "naive")
```

```{r}
alpha_div_ahr <- calc_alpha(ps_wt_ko)
meta_ahr <-meta_to_df(ps_wt_ko)
# merge alpha diversity df and metadata by rownames (0)
alpha_div_ahr <- merge(alpha_div_ahr, meta_ahr, by=0)
```

```{r}
for (i in metrics){
  p <- plot_boxplot(df = alpha_div_ahr, variable_col = "group", value_col = i, 
             fill_var = "group", comparisons_list = list(c("WT", "KO")), xlab = "group", 
             ylab = i, p_title = "", 
             cols = colours, group.order = c("WT", "KO")) +
  theme(axis.text = element_text(size=16))
  name <- paste0("Alpha-Diversity/", gsub("\\.", "_",i), "_ahr.pdf")
  ggsave(filename = name,plot=p, device = "pdf",dpi = 300, height=4, width=4.25)
  print(p)
}
```

## Beta Diversity (Ordinations) {.tabset}

#### Bray-Curtis distance PCoA

```{r BC-NMDS, fig.height=4, fig.width=8, warning=FALSE, message=FALSE, echo=FALSE}
bc <- calc_betadiv(ps_norm, dist = "bray", ord_method = "NMDS")
plot_bc <- plot_beta_div(ps_norm, dist_matrix = bc$Distance_Matrix, ordination = bc$Ordination, 
              group_variable = "group", cols = colours)
ggsave("Beta-Diversity/bray-curtis_plot.pdf", plot = plot_bc, device = "pdf", height=5, width=6)
```

#### GUnifrac distance PCoA

fix issue with fig size

```{r gunifrac-NMDS, fig.height=4, fig.width=8, warning=FALSE, message=FALSE, echo=FALSE}
gu <- calc_betadiv(ps_norm, dist = "gunifrac", ord_method = "NMDS")
p_gu <- plot_beta_div(ps_norm,dist_matrix = gu$Distance_Matrix, ordination = gu$Ordination, 
              group_variable = "group", cols = colours)
ggsave("Beta-Diversity/gunifrac_plot.pdf", plot = p_gu, device = "pdf", height=5, width=6)
```

#### Aitchison 

```{r}
aitch <- calc_betadiv(ps, dist = "aitchison", ord_method = "NMDS")
p_aitch <- plot_beta_div(ps,dist_matrix = aitch$Distance_Matrix, ordination = aitch$Ordination, 
              group_variable = "group", cols = colours)
p_aitch
#ggsave("Beta-Diversity/aitchison_plot.pdf", plot = p_aitch, device = "pdf", height=5, width=6)
```
### AhR mice only - Betadiv

#### Bray-Curtis distance PCoA

```{r BC-NMDS, fig.height=4, fig.width=8, warning=FALSE, message=FALSE, echo=FALSE}
bc_ahr <- calc_betadiv(ps_wt_ko, dist = "bray", ord_method = "NMDS")
p_bc_ahr <- plot_beta_div(ps_wt_ko, dist_matrix = bc_ahr$Distance_Matrix, ordination = bc_ahr$Ordination, 
              group_variable = "group", cols = colours)
ggsave("Beta-Diversity/bray-curtis_plot_ahr.pdf", plot = p_bc_ahr, device = "pdf", height=5, width=6)
```

#### GUnifrac distance PCoA

fix issue with fig size

```{r gunifrac-NMDS, fig.height=4, fig.width=8, warning=FALSE, message=FALSE, echo=FALSE}
gu_ahr <- calc_betadiv(ps_wt_ko, dist = "gunifrac", ord_method = "NMDS")
p_gu_ahr <- plot_beta_div(ps_wt_ko,dist_matrix = gu_ahr$Distance_Matrix, ordination = gu_ahr$Ordination, 
              group_variable = "group", cols = colours)
ggsave("Beta-Diversity/gunifrac_plot_ahr.pdf", plot = p_gu_ahr, device = "pdf", height=5, width=6)
```

#### Aitchison 

```{r}
aitch <- calc_betadiv(ps_wt_ko, dist = "aitchison", ord_method = "NMDS")
p_aitch <- plot_beta_div(ps_wt_ko,dist_matrix = aitch$Distance_Matrix, ordination = aitch$Ordination, 
              group_variable = "group", cols = colours)
p_aitch
ggsave("Beta-Diversity/aitchison_plot_ahr.pdf", plot = p_aitch, device = "pdf", height=5, width=6)
```
## Taxonomic composition {.tabset}

Batch 2 -

remaining analyses could focus on differential taxa -

```{r}
taxa <- c("Bacteroidaceae", "Muribaculaceae", "Prevotellaceae", "Rikenellaceae", "Bacteroidales Order", "Bacteria Kingdom", "Bacteroidia Class", "Lactobacillaceae", "Lachnospiraceae", "Oscillospiraceae", "Ruminococcaceae", "Akkermansiaceae")

tax_col <- c(taxa,"other")

pal_fig <- pal_npg()(10)
pal_fig <- c(pal_fig, "#E4E9B2", "#F9A620","#52414C", "#054A29", "#DCDCDC")
names(pal_fig) <- tax_col
ps_norm_f <- phylosmith::set_sample_order(ps_wt_ko, treatment = "group") %>% 
  aggregate_taxa(level="Family")

fig <- ps_norm_f %>% 
  tax_fix(unknowns = c("")) %>%
  ps_arrange(desc(Akkermansiaceae), .target = "otu_table") %>% 
  comp_barplot(
    tax_level = "Family", n_taxa = 10,
    bar_outline_colour = "black", bar_width = 0.9, palette = pal_fig) + 
  theme_cowplot() +
  facet_grid(.~factor(group, rev(stringr::str_sort(unique(group)))),
    scales = "free", space = "free" # these options are critically important!
  ) + 
  scale_y_continuous(labels = scales::percent, expand = c(0,0)) +
  theme(
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=22),
        axis.text.y = element_text(size=18),
        axis.ticks = element_blank())
fig + 
  rotate_x_text(angle=45)
ggsave2("Taxonomic-Composition/genus_level_comp10.pdf",, device = "pdf", height=5, width=6)
```

## Differential abundance {.tabset}



WT vs naive

```{r, warning=FALSE, message=FALSE, echo=FALSE}
res_wt_naive <- ancombc2(
      data = format_taxonomy(ps_naive_wt),
      fix_formula = "group",
      p_adj_method = "BH",
      prv_cut = 0.33,
      group = "group",
      struc_zero = TRUE,
      neg_lb = FALSE,
      pseudo_sens = T,
      global = FALSE
    )
```

```{r}
plot_da(ancom_res, groups=c("Sham", "Stroke"), cols = colours)
```

KO vs naive

```{r, warning=FALSE, message=FALSE, echo=FALSE}
res_ko_naive <- ancombc2(
      data = format_taxonomy(ps_naive_ko),
      fix_formula = "group",
      p_adj_method = "BH",
      prv_cut = 0.33,
      group = "group",
      struc_zero = TRUE,
      neg_lb = FALSE,
      pseudo_sens = T,
      global = FALSE
    )
```

```{r}
plot_da(ancom_res, groups=c("Sham", "Stroke"), cols = colours)
```

WT vs KO

No global difference between WT and AhR KO

```{r, warning=FALSE, message=FALSE, echo=FALSE}
res_wt_ko <- ancombc2(
      data = format_taxonomy(ps_wt_ko),
      fix_formula = "group",
      p_adj_method = "BH",
      prv_cut = 0.33,
      group = "group",
      struc_zero = TRUE,
      neg_lb = FALSE,
      pseudo_sens = T,
      global = FALSE
    )
```

No Entero detected 

```{r session-info}

sessionInfo()

```