---
title: "Ileal analysis"
author: "Adam Sorbie"
date: "2026-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of Ileal data from AHR project 


Key figures from this are: 

Donor composition barplots
NMDS plot
colonization heatmap 
any associations with Enterobacteriaceae ASVs and indole/ DCs 
```{r}
source("../scripts/analysis_functions.R")
source("../scripts/utils.R")
library(microViz)
library(microbiome)
library(ggsci)
library(cowplot)
library(dendextend)
library(gridExtra)
library(ggplotify)
library(grid)
library(Tjazi)
conflicted::conflict_prefer_all("dplyr")
```

```{r}
cols_hfmt <- c("hFMT 14"= "#797979","hFMT 15" = "#FFB92E", "patient_11" = "#1A1F16", "patient_15"= "#E3170A")
data_dir <- "../data/2025/dada2_out/"

ps <- import_as_pseq(paste0(data_dir, "ASV_seqtab_tax.tab"),
                     "../data/2025/metadata.txt", 
                     paste0(data_dir, "ASV_tree.tre"))

```

QC 

```{r}
pdf("figures/Library_size.pdf")
barplot(sort(sample_sums(ps)), horiz = TRUE, las = 2, xlab=NULL, main="Library sizes")
hist(sort(sample_sums(ps)), las = 2)
dev.off()
```

Remove two additional human samples
```{r}
ps_exc_patients <- ps %>% 
  subset_samples(Donor %in% c("patient_4", "patient_6"))

ps_all <- ps
ps <- ps %>% 
  subset_samples(!Donor %in% c("patient_4", "patient_6"))



# exclude CB6 - high levels of chloroplast and mitochondrial sequences and low library size
ps <- ps %>% 
  subset_samples(OriginalID != "d21_H7")
```

```{r}
ps_norm_mouse <- subset_samples(ps, Source == "Mouse") %>% 
  transform()
ps_norm_human <- subset_samples(ps, Source == "Human") %>% 
  transform()
ps_norm_all <- transform(ps_all)
ps_norm <- transform(ps)
```

What questions do we want to answer here? 

a) What is the composition and diversity of the human samples?
b) What is the transfer efficiency? 
c) Are the two groups different in terms of diversity? 
d) Which taxa differ between the two groups? 
e) are these likely to be indole producing taxa? 

```{r}
div_human <- calc_alpha(ps_norm_human, detection = 1)
div_human_meta <- merge(div_human, meta_to_df(ps_norm_human), by=0)
```

### Composition 

```{r}
pal_fig <- pal_npg()(10)
pal_fig <- c(pal_fig,"#DCDCDC")
```

```{r}
ps_norm_human_all <- subset_samples(ps_norm_all, Source == "Human")
ps_norm_human_f <- ps_reorder(ps_norm_human_all, sample_order = c("CB12",  "P11","P4", "CB13","P15", "P6")) %>% 
  aggregate_taxa(level="Family")

fig <- ps_norm_human_f %>% 
  tax_fix(unknowns = c("")) %>%
  comp_barplot(
    tax_level = "Family", n_taxa = 10,
    bar_outline_colour = "black", bar_width = 0.9, palette = pal_fig) + #
      facet_grid(~Severity, scales="free_x", space="free_x") +
  theme_cowplot() +
  scale_y_continuous(labels = scales::percent, expand = c(0,0)) +
  theme(
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=22),
        axis.text.y = element_text(size=18),
        axis.ticks = element_blank(),
        strip.background = element_rect(fill="white", color="black"))
fig 
ggsave("figures/donors_comp_all.pdf", fig, device = "pdf", dpi=300, height = 7.5, width = 6)
```

For excluded patients
```{r}
ps_norm_human_exc <- transform(ps_exc_patients)
ps_norm_human_exc_f <- ps_reorder(ps_norm_human_exc, sample_order = c("P4",  "P6")) %>% 
  aggregate_taxa(level="Family")

fig <- ps_norm_human_exc_f %>% 
  tax_fix(unknowns = c("")) %>%
  comp_barplot(
    tax_level = "Family", n_taxa = 10,
    bar_outline_colour = "black", bar_width = 0.9, palette = pal_fig) + #
      facet_grid(~Severity, scales="free_x", space="free_x") +
  theme_cowplot() +
  scale_y_continuous(labels = scales::percent, expand = c(0,0)) +
  theme(
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=22),
        axis.text.y = element_text(size=18),
        axis.ticks = element_blank(),
        strip.background = element_rect(fill="white", color="black"))
fig 
ggsave("figures/donors_comp_exc.pdf", fig, device = "pdf", dpi=300, height = 7.5, width = 6)
```

### Transfer efficiency

How to best answer this? 


Get percentages first - stratified by donor 

number of ASVs in donor sample above 0 - baseline 
for each mouse get number of ASVs above 0, subset by donor list and calc percentage
```{r}
asvtab <- ps_to_asvtab(ps_norm)
p14_asvs <- asvtab %>% select(CB12) %>% filter(CB12 > 0) %>% rownames()
p15_asvs <- asvtab %>% select(CB13) %>% filter(CB13 > 0) %>% rownames()

p11_asvs <- asvtab %>% select(P11) %>% filter(P11 > 0) %>% rownames()
p15b_asvs <- asvtab %>% select(P15) %>% filter(P15 > 0) %>% rownames()
```

#### Looking specifically at those that colonise

```{r}
ps_hFMT14 <- subset_samples(ps_norm, Donor =="hFMT 14")
ps_hFMT15 <- subset_samples(ps_norm, Donor =="hFMT 15")
ps_p11 <- subset_samples(ps_norm, Donor =="patient_11")
ps_p15 <- subset_samples(ps_norm, Donor =="patient_15")
```


```{r}
get_colonised <- function(ps, donor, recipients) {
  asvtab <- ps_to_asvtab(ps)
  # select donor sample
  asvtab_donor <- asvtab[donor]
  # select mouse recipients
  asvtab_recipients <- asvtab[recipients]
  # filter to ASVs present (>0) in donor 
  asvs_donor <- asvtab_donor %>% filter(rowSums(.) > 0)
  # filter the recipients by those present in donor by name
  asvs_recipients_donor <- filter_rownames(asvtab_recipients, rownames(asvs_donor))
  # finally filter by abundance (>0) keeping only those ASVs found in both donor and recipient
  # with a abundance >0 in at least one recipient 
  asvtab_recipients <- asvs_recipients_donor %>% filter(rowSums(.) > 0)

  asvtab_out <- asvtab %>%
    filter_rownames(rownames(asvtab_recipients))

  return(asvtab_out)
}
```

```{r}
meta_hFMT14 <- meta_to_df(ps_hFMT14)
mice_hFMT14 <- meta_hFMT14 %>% filter(Source == "Mouse")
asvtab_col_hFMT14 <- get_colonised(ps_hFMT14, donor="CB12", recipients = rownames(mice_hFMT14))

colonisation_efficiencyhFMT14 <- round(nrow(asvtab_col_hFMT14) / length(p14_asvs) * 100, digits = 2)
```

```{r}
meta_hFMT15 <- meta_to_df(ps_hFMT15)
mice_hFMT15 <- meta_hFMT15 %>% filter(Source == "Mouse")
asvtab_col_hFMT15 <- get_colonised(ps_hFMT15, donor="CB13", recipients = rownames(mice_hFMT15))
colonisation_efficiencyhFMT15 <-round(nrow(asvtab_col_hFMT15) / length(p15_asvs) * 100, digits = 2)
```

```{r}
meta_p11 <- meta_to_df(ps_p11)
mice_p11 <- meta_p11 %>% filter(Source == "Mouse")
asvtab_col_p11 <- get_colonised(ps_p11, donor="P11", recipients = rownames(mice_p11))
colonisation_efficiencyp11 <- round(nrow(asvtab_col_p11) / length(p11_asvs) * 100, digits = 2)
```

```{r}
meta_p15 <- meta_to_df(ps_p15)
mice_p15 <- meta_p15 %>% filter(Source == "Mouse")
asvtab_col_p15 <- get_colonised(ps_p15, donor="P15", recipients = rownames(mice_p15))
colonisation_efficiencyp15 <- round(nrow(asvtab_col_p15) / length(p15b_asvs) * 100, digits = 2)
```
```{r}
ps_hFMT14_hm <- prune_taxa(rownames(asvtab_col_hFMT14), ps_hFMT14)
ps_hFMT15_hm <- prune_taxa(rownames(asvtab_col_hFMT15), ps_hFMT15)
ps_p11_hm <- prune_taxa(rownames(asvtab_col_p11), ps_p11)
ps_p15_hm <- prune_taxa(rownames(asvtab_col_p15), ps_p15)

ps_hm <- prune_taxa(c(rownames(asvtab_col_hFMT14), rownames(asvtab_col_hFMT15), rownames(asvtab_col_p11), rownames(asvtab_col_p15)), ps_norm)
```

#### With all colonising ASVs

```{r}
mean_col_eff <- mean(c(colonisation_efficiencyhFMT14, colonisation_efficiencyhFMT15,
                          colonisation_efficiencyp11, colonisation_efficiencyp15))

sink(file = "colonisation-efficiency.txt")
paste("Mean: ", mean_col_eff)
paste("hFMT14: ", colonisation_efficiencyhFMT14)
paste("hFMT15: ", colonisation_efficiencyhFMT15)
paste("P11: ", colonisation_efficiencyp11)
paste("P15: ", colonisation_efficiencyp15)
sink(file = NULL)
```

Heatmap per donor
```{r}
pal <- RColorBrewer::brewer.pal(9, "PuBuGn")
list_hm <- list(ps_hFMT14, ps_p11, ps_p15, ps_hFMT15)

list_hm <- map(list_hm, function(x){
  ps_hm_sub <- ps_arrange(x, hmap_ord)
  p <- microbiomeutilities::plot_taxa_heatmap(ps_hm_sub, subset.top = 40, 
                                         transformation = "clr", VariableA = "hmap_ord", 
                                         heatcolors = pal,
                                         border_color=NA, cluster_cols=F, show_colnames=F)
  return(as.grob(p$plot$gtable))
})

names(list_hm) <- c("hFMT14", "p11", "p15", "hFMT15")

ggsave("figures/hm_colonisers-separate_donors-2025.pdf", grid.arrange(grobs=list_hm,ncol=4), device = "pdf", dpi=300, width = 32, height = 7.5)
```

Combined heatmap 
```{r}

ps_hm <- ps_arrange(ps_hm, hmap_ord)
p <- microbiomeutilities::plot_taxa_heatmap(ps_hm, subset.top = 40, 
                                       transformation = "clr", VariableA = "hmap_ord", 
                                       heatcolors = pal, 
                                       border_color=NA, cluster_cols=F, show_colnames=F)
ggsave("figures/hm_colonisers-both_groups-2025.pdf", p$plot$gtable, device = "pdf", dpi=300, width = 12, height = 5.5)
```


### Diversity - colonised mice 


```{r}
div_mice <- calc_alpha(ps_norm_mouse)
div_mice_meta <- merge(div_mice, meta_to_df(ps_norm_mouse), by=0)

comps <- list(unique(div_mice_meta$Donor))

for (i in c("Richness", "Shannon.Effective", "Faiths.PD")){
  p <- plot_boxplot(div_mice_meta, variable_col = "Donor", value_col = i, 
             comparisons_list = comps, cols = cols_hfmt, multiple_groups=T)
  if (i == "Faiths.PD"){
    p <- p + ylim(0, 8)
  }
  print(p)
  
}


```


#### Beta-diversity inc. human samples 

bray 
```{r}
betadiv_bray_all <- calc_betadiv(ps_norm, dist="bray", ord_method = "NMDS")
plot_beta_div(ps_norm, betadiv_bray_all$Distance_Matrix, betadiv_bray_all$Ordination,group_variable =  "Donor", size=4, shape = "Source", cols = cols_hfmt) 
ggsave("figures/betadiv_all-bray.pdf", device = "pdf", dpi=300, height = 6, width = 8)
```

unifrac 

```{r}
betadiv_uf_all <- calc_betadiv(ps_norm, dist="unifrac", ord_method = "NMDS")
plot_beta_div(ps_norm, betadiv_uf_all$Distance_Matrix, betadiv_uf_all$Ordination, group_variable = "Donor", size = 4, shape = "Source", cols = cols_hfmt) 
ggsave("figures/betadiv_all-unifrac.pdf", device = "pdf", dpi=300, height = 6, width = 8)
```
wunifrac 

```{r}
betadiv_wf_all <- calc_betadiv(ps_norm, dist="wunifrac", ord_method = "NMDS")
plot_beta_div(ps_norm, betadiv_wf_all$Distance_Matrix, betadiv_wf_all$Ordination, , group_variable = "Donor", size = 4, shape = "Source", cols = cols_hfmt) 
ggsave("figures/betadiv_all-wunifrac.pdf", device = "pdf", dpi=300, height = 6, width = 8)
```
gunifrac

```{r}
betadiv_gf_all <- calc_betadiv(ps_norm, dist="gunifrac", ord_method = "NMDS")
plot_beta_div(ps_norm, betadiv_gf_all$Distance_Matrix, betadiv_gf_all$Ordination, group_variable = "Donor", size = 4, shape = "Source", cols = cols_hfmt) 
ggsave("figures/betadiv_all-gunifrac.pdf", device = "pdf", dpi=300, height = 6, width = 8)
```

compositional 

```{r}
betadiv_aitch_all <- calc_betadiv(ps_norm, dist="aitchison", ord_method = "NMDS")
plot_beta_div(ps_norm, betadiv_aitch_all$Distance_Matrix, betadiv_aitch_all$Ordination, , group_variable = "Donor", size = 4, shape = "Source", cols = cols_hfmt) 
ggsave("figures/betadiv_all-aitchison.pdf", device = "pdf", dpi=300, height = 6, width = 8)
```

### Hierarchical clustering

Compositional - all data points (inc. donors)


```{r}
#Save as dendrogram
ward <- as.dendrogram(hclust(betadiv_aitch_all$Distance_Matrix, method = "ward.D2"))
#Provide color codes
meta <- data.frame(phyloseq::sample_data(ps_norm))
#colorCode <- c(`hFMT 15` = "#E64B35FF", `hFMT 14` = "#4DBBD5FF")
labels_colors(ward) <- cols_hfmt[meta$Donor][order.dendrogram(ward)]
#Plot
pdf("figures/dendrogram_hclust.pdf", width = 7, height = 4)
plot(ward)
dev.off()
```


### Differential abundance

Prep data

```{r}
ps_mouse_clr <- ps_norm_mouse %>%
  transform("relative") %>% 
  format_taxonomy() %>% 
  microViz::tax_filter(min_prevalence = 0.33, min_total_abundance = 0) 
  
#   %>%
#   ps_get() 
#   %>% 
#   transform("clr")

# ps_mouse_clr_g <- ps_norm_mouse %>% 
#   format_taxonomy() %>% 
#   aggregate_taxa(level = "Genus") %>%
#   microViz::tax_filter(min_prevalence = 0.33, min_total_abundance = 0) %>% 
#   transform("clr")


# ps_mouse_clr_f <- ps_norm_mouse %>% 
#   format_taxonomy() %>% 
#   aggregate_taxa(level = "Family") %>%
#   microViz::tax_filter(min_prevalence = 0.33, min_total_abundance = 0) %>% 
#   transform("clr")
```

```{r}
mouse_counts_clr <- ps_to_asvtab(ps_mouse_clr)
metadata_mouse <- meta_to_df(ps_mouse_clr)
da_tax_clr <- fw_glm(x= mouse_counts_clr,
                     f= ~ Donor, 
                     metadata = metadata_mouse,
                     adjust.method = "BH", 
                     format = "brief")

da_tax_clr_sig <- da_tax_clr %>% 
  filter(`DonorhFMT 15 Pr(>|t|).BH` <= 0.05) %>% 
  mutate(value = case_when(`DonorhFMT 15 Estimate` < 0 ~ "hFMT 14",
                           TRUE ~ "hFMT 15"))
```

Diverging lollipop of beta values

Add tnaA prevalence 
```{r}
da_tax_clr_sig  %>%
    arrange(`DonorhFMT 15 Estimate`) %>%
    mutate(feature = factor(feature, levels = feature)) %>% 
    ggplot(
           aes(x = feature, y = `DonorhFMT 15 Estimate`, color = value)) +
    geom_point(size = 5) +
    labs(y = paste("beta \n", "hFMT 14 vs hFMT 15",sep =
                     " "),
         x = "") +
    theme_bw() +
    theme(
      axis.text.x = element_text(color = "black", size = 14),
      axis.text.y = element_text(color = "black", size = 14),
      axis.title.y = element_text(size = 16),
      axis.title.x = element_text(size = 16),
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 14),
      legend.position = "none"
    ) +
    coord_flip() +
  theme_cowplot() +
    geom_hline(yintercept = 0, linetype = "dotted") +
    scale_color_manual(values = cols_hfmt)
ggsave("figures/differential_abundance_donor.pdf", device = "pdf", dpi=300, height = 5, width = 7)
```
#### Genus level

```{r}
da_tax_g_clr <- fw_glm(x= ps_to_asvtab(ps_mouse_clr_g),
                     f= ~ Donor, 
                     metadata = meta_to_df(ps_mouse_clr_g),
                     adjust.method = "BH", 
                     format = "brief")

da_tax_clr_g_sig <- da_tax_g_clr %>% 
  filter(`DonorhFMT 15 Pr(>|t|).BH` <= 0.1) %>% 
  mutate(value = case_when(`DonorhFMT 15 Estimate` < 0 ~ "hFMT 14",
                           TRUE ~ "hFMT 15"))
```

Diverging lollipop of beta values

Add tnaA prevalence 
```{r}
da_tax_clr_g_sig  %>%
    arrange(`DonorhFMT 15 Estimate`) %>%
    mutate(feature = factor(feature, levels = feature)) %>% 
    ggplot(
           aes(x = feature, y = `DonorhFMT 15 Estimate`, color = value)) +
    geom_point(size = 5) +
    labs(y = paste("beta \n", "hFMT 14 vs hFMT 15",sep =
                     " "),
         x = "") +
    theme_bw() +
    theme(
      axis.text.x = element_text(color = "black", size = 14),
      axis.text.y = element_text(color = "black", size = 14),
      axis.title.y = element_text(size = 16),
      axis.title.x = element_text(size = 16),
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 14),
      legend.position = "none"
    ) +
    coord_flip() +
  theme_cowplot() +
    geom_hline(yintercept = 0, linetype = "dotted") +
    scale_color_manual(values = cols_hfmt)
ggsave("figures/differential_abundance_donor_genus.pdf", device = "pdf", dpi=300, height = 5, width = 7)
```

#### Family level


```{r}
da_tax_f_clr <- fw_glm(x= ps_to_asvtab(ps_mouse_clr_f),
                     f= ~ Donor, 
                     metadata = meta_to_df(ps_mouse_clr_f),
                     adjust.method = "BH", 
                     format = "brief")

da_tax_clr_f_sig <- da_tax_f_clr %>% 
  filter(`DonorhFMT 15 Pr(>|t|).BH` <= 0.25) %>% 
  mutate(value = case_when(`DonorhFMT 15 Estimate` < 0 ~ "hFMT 14",
                           TRUE ~ "hFMT 15"))
```   

```{r}
da_tax_clr_f_sig  %>%
    arrange(`DonorhFMT 15 Estimate`) %>%
    mutate(feature = factor(feature, levels = feature)) %>% 
    ggplot(
           aes(x = feature, y = `DonorhFMT 15 Estimate`, color = value)) +
    geom_point(size = 5) +
    labs(y = paste("beta \n", "hFMT 14 vs hFMT 15",sep =
                     " "),
         x = "") +
    theme_bw() +
    theme(
      axis.text.x = element_text(color = "black", size = 14),
      axis.text.y = element_text(color = "black", size = 14),
      axis.title.y = element_text(size = 16),
      axis.title.x = element_text(size = 16),
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 14),
      legend.position = "none"
    ) +
    coord_flip() +
  theme_cowplot() +
    geom_hline(yintercept = 0, linetype = "dotted") +
    scale_color_manual(values = cols_hfmt)
ggsave("figures/differential_abundance_donor_family.pdf", device = "pdf", dpi=300, height = 5, width = 7)
```


### Correlations

Correlate clr-transformed taxon abundance with a) urinary indole concentration b) flow cytometry data 

P-value less than 0.05 and correlation threshold higher than or equal to 0.3
```{r}
cor_indole <- ps_mouse_clr %>% 
  ps_melt() %>% 
  group_by(OTU) %>% 
  rstatix::cor_test(Abundance, Urinary_indole, method = "spearman") %>% 
  filter(p < 0.05 & abs(cor) >=  0.3)

cor_mig_dc <- ps_mouse_clr %>% 
  ps_melt() %>% 
  group_by(OTU) %>% 
  rstatix::cor_test(Abundance, CD11cposMHCIIposCD11bposCD103pos, method = "spearman") %>% 
  filter(p < 0.05 & abs(cor) >= 0.3) 

cor_mig_dc1 <- ps_mouse_clr %>% 
  ps_melt() %>% 
  group_by(OTU) %>% 
  rstatix::cor_test(Abundance, CD11cHighMHCIIpos, method = "spearman") %>% 
  filter(p < 0.05 & abs(cor) >= 0.3)

cor_mig_dc2 <- ps_mouse_clr %>% 
  ps_melt() %>% 
  group_by(OTU) %>% 
  rstatix::cor_test(Abundance, CD11cposMHCIIpos.FITC, method = "spearman") %>% 
  filter(p < 0.05 & abs(cor) >= 0.3)

cor_mig_dc3 <- ps_mouse_clr %>% 
  ps_melt() %>% 
  group_by(OTU) %>% 
  rstatix::cor_test(Abundance, CD11cHighMHCIIpos.1, method = "spearman") %>% 
  filter(p < 0.05 & abs(cor) >= 0.3)
```

Get overlap between significant cors and DA microbes enriched in hFMT15 

```{r}
hFMT_feat <- filter(da_tax_clr_sig, `DonorhFMT 15 Estimate` > 0) %>% pull(feature)
indole_da_overlap <- intersect(cor_indole$OTU, hFMT_feat)
dc_da_overlap <- intersect(cor_mig_dc$OTU, hFMT_feat)
dc1_da_overlap <- intersect(cor_mig_dc1$OTU, hFMT_feat)
dc2_da_overlap <- intersect(cor_mig_dc2$OTU, hFMT_feat)
dc3_da_overlap <- intersect(cor_mig_dc3$OTU, hFMT_feat)

all_int_cors <- Reduce(union, list(indole_da_overlap, dc_da_overlap, 
                      dc1_da_overlap, dc2_da_overlap, dc3_da_overlap)) %>% 
  print()
```

Final corrplots

```{r}
plot_cor <- function(df, x, xlab, y, cor.pos) {
  p <- df %>%
    filter(OTU == y) %>%
    ggscatter(
      x = x,
      y = "Abundance",
      size = 3,
      title = y,
      color = "Donor",
      xlab = xlab,
      ylab = "CLR-transformed abundance"
    ) +
    stat_cor(label.x = cor.pos) +
    scale_color_manual(values = cols_hfmt) +
    scale_fill_manual(values = cols_hfmt) +
    theme_cowplot(font_size = 15) +
    theme(plot.title = element_text(hjust = 0.5))
  print(p)
}
```


"ASV_16; Bacteroides"  
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="Urinary_indole", xlab = "Urinary Indole [μM]", y="ASV_16; Bacteroides", cor.pos=30)
ggsave("figures/ASV_16-Bacteroides-indole-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="CD11cHighMHCIIpos.1", xlab = "CD11CHighMHCII+ [%]", y="ASV_16; Bacteroides", cor.pos = 5)
ggsave("figures/ASV_16-Bacteroides-CD11chighMHCII+-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="CD11cposMHCIIpos.FITC", xlab = "CD11c+MHCII+ [FITC]", y="ASV_16; Bacteroides", cor.pos=40000)
ggsave("figures/ASV_16-Bacteroides-CD11cMHCII-fitc-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```


"ASV_41; Bacteroides"

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="Urinary_indole", xlab = "Urinary Indole [μM]", y="ASV_41; Bacteroides", cor.pos = 30)
ggsave("figures/ASV_41-Bacteroides-indole-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
 plot_cor(x="CD11cHighMHCIIpos.1", xlab = "CD11CHighMHCII+ [%]", y="ASV_41; Bacteroides", cor.pos = 5)
ggsave("figures/ASV_41-Bacteroides-CD11chighMHCII+-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
 plot_cor(x="CD11cposMHCIIpos.FITC", xlab = "CD11c+MHCII+ [FITC]", y="ASV_41; Bacteroides", cor.pos = 40000)
ggsave("figures/ASV_41-Bacteroides-CD11cMHCII-fitc-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

"ASV_88; Bacteroides"

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="Urinary_indole", xlab = "Urinary Indole [μM]", y="ASV_88; Bacteroides", cor.pos = 30)
ggsave("figures/ASV_88-Bacteroides-indole-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="CD11cHighMHCIIpos.1", xlab = "CD11CHighMHCII+ [%]", y="ASV_88; Bacteroides", cor.pos = 5)
ggsave("figures/ASV_88-Bacteroides-CD11chighMHCII+-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="CD11cposMHCIIpos.FITC", xlab = "CD11c+MHCII+ [FITC]", y="ASV_88; Bacteroides", cor.pos = 40000)
ggsave("figures/ASV_88-Bacteroides-CD11cMHCII-fitc-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

"ASV_26; Phascolarctobacterium"
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
    plot_cor(x="Urinary_indole", xlab = "Urinary Indole [μM]", y="ASV_26; Phascolarctobacterium", cor.pos = 30)
ggsave("figures/ASV_26-Phascolarctobacterium-indole-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="CD11cHighMHCIIpos.1", xlab = "CD11CHighMHCII+ [%]", y="ASV_26; Phascolarctobacterium", cor.pos = 5)
ggsave("figures/ASV_26-Phascolarctobacterium-CD11chighMHCII+-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="CD11cposMHCIIpos.FITC", xlab = "CD11c+MHCII+ [FITC]", y="ASV_26; Phascolarctobacterium", cor.pos = 40000)
ggsave("figures/ASV_26-Phascolarctobacterium-CD11cMHCII-fitc-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

"ASV_79; Lachnospiraceae UCG-001"

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
    plot_cor(x="Urinary_indole", xlab = "Urinary Indole [μM]", y="ASV_79; Lachnospiraceae UCG-001", cor.pos = 30)
ggsave("figures/ASV_79-Lachnospiraceae-UCG-001-indole-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="CD11cHighMHCIIpos.1", xlab = "CD11CHighMHCII+ [%]", y="ASV_79; Lachnospiraceae UCG-001", cor.pos = 5)
ggsave("figures/ASV_79-Lachnospiraceae-UCG-001-CD11chighMHCII+-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="CD11cposMHCIIpos.FITC", xlab = "CD11c+MHCII+ [FITC]", y="ASV_79; Lachnospiraceae UCG-001", cor.pos = 40000)
ggsave("figures/ASV_79-Lachnospiraceae-UCG-001-CD11cMHCII-fitc-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

"ASV_15; Parabacteroides"

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
    plot_cor(x="Urinary_indole", xlab = "Urinary Indole [μM]", y="ASV_15; Parabacteroides", cor.pos = 30)
ggsave("figures/ASV_15-Parabacteroides-indole-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="CD11cHighMHCIIpos.1", xlab = "CD11CHighMHCII+ [%]", y="ASV_15; Parabacteroides", cor.pos = 5)
ggsave("figures/ASV_15-Parabacteroides-CD11chighMHCII+-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
  plot_cor(x="CD11cposMHCIIpos.FITC", xlab = "CD11c+MHCII+ [FITC]", y="ASV_15; Parabacteroides", cor.pos = 40000)
ggsave("figures/ASV_15-Parabacteroides-CD11cMHCII-fitc-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

"ASV_139; Roseburia" 

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
    plot_cor(x="Urinary_indole", xlab = "Urinary Indole [μM]", y="ASV_139; Roseburia", cor.pos = 30)
ggsave("figures/ASV_139-Roseburia-indole-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
 plot_cor(x="CD11cHighMHCIIpos.1", xlab = "CD11CHighMHCII+ [%]", y="ASV_139; Roseburia", cor.pos = 5)
ggsave("figures/ASV_139-Roseburia-CD11chighMHCII+-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
 plot_cor(x="CD11cposMHCIIpos.FITC", xlab = "CD11c+MHCII+ [FITC]", y="ASV_139; Roseburia", cor.pos = 40000)
ggsave("figures/ASV_139-Roseburia-CD11cMHCII-fitc-cor.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

#### Targeted correlations - Enterobacterales/acieae 

```{r}
ps_mouse_clr <- ps_norm_mouse %>%
  transform("clr") %>% 
  format_taxonomy() 
```

##### Indole producers relative abundance

ASV_8
```{r}
ASV_8 <- ps_mouse_clr %>% 
  ps_melt() %>% 
  filter(OTU == "ASV_8; unknown_Enterobacteriaceae")
```

```{r}
sev_cols <- c("mild" = "#797979","severe" = "#FFB92E")
comps <- microbiomeutilities::make_pairs(ASV_8$Severity)
ggplot(ASV_8, aes(x = Severity, y = Abundance, fill = Severity)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(shape=Donor), width = 0.075) +
  labs(y = "CLR-transformed abundance", x = "Donor Stroke Severity") +
  theme_bw() +
  theme(
    axis.text.x = element_text(color = "black", size = 12),
    axis.text.y = element_text(color = "black", size = 12),
    axis.title.y = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.position = "none"
  ) +
  scale_fill_manual(values = sev_cols) + 
  theme_cowplot() + 
  rotate_x_text(45) + 
  stat_compare_means(comparisons = comps, method = "wilcox.test", hide.ns = TRUE)
ggsave("figures/ASV_8-Enterobacteriaceae-donor-boxplot-grouped.pdf", device = "pdf", dpi=300, height = 4, width = 4.25)
```

Other potential indole producers

```{r}
indole_producers <- c("Enterobacteriaceae", "Escherichia", "Akkermansia", "Bacteroides", "Bifidobacterium", "Oscillibacter", "Klebsiella", "Butyricicoccus")
indole_producers_asvs <- ps_mouse_clr %>% 
  ps_melt() %>% 
  filter(grepl(paste(indole_producers, collapse="|"), OTU)) %>% 
  distinct(OTU) %>% 
  pull(OTU)
```

```{r}
map(indole_producers_asvs, function(x) {
  ASV_dat <- ps_mouse_clr %>% 
    ps_melt() %>% 
    filter(OTU == x) 

  ggplot(ASV_dat, aes(x = Severity, y = Abundance, fill = Severity)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(shape=Donor), width = 0.075) +
  labs(y = "CLR-transformed abundance", x = "Donor Stroke Severity") +
  theme_bw() +
  theme(
    axis.text.x = element_text(color = "black", size = 12),
    axis.text.y = element_text(color = "black", size = 12),
    axis.title.y = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.position = "none"
  ) +
  scale_fill_manual(values = sev_cols) + 
  theme_cowplot() + 
  rotate_x_text(45) + 
  stat_compare_means(comparisons = comps, method = "wilcox.test", hide.ns = TRUE)
name <- ASV_dat$OTU[1] %>% gsub(";", "-", .)
ggsave(paste0("figures/", name, "-donor-boxplot-grouped.pdf"), device = "pdf", dpi=300, height = 4, width = 4.25)
})
```

##### Correlations 

```{r}
map(indole_producers_asvs, function(x) {
  ASV_dat <- ps_mouse_clr %>% 
    ps_melt() %>% 
    filter(OTU == x) 

  ggscatter(ASV_dat,
      x = "Urinary_indole_ummol_ml",
      y = "Abundance",
      add = "reg.line",
      conf.int = T,
      size = 3,
      color = "Severity",
      xlab = "Urinary Indole [μmol/mL]",
      ylab = "CLR-transformed abundance"
    ) +
    stat_cor(aes(color=Severity), label.x = 0.35) +
    scale_color_manual(values = sev_cols) +
    scale_fill_manual(values = sev_cols) +
    theme_cowplot(font_size = 15) +
    theme(plot.title = element_text(hjust = 0.5))
name <- ASV_dat$OTU[1] %>% gsub(";", "-", .)
ggsave(paste0("figures/Indole_producers/", name, "-donor-corrplot-grouped.pdf"), device = "pdf", dpi=300, height = 4, width = 4.25)
})
```

Entero ASVs
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
    filter(grepl("Enterobacterales", OTU) | grepl("Enterobacteriaceae", OTU)) %>%
    distinct(OTU) %>%
    arrange(OTU) %>%
    print()
```

Correlations grouped by donor
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
    filter(OTU == "ASV_8; unknown_Enterobacteriaceae") %>%
    ggscatter(
      x = "Urinary_indole_ummol_ml",
      y = "Abundance",
      add = "reg.line",
      conf.int = T,
      size = 3,
      title =  "ASV_8; unknown_Enterobacteriaceae",
      color = "Donor",
      xlab = "Urinary Indole [μmol/mL]",
      ylab = "CLR-transformed abundance",
    ) +
    stat_cor(aes(color=Donor)) +
    facet_wrap(~Round, scales = "free_x") +
    scale_color_manual(values = cols_hfmt) +
    scale_fill_manual(values = cols_hfmt) +
    theme_cowplot(font_size = 15) +
    theme(plot.title = element_text(hjust = 0.5))
ggsave("figures/ASV_8-Enterobacterales-indole-cor-facet.pdf", device = "pdf", dpi=300, height = 3.75, width = 8)
```

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
    filter(OTU == "ASV_104; unknown_Enterobacteriaceae") %>%
    ggscatter(
      x = "Urinary_indole_ummol_ml",
      y = "Abundance",
      add = "reg.line",
      conf.int = T,
      size = 3,
      title = "ASV_104; unknown_Enterobacteriaceae",
      color = "Donor",
      xlab = "Urinary Indole [μmol/mL]",
      ylab = "CLR-transformed abundance"
    ) +
    stat_cor(aes(color=Donor)) +
    facet_wrap(~Round, scales = "free_x") +
    scale_color_manual(values = cols_hfmt) +
    scale_fill_manual(values = cols_hfmt) +
    theme_cowplot(font_size = 15) +
    theme(plot.title = element_text(hjust = 0.5))
ggsave("figures/ASV_104-Enterobacteriaceae-indole-cor-facet.pdf", device = "pdf", dpi=300, height = 3.75, width = 8)
```

Grouped by donor severity

```{r}
cols_sev <- c("dodgerblue","firebrick3")
```
```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
    filter(OTU == "ASV_8; unknown_Enterobacteriaceae") %>%
    ggscatter(
      x = "Urinary_indole_ummol_ml",
      y = "Abundance",
      add = "reg.line",
      conf.int = T,
      size = 3,
      title =  "ASV_8; unknown_Enterobacteriaceae",
      color = "Severity",
      xlab = "Urinary Indole [μmol/mL]",
      ylab = "CLR-transformed abundance"
    ) +
    stat_cor(aes(color=Severity), label.x = 0.35) +
    scale_color_manual(values = cols_sev) +
    scale_fill_manual(values = cols_sev) +
    theme_cowplot(font_size = 15) +
    theme(plot.title = element_text(hjust = 0.5))
ggsave("figures/ASV_8-Enterobacterales-indole-cor-sev.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```

```{r}
ps_mouse_clr %>% 
  ps_melt() %>%
    filter(OTU == "ASV_104; unknown_Enterobacteriaceae") %>%
    ggscatter(
      x = "Urinary_indole_ummol_ml",
      y = "Abundance",
      add = "reg.line",
      conf.int = T,
      size = 3,
      title = "ASV_104; unknown_Enterobacteriaceae",
      color = "Severity",
      xlab = "Urinary Indole [μmol/mL]",
      ylab = "CLR-transformed abundance"
    ) +
    stat_cor(aes(color=Severity), label.x = 0.35) +
    scale_color_manual(values = cols_sev) +
    scale_fill_manual(values = cols_sev) +
    theme_cowplot(font_size = 15) +
    theme(plot.title = element_text(hjust = 0.5))
ggsave("figures/ASV_104-Enterobacteriaceae-indole-cor-sev.pdf", device = "pdf", dpi=300, height = 3.75, width = 5.75)
```