---
title: "ahr_ileum_func"
author: "Adam Sorbie"
date: "2023-06-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of functional profiles from ileum content of sham/stroke mice - AHR project

This document contains code to analyse functional profiles produced from 150bp PE
Illumina Nextseq sequencing of DNA extracted from the ileum content of sham
and stroke (fMCAO) mice. 

Sequencing Methods overview: 

Platform: Illumina Nextseq
Adapters: Nextera PE
Paired: TRUE
Read length: 150bp
Depth: 2Gbp (approximate target)
Avg. reads per sample:14078731.33

Note that 1 sample did not pass QC (M14). 


QC overview: 

Reads were merged (simple cat), trimmed and filtered and then aligned against
the mouse/phiX genomes to filter non-microbial reads 

Trimming: Atria 
Adapters: Nextera PE
Length filtering: 75-200bp 
Quality filtering: Q20
polyG trimming: TRUE
bowtie2: very-sensitive 


## Functional profiling 

Post-QC reads were profiled using HUMAnN 3.7 against the uniref50 database. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
library(tidytext)
source("../../../Metagenomics/scripts/func_analysis.R")
meta_path <- "../data/ahr_metadata-2025.txt"
ec_path <- "../data/humann4/search_coverage_30/merged_table/unstrat/humann_genefamilies_ec_unstratified.tsv"
ko_path <- "../data/humann4/search_coverage_30/merged_table/unstrat/humann_genefamilies_ko_unstratified.tsv"
py_path <- "../data/humann4/search_coverage_30/merged_table/unstrat/humann_pathabundance_relab_unstratified.tsv"
gbm_path <- "../data/humann4/search_coverage_30/merged_table/unstrat/GBM_abundance_default.tsv"
#uniref_path <- "../data/humann_com/merged_table/unstrat/humann_genefamilies_relab_unstratified.tsv"
outdir <- "func_figures_2025"
conflicted::conflict_prefer_all("dplyr")
conflicted::conflict_prefer_all("purrr")
```

Sample exclusion

"RD275-ileum", "M-21-ileum", "RD521-ileum" may need to be excluded 

```{r}

```


```{r}
func_profile <- import_func_profile(meta = meta_path,
                                    ec = ec_path, ko = ko_path,
                                    py = py_path, filter_ungrouped = T, 
                                    read = T, renorm=T, humann_version = 4, sum=1)

func_profile_filt <- subset_samples_func(func_profile, sample_type == 'ileum') %>% 
 filter_func(abund=1e-8, prev=0.1, renorm = F)  %>% 
  transform_func("relative", sum=1) 
```

### Analysis 

PCoA plot and feature (UNIREF, EC, KO and PY) DA, specific indole metabolism focused analysis 


With all samples 

```{r}
cols <- c("sham"= "#5E5E5E","fMCAO" = "#D00000",mild = "#5E5E5E", "moderate" = "#FF7E79","severe" = "#D00000")
```


```{r}
plot_pcoa(func_profile_filt, what = "KO", var = "condition", colours = cols, size=4)
#ggsave(filename = paste0(outdir, "KO_PCoA.pdf"), width = 9, height = 6, dpi=300, device = "pdf")
```

```{r}
plot_pcoa(func_profile_filt, what = "EC", var = "condition", colours = cols, size = 4, plot = "MDS")
#ggsave(filename = "func_figures/EC_PCoA.pdf", width = 9, height = 6, dpi=300, device = "pdf")
```


```{r}
plot_pcoa(transform_func(func_profile_filt, "arcsin"), what = "KO", var = "severity", colours = cols, size=4, add_ellipse=F) + 
  geom_polygon(stat = "ellipse", aes(fill = severity), alpha = 0.1) 
  ggsave(filename = "func_figures_2025/KO_PCoA_sev.pdf", width = 5.25, height = 4, dpi=300, device = "pdf")
```
```{r}
plot_pcoa(transform_func(func_profile_filt, "arcsin"), what = "EC", var = "severity", colours = cols, size=4, add_ellipse=F) + 
  geom_polygon(stat = "ellipse", aes(fill = severity), alpha = 0.1) 
  ggsave(filename = "func_figures_2025/EC_PCoA_sev.pdf", width = 5.25, height = 4, dpi=300, device = "pdf")
```

```{r}
plot_pcoa(func_profile_filt, what = "pathways", var = "severity", colours = cols)
#ggsave(filename = "func_figures/PY_PCoA_sev.pdf", width = 9, height = 6, dpi=300, device = "pdf")
```

```{r}
func_profile_filt_st <- subset_samples_func(func_profile, sample_type == 'stool') %>% 
 filter_func(abund=1e-8, prev=0.1, renorm = F)  %>% 
  transform_func("relative", sum=1) 
```

```{r}
plot_pcoa(transform_func(func_profile_filt_st, "arcsin"), what = "KO", var = "severity", colours = cols, size=4, add_ellipse=F) + 
  geom_polygon(stat = "ellipse", aes(fill = severity), alpha = 0.1) 
  ggsave(filename = "func_figures_2025/KO_PCoA_sev_stool.pdf", width = 5.25, height = 4, dpi=300, device = "pdf")
```

```{r}
plot_pcoa(transform_func(func_profile_filt_st, "arcsin"), what = "EC", var = "severity", colours = cols, size=4, add_ellipse=F) + 
  geom_polygon(stat = "ellipse", aes(fill = severity), alpha = 0.1) 
  ggsave(filename = "func_figures_2025/EC_PCoA_sev_stool.pdf", width = 5.25, height = 4, dpi=300, device = "pdf")
```

## Pairwise-feature testing 

### Feature fold change

KO log2 FC - sham vs moderate
```{r}
sham_vs_moderate_KO <- func_profile_filt %>% 
  subset_samples_func(severity != 'severe') %>% 
  feat_fold_change(feat_type = "KO",var = "severity", case = "moderate", ctrl = "mild")
```

KO log2 FC - sham vs severe
```{r}
sham_vs_severe_KO <- func_profile_filt %>% 
  subset_samples_func(severity != 'moderate') %>% 
  feat_fold_change(feat_type = "KO",var = "severity", case = "severe", ctrl = "mild")
```

```{r}
moderate_vs_severe_KO <- func_profile_filt %>% 
  subset_samples_func(severity != 'mild') %>% 
  feat_fold_change(feat_type = "KO",var = "severity", case = "severe", ctrl = "moderate")
```

PY analysis 

```{r}
sham_vs_moderate_PY <- func_profile_filt %>% 
  subset_samples_func(severity != 'severe') %>% 
  feat_fold_change(feat_type = "pathways",var = "severity", case = "moderate", ctrl = "mild")
```

```{r}
sham_vs_severe_PY <- func_profile_filt %>% 
  subset_samples_func(severity != 'moderate') %>% 
  feat_fold_change(feat_type = "pathways",var = "severity", case = "severe", ctrl = "mild")
```

```{r}
moderate_vs_severe_PY <- func_profile_filt %>% 
  subset_samples_func(severity != 'mild') %>% 
  feat_fold_change(feat_type = "pathways",var = "severity", case = "severe", ctrl = "moderate")
```

#### Better DA plots 

```{r}
pval_colors <- c("#889296FF", "#A7ADAEFF", "#949C9FFF", "#F59885FF", "#ED5F55FF", "#DB3842FF", "#D13140FF")
pvals <- c(0.00, 0.30, 0.60, 1, 1.3,1.60, 1.9, 2.25)
```

```{r}
fc_strat <- bind_rows(
  sham_vs_moderate_KO$res %>%
    rownames_to_column("feature") %>%
    mutate(comparison = "Mild vs Moderate"),
  sham_vs_severe_KO$res %>%
    rownames_to_column("feature") %>%
    mutate(comparison = "Mild vs Severe"),
  moderate_vs_severe_KO$res %>%
    rownames_to_column("feature") %>%
    mutate(comparison = "Moderate vs Severe")
)

fc_strat_py <- bind_rows(
  sham_vs_moderate_PY$res %>%
    rownames_to_column("feature") %>%
    mutate(comparison = "Mild vs Moderate"),
  sham_vs_severe_PY$res %>%
    rownames_to_column("feature") %>%
    mutate(comparison = "Mild vs Severe"),
  moderate_vs_severe_PY$res %>%
    rownames_to_column("feature") %>%
    mutate(comparison = "Moderate vs Severe")
)

```

Sorting by effect size 
```{r}
colpal2 <- c(cols[3:5], "lightgrey")
names(colpal2) <- c("mild", "moderate", "severe", "ns")
fc_strat %>%
  mutate(enrichment = case_when(log2FC < 1 ~ "depleted",
  log2FC > 1 ~ "enriched")) %>% 
  # filter to top 10+
  group_by(comparison, enrichment) %>%
    mutate(`-log10(qval)` = -log10(p.adj)) %>% 
  slice_max(abs(log2FC), n = 10) %>%
  ungroup() %>% 
  mutate(feature = gsub(",", ";", feature)) %>% 
  mutate(KO = gsub(";.*$", "",feature)) %>%
  mutate(
    signif = case_when(
      p.adj < 0.1 & log2FC < 0 & comparison == "Mild vs Moderate" ~ "mild",
      p.adj < 0.1 & log2FC < 0 & comparison == "Mild vs Severe" ~ "mild",
      p.adj < 0.1 & log2FC > 0 & comparison == "Mild vs Moderate" ~ "moderate",
      p.adj < 0.1 & log2FC > 0 & comparison == "Mild vs Severe" ~ "severe",
      TRUE ~ "ns"
    )
  ) %>%
  ggplot(aes(x = log2FC, y = reorder_within(KO, log2FC, comparison), fill = `-log10(qval)`)) +
  geom_col() +
  scale_y_reordered() +
  theme_cowplot() +
 scale_fill_gradientn(
  colours = pval_colors,
  values = scales::rescale(pvals),
  limits = c(0, 2.25)
) + 
  #paletteer::scale_fill_paletteer_c("ggthemes::Red-Black Diverging", direction=-1) + 
  theme(panel.border = element_blank(), axis.line = element_line()) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18, face = "italic"),
    axis.title = element_text(size = 20),
    strip.background = element_rect(fill="white", color="black"),
    legend.text = element_text(size = 18)
  ) +
  xlab("Log2FC") +
  ylab("") +
  facet_wrap(vars(comparison), scales = "free")
ggsave("func_figures_2025/fc_KO-effsize.pdf", device = "pdf", dpi = 300, height = 5.5, width = 18)
```

selecting significant only 

```{r}
fc_strat %>%
  mutate(enrichment = case_when(log2FC < 1 & p.adj <= 0.1 ~ "depleted",
  log2FC > 1 & p.adj <=0.1 ~ "enriched")) %>% 
  filter(p.adj <= 0.1) %>% 
  # filter to top 10+
  group_by(comparison, enrichment) %>%
    mutate(`-log10(qval)` = -log10(p.adj)) %>% 
  slice_max(abs(log2FC), n = 10) %>%
  ungroup() %>% 
  mutate(feature = gsub(",", ";", feature)) %>% 
  mutate(KO = gsub(";.*$", "",feature)) %>%
  mutate(
    signif = case_when(
      p.adj < 0.1 & log2FC < 0 & comparison == "Mild vs Moderate" ~ "mild",
      p.adj < 0.1 & log2FC < 0 & comparison == "Mild vs Severe" ~ "mild",
      p.adj < 0.1 & log2FC > 0 & comparison == "Mild vs Moderate" ~ "moderate",
      p.adj < 0.1 & log2FC > 0 & comparison == "Mild vs Severe" ~ "severe",
      TRUE ~ "ns"
    )
  ) %>%
  ggplot(aes(x = log2FC, y = reorder_within(KO, log2FC, comparison), fill = `-log10(qval)`)) +
  geom_col() +
  scale_y_reordered() +
  theme_cowplot() +
 scale_fill_gradientn(
  colours = pval_colors,
  values = scales::rescale(pvals),
  limits = c(0, 2.25)
) + 
  #paletteer::scale_fill_paletteer_c("ggthemes::Red-Black Diverging", direction=-1) + 
  theme(panel.border = element_blank(), axis.line = element_line()) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18, face = "italic"),
    axis.title = element_text(size = 20),
    strip.background = element_rect(fill="white", color="black"),
    legend.text = element_text(size = 18)
  ) +
  xlab("Log2FC") +
  ylab("") +
  facet_wrap(vars(comparison), scales = "free")
ggsave("func_figures_2025/fc_KO-pval.pdf", device = "pdf", dpi = 300, height = 5.5, width = 18)
```

Pathways  

Fix names 
```{r}
fc_strat_py <- fc_strat_py %>% 
  mutate(PY = str_extract(feature, "(?<=: ).*(?=_NA)"))
```

```{r}
fc_strat_py %>%
  mutate(enrichment = case_when(log2FC < 1 & p.adj <= 0.1 ~ "depleted",
  log2FC > 1 & p.adj <=0.1 ~ "enriched")) %>% 
  filter(p.adj <= 0.1) %>% 
  # filter to top 10+
  group_by(comparison, enrichment) %>%
    mutate(`-log10(qval)` = -log10(p.adj)) %>% 
  slice_max(abs(log2FC), n = 10) %>%
  ungroup() %>% 
  mutate(PY = str_wrap(PY, width = 30)) %>% 
  mutate(
    signif = case_when(
      p.adj < 0.1 & log2FC < 0 & comparison == "Mild vs Moderate" ~ "mild",
      p.adj < 0.1 & log2FC < 0 & comparison == "Mild vs Severe" ~ "mild",
      p.adj < 0.1 & log2FC > 0 & comparison == "Mild vs Moderate" ~ "moderate",
      p.adj < 0.1 & log2FC > 0 & comparison == "Mild vs Severe" ~ "severe",
      TRUE ~ "ns"
    )
  ) %>%
  ggplot(aes(x = log2FC, y = reorder_within(PY, log2FC, comparison), fill = `-log10(qval)`)) +
  geom_col() +
  scale_y_reordered() +
  theme_cowplot() +
 scale_fill_gradientn(
  colours = pval_colors,
  values = scales::rescale(pvals),
  limits = c(0, 2.25)
)  +
  #paletteer::scale_fill_paletteer_c("ggthemes::Red-Black Diverging", direction=-1) + 
  theme(panel.border = element_blank(), axis.line = element_line()) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18, face = "italic"),
    axis.title = element_text(size = 20),
    strip.background = element_rect(fill="white", color="black"),
    legend.text = element_text(size = 18)
  ) +
  xlab("Log2FC") +
  ylab("") +
  facet_wrap(vars(comparison), scales = "free") 
ggsave("func_figures_2025/fc_PY-pval.pdf", device = "pdf", dpi = 300, height = 8.25, width = 18)
```

Different plot style

```{r}

```

```{r}
py_mild_moderate <- fc_strat_py %>% 
  filter(p.adj <= 0.1, comparison == "Mild vs Moderate") %>% 
  mutate(
    enrichment = if_else(log2FC < 0, "depleted", "enriched"),
    PY = str_wrap(PY, width = 30)
  ) %>% 
  group_by(enrichment) %>%
  slice_max(abs(log2FC), n = 10) %>%
  ungroup() %>% 
  # Calculate rectangle positions
  mutate(
    is_depleted = log2FC < 0,
    depleted_count = sum(is_depleted),
    total_count = n()
  ) %>%
  ggplot(aes(x = 1, y = fct_reorder(PY, log2FC))) + 
  geom_point(aes(fill = abs(log2FC)), alpha = 0.75, shape = 21, size = 10, color = "white") +
  # Dynamic rectangle positioning
  geom_rect(
  data = . %>% slice(1),
  aes(xmin = 0.75, xmax = 1.25, ymin = 0.5, ymax = depleted_count + 0.5),
  alpha = 0.2, fill = cols[3], inherit.aes = FALSE
) +
geom_rect(
  data = . %>% slice(1),
  aes(xmin = 0.75, xmax = 1.25, ymin = depleted_count + 0.5, ymax = total_count + 0.5),
  alpha = 0.2, fill = cols[4], inherit.aes = FALSE
) +
  scale_fill_gradient2(
    low = "#FCEDED", mid = "#E14141", high = "#780101", 
    midpoint = 12.5, limits = c(0, 25),
    name = "Absolute log2FC",
    guide = guide_colorbar(
      frame.colour = "black", 
      ticks.colour = "white", 
      frame.linewidth = 0.75, 
      ticks.linewidth = 0.75
    )
  ) + 
  cowplot::theme_cowplot() + 
  theme(
    axis.line.x = element_blank(), 
    axis.ticks.x = element_blank(),
    axis.title = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(colour = "black", size = 16),
    legend.position = "None",
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_cartesian(clip = "off")

py_mild_moderate
```

```{r}
py_mild_severe <- fc_strat_py %>% 
  filter(p.adj <= 0.1, comparison == "Mild vs Severe") %>% 
  mutate(
    enrichment = if_else(log2FC < 0, "depleted", "enriched"),
    PY = str_wrap(PY, width = 30)
  ) %>% 
  group_by(enrichment) %>%
  slice_max(abs(log2FC), n = 10) %>%
  ungroup() %>% 
  # Calculate rectangle positions
  mutate(
    is_depleted = log2FC < 0,
    depleted_count = sum(is_depleted),
    total_count = n()
  ) %>%
  ggplot(aes(x = 1, y = fct_reorder(PY, log2FC))) + 
  geom_point(aes(fill = abs(log2FC)), alpha = 0.75, shape = 21, size = 10, color = "white") +
  # Dynamic rectangle positioning
  geom_rect(
  data = . %>% slice(1),
  aes(xmin = 0.75, xmax = 1.25, ymin = 0.5, ymax = depleted_count + 0.5),
  alpha = 0.2, fill = cols[3], inherit.aes = FALSE
) +
geom_rect(
  data = . %>% slice(1),
  aes(xmin = 0.75, xmax = 1.25, ymin = depleted_count + 0.5, ymax = total_count + 0.5),
  alpha = 0.2, fill = cols[5], inherit.aes = FALSE
) +
  scale_fill_gradient2(
    low = "#FCEDED", mid = "#E14141", high = "#780101", 
    midpoint = 12.5, limits = c(0, 25),
    name = "Absolute log2FC",
    guide = guide_colorbar(
      frame.colour = "black", 
      ticks.colour = "white", 
      frame.linewidth = 0.75, 
      ticks.linewidth = 0.75
    )
  ) + 
  cowplot::theme_cowplot() + 
  theme(
    axis.line.x = element_blank(), 
    axis.ticks.x = element_blank(),
    axis.title = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(colour = "black", size = 16),
    legend.position = "None",
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_cartesian(clip = "off")
py_mild_severe
```


```{r}
py_moderate_severe <- fc_strat_py %>% 
  filter(p.adj <= 0.1, comparison == "Moderate vs Severe") %>% 
  mutate(
    enrichment = if_else(log2FC < 0, "depleted", "enriched"),
    PY = str_wrap(PY, width = 30)
  ) %>% 
  group_by(enrichment) %>%
  slice_max(abs(log2FC), n = 10) %>%
  ungroup() %>% 
  # Calculate rectangle positions
  mutate(
    is_depleted = log2FC < 0,
    depleted_count = sum(is_depleted),
    total_count = n()
  ) %>%
  ggplot(aes(x = 1, y = fct_reorder(PY, log2FC))) + 
  geom_point(aes(fill = abs(log2FC)), alpha = 0.75, shape = 21, size = 10, color = "white") +
  # Dynamic rectangle positioning
  geom_rect(
  data = . %>% slice(1),
  aes(xmin = 0.75, xmax = 1.25, ymin = 0.5, ymax = depleted_count + 0.5),
  alpha = 0.2, fill = cols[4], inherit.aes = FALSE
) +
geom_rect(
  data = . %>% slice(1),
  aes(xmin = 0.75, xmax = 1.25, ymin = depleted_count + 0.5, ymax = total_count + 0.5),
  alpha = 0.2, fill = cols[5], inherit.aes = FALSE
) +
  scale_fill_gradient2(
    low = "#FCEDED", mid = "#E14141", high = "#780101", 
    midpoint = 12.5, limits = c(0, 25),
    name = "Absolute log2FC",
    guide = guide_colorbar(
      frame.colour = "black", 
      ticks.colour = "white", 
      frame.linewidth = 0.75, 
      ticks.linewidth = 0.75
    )
  ) + 
  cowplot::theme_cowplot() + 
  theme(
    axis.line.x = element_blank(), 
    axis.ticks.x = element_blank(),
    axis.title = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(colour = "black", size = 16),
    legend.text = element_text(size = 14, colour = "black"), 
    legend.title = element_text(size = 16),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_cartesian(clip = "off")
py_moderate_severe
```

Final PY plot 

```{r}
final_py_plot <- cowplot::plot_grid(
  py_mild_moderate + ggtitle("Mild vs Moderate"),
  py_mild_severe + ggtitle("Mild vs Severe"),
  py_moderate_severe + ggtitle("Moderate vs Severe"),
  ncol = 3,
  align = "hv"
)
final_py_plot
ggsave("func_figures_2025/fc_PY-absFC.pdf", plot = final_py_plot, device = "pdf", dpi = 300, height = 6.5, width = 20)
```

### GBM analysis 

```{r}
gutbrain_modules <- read_tsv(gbm_path) %>% 
  unite("GBM", Module:Description, sep=":") %>% 
  column_to_rownames("GBM")
# remove stool 
gutbrain_modules <- gutbrain_modules %>% select(contains("ileum"))
```

Here we want to test if any GBMs are associated with severity score 

```{r}
gutbrain_modules<- gutbrain_modules + 1e-10
gutbrain_modules_rel <- t(100 * t(gutbrain_modules) / colSums(gutbrain_modules))
gutbrain_modules_clr <- t(mixOmics::logratio.transfo(t(gutbrain_modules_rel), logratio = "CLR"))
class(gutbrain_modules_clr) <- "matrix"
```

```{r}
gutbrain_modules_df <- gutbrain_modules_clr %>%
  as.data.frame() %>%
  rownames_to_column("GBM") %>%
  pivot_longer(-GBM)
gutbrain_modules_df$name <- gsub("\\.", "-", gutbrain_modules_df$name)
gutbrain_modules_df$name <- gsub("_merged_clean", "", gutbrain_modules_df$name)
```

Merge with metadata 

```{r}
# Join the dataframes
gutbrain_modules_df_long <- gutbrain_modules_df %>%
  left_join(func_profile_filt$Metadata %>% rownames_to_column("name"), by = "name")
```
Remove sham samples 

```{r}
gutbrain_modules_df_long <- gutbrain_modules_df_long %>% 
  filter(condition !="sham") %>% 
  filter(name != "RD521-ileum")
```

glm value ~ severity_score 
```{r}
gbm_res <-
  data.frame(
    feature = character(0),
    coef = numeric(0),
    std.err = numeric(0),
    pval= numeric(0),
    name=character(0),
    qval=numeric(0),
    N=numeric(0)
  )

for (i in unique(gutbrain_modules_df_long$GBM)) {
  fit <- glm(
    severity_score ~ value,
    data = filter(gutbrain_modules_df_long, GBM == i),
    family = gaussian(link = "identity")
  )

  # Extract coefficients summary
  coef_summary <- summary(fit)$coefficients

  # Get the data used in the model (handles NA removal automatically)
  model_data <- fit$model

  # Create the output dataframe
  gbm_res <- bind_rows(gbm_res, data.frame(
    feature = i, # or the OTU name
    coef = coef_summary[-1, "Estimate"],
    stderr = coef_summary[-1, "Std. Error"],
    pval = coef_summary[-1, "Pr(>|t|)"], # or "Pr(>|t|)" depending on family
    name = paste(i, rownames(coef_summary)[-1], sep = "_"),
    qval = NA, # calculate later with p.adjust()
    N = nrow(model_data)
  ))

 
}

 # Add q-values (FDR correction)
gbm_res$qval <- p.adjust(gbm_res$pval, method = "BH")
```

GBM plot 

```{r}
sig_gbm <- gbm_res %>% 
  filter(qval < 0.25) %>% 
  mutate(enrichment = case_when(coef < 0 ~ "depleted",
  coef > 0 ~ "enriched")) %>% 
  group_by(enrichment) %>% 
  slice_max(abs(coef), n=20) %>% 
  ungroup()
```

```{r}
sig_gbm %>% 
  mutate(group = "group") %>% 
  mutate(`-log10(qval)` = -log10(qval)) %>% 
ggplot(aes(x= fct_reorder(feature, coef), y=group, size = abs(coef), color=`-log10(qval)`)) +
    geom_point(alpha=0.5) +
 scale_color_gradientn(
  colours = pval_colors,
  values = scales::rescale(pvals),
  limits = c(0, 2.25)
) +
  theme_cowplot() + 
  theme(axis.title.x= element_blank(),
axis.title.y = element_text("          "), 
  axis.text.y = element_blank())  +
    scale_size(range = c(.5, 10)) +
    rotate_x_text(45)
ggsave("func_figures_2025/gbms_effsize_horizontal.pdf", device="pdf", dpi=300, height=6.5, width=15)
```

### Specific indole genes

Indole genes table 

```{r}
indole_genes <- read_tsv("../data/indole_metabolism_ko_ec.txt")
indole_ko <- indole_genes %>% 
  drop_na(KO) %>% 
  pull(KO) %>% 
  unique()
indole_ko <- paste0(indole_ko, "_")
indole_ec <- indole_genes %>% 
  drop_na(EC) %>% 
  pull(EC)
indole_ec <- stringr::str_conv(indole_ec, "UTF-8")
indole_ec <- c(indole_ec, "4.1.1.115")
indole_ec <- paste0("EC:", indole_ec, "_")
```

```{r}
ec_filt <- func_profile$EC %>% 
  rownames_to_column("EC") %>% 
  filter(grepl(paste(indole_ec, collapse="|"), EC)) %>% 
  column_to_rownames("EC") %>% 
  t_df()
```


```{r}
for (i in colnames(ec_filt)){
  sub <- ec_filt %>% 
    select(i) %>% 
    merge(func_profile_filt$Metadata, by=0) %>% 
    rename(ec = i) %>% 
    mutate(ec = asin(sqrt(ec)))
   sub[, "condition"] <- factor(sub[, "condition"], levels = c("sham", "fMCAO"))

  p <- ggplot(sub, aes(x = condition, y = ec, color=condition)) + 
    geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75) + 
    scale_color_manual(values = cols) + 
    theme_cowplot() + 
    ylab(paste0(i, "\n", "Asin(sqrt(rel. abundance))")) +
    ggtitle(i) + 
    theme(axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") +
  stat_compare_means(method = "wilcox.test", label = "p.format", comparisons = list(c("sham", "fMCAO"))) + 
  EnvStats::stat_n_text()
  name <- sub("_.*", "", i)
  name <- gsub(":", "_", i)
  ggsave(paste0(outdir, "/indole_genes/", name, "-2025.pdf"), plot = p, device = "pdf",dpi=300, height = 4, width = 4.5)
}
```


```{r}
comps <- microbiomeutilities::make_pairs(func_profile$Metadata$severity)
for (i in colnames(ec_filt)){
  sub <- ec_filt %>% select(i) %>% 
    merge(func_profile_filt$Metadata, by=0) %>% 
    rename(ec = i) %>% 
    mutate(ec = asin(sqrt(ec))) 

  sub[, "severity"] <- factor(sub[, "severity"], levels = c("mild", "moderate", "severe"))

  p <- ggplot(sub, aes(x = severity, y = ec, color=severity, label=Row.names)) + 
    geom_jitter(position = position_jitterdodge(jitter.width = 0.175), size=3.75, alpha=0.5) + 
   #ggrepel::geom_text_repel() + 
    theme_cowplot() + 
    scale_color_manual(values = cols)+ 
    ylab(paste0(i, "\n", "Asin(sqrt(rel. abundance))")) +
    #ylab("rel. abundance") + 
    ggtitle(i) + 
    theme(axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") +
  stat_compare_means(method = "wilcox.test", label = "p.format", comparisons = comps) + 
  EnvStats::stat_n_text()
  name <- sub("_.*", "", i)
  name <- gsub(":", "_", i)
  ggsave(paste0(outdir, "/indole_genes/", name, "_severity.pdf"), plot = p, device = "pdf",dpi=300, height = 4, width = 5)
  print(p)
}

```
Pairwise comparisons 
```{r}
pairwise_comps <- list(c("mild", "moderate"), c("mild", "severe"), c("moderate", "severe"))
for (i in colnames(ec_filt)){
  for (j in 1:length(pairwise_comps)){
    sub <- ec_filt %>% select(i) %>% 
    merge(func_profile_filt$Metadata, by=0) %>% 
    rename(ec = i) %>% 
    filter(severity %in% pairwise_comps[[j]]) %>% 
    mutate(ec = asin(sqrt(ec)))
    
  comps <- microbiomeutilities::make_pairs(sub$severity)
  sub[, "severity"] <- factor(sub[, "severity"], levels = pairwise_comps[[j]])
  
  name <- gsub("_", " ", i)
  
  p <- ggplot(sub, aes(x = severity, y = ec, color=severity, label=Row.names)) + 
    geom_jitter(position = position_jitterdodge(jitter.width = 0.1), size=3.75) + 
    theme_cowplot() + 
    scale_color_manual(values = cols)+ 
    ylab(paste0(name, "\n", "Asin(sqrt(rel. abundance))")) +
    #ylab("rel. abundance") + 
    theme(axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") +
  stat_compare_means(method = "wilcox.test", label = "p.format", comparisons = comps)
  name <- sub("_.*", "", i)
  name <- gsub(":", "_", i)
  comparison <- paste(pairwise_comps[[j]], collapse = "_")
  ggsave(paste0("func_figures_2025/indole_genes/", name, "_", comparison,  "_severity.pdf"), plot = p, device = "pdf",dpi=300, height = 4, width = 4.5)
   print(p) 
  }
 
}

```

### Correlation of tnaA with severity

```{r}
ec_filt %>% select(`EC:4.1.99.1_Tryptophanase`) %>% 
    merge(func_profile_filt$Metadata, by=0) %>% 
  filter(condition != "sham") %>% 
  plot_scatter(x="severity_score", y="EC:4.1.99.1_Tryptophanase", line_color = "firebrick1", fill_color = "firebrick1", point_color = "firebrick1", xlab="Severity score", ylab="Asin(sqrt(rel. abundance))",corr.method = "spearman")
ggsave("func_figures_2025/indole_genes/tnaA_severity_cor.pdf", device = "pdf", dpi=300, height = 4, width = 5)
```


### Stratified 

```{r}
ec_strat <- read_tsv("../data/humann4/search_coverage_30/merged_table/humann_genefamilies_ec.tsv") %>% 
  rename(contrib = `# Gene Family HUMAnN v4.0.0.alpha.1 Adjusted CPMs`) %>% 
  pivot_longer(-contrib, names_to = "sampleID", values_to = "Abundance")
ec_strat$sampleID <- gsub("_merged_clean", "", ec_strat$sampleID)
```

```{r}
plot_contrib <- function(strat_profile, pattern, ylabel, mean=F) {
  pattern <- paste0(pattern, "\\|")

  strat_profile <- strat_profile %>% 
    filter(grepl(pattern, contrib))
  
  if (mean == TRUE){
    strat_profile <- strat_profile %>% 
      group_by(severity, contrib) %>% 
      summarise(Abundance = mean(Abundance))  %>%
    ggplot(aes(fill = contrib, y = Abundance, x = severity)) +
    geom_bar(position = "stack", stat = "identity") +
    facet_grid(. ~ ~ factor(severity, levels = c("mild", "moderate", "severe")),
               scales = "free_x",
               space = "free_x") +
    theme_classic() +
    theme(
      panel.spacing = unit(0, "pt"),
      panel.border = element_rect(colour = "grey50", fill = NA)
    ) +
    scale_y_continuous(expand = c(0, 0)) +
    paletteer::scale_fill_paletteer_d("ggsci::default_igv") + 
    theme(
      axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text = element_text(size = 14),
      axis.title = element_text(size = 16),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 14),
      strip.text = element_text(size = 14)
    ) +
    ylab(ylabel)
  } else {

    strat_profile %>%
    ggplot(aes(fill = contrib, y = Abundance, x = sampleID)) +
    geom_bar(position = "stack", stat = "identity") +
    facet_grid(. ~ ~ factor(severity, levels = c("mild", "moderate", "severe")),
               scales = "free_x",
               space = "free_x") +
    theme_classic() +
    theme(
      panel.spacing = unit(0, "pt"),
      panel.border = element_rect(colour = "grey50", fill = NA)
    ) +
    scale_y_continuous(expand = c(0, 0)) +
    paletteer::scale_fill_paletteer_d("ggsci::default_igv") + 
    theme(
      axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text = element_text(size = 14),
      axis.title = element_text(size = 16),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 14),
      strip.text = element_text(size = 14)
    ) +
    ylab(ylabel)
  }
  
}
```



```{r}
mild_samples <- func_profile_filt$Metadata %>% 
  filter(severity == "mild") %>% 
  rownames()
moderate_samples <- func_profile_filt$Metadata %>% 
  filter(severity == "moderate") %>% 
  rownames()
```

```{r}
ec_strat_meta <- ec_strat %>%
  mutate(
    severity = case_when(
      sampleID %in% mild_samples ~ "mild",
      sampleID %in% moderate_samples ~ "moderate",
      TRUE ~ "severe"
    )
  ) 
```

#### Indole genes
 
##### KO

plot - tnaA

```{r}
ec <- indole_genes$EC
names(ec) <- indole_genes$Gene
#ec <- c(ko, "low affinity tryptophan permease" = "K03836")
```

```{r}
ec_detected <- ec[ec %in% ec_strat_meta$contrib]
```


Use code from above but adjust so values are converted to % and group low abundant into other 
```{r, fig.height=7, fig.width=12}
#imap(ko_detected,function(.x, .y) plot_contrib(ko_strat_meta, pattern = .x, ylabel = paste0(.y, "[CPM]")))
plot_contrib(ec_strat_meta, pattern = "K03836", ylabel = "Tryptophanase (tnaA) [CPM]", mean=T) + 
  theme(legend.text = element_text(size=6))
```

```{r}
pattern <- paste0("4.1.99.1", "\\|")

# Set threshold for grouping low abundant species
abundance_threshold <- 5  # Adjust this value as needed (e.g., 1% of composition)

encoding_species <- ec_strat_meta %>%
  filter(grepl(pattern, contrib)) %>%
  group_by(severity, contrib) %>%
  summarise(Abundance = mean(Abundance), .groups = "drop") %>%
  # Calculate composition (convert to percentage)
  group_by(severity) %>%
  mutate(
    total = sum(Abundance),
    composition = (Abundance / total) * 100
  ) %>%
  ungroup() %>%
  # Group low abundant species into "Other"
  mutate(
    contrib = ifelse(composition < 5, "Other", contrib)
  ) %>%
  # Recalculate after grouping
  group_by(severity, contrib) %>%
  summarise(composition = sum(composition), .groups = "drop") %>%
  distinct(contrib) %>%
  pull(contrib)
```

```{r}
strat_pal <- c("#E4E9B2","#7E6148FF","#91D1C2FF","#E64B35FF", "#B09C85FF","#DCDCDC", "#F9A620","#054A29","#52414C" ,"#F39B7FFF", "white")
names(strat_pal) <- encoding_species
```

```{r}

ec_strat_meta %>% 
  filter(grepl(pattern, contrib)) %>%
  group_by(severity, contrib) %>% 
  summarise(Abundance = mean(Abundance), .groups = "drop") %>%
  # Calculate composition (convert to percentage)
  group_by(severity) %>%
  mutate(
    total = sum(Abundance),
    composition = (Abundance / total) * 100
  ) %>%
  ungroup() %>%
  # Group low abundant species into "Other"
  mutate(
    contrib = ifelse(composition < abundance_threshold, "Other", contrib)
  ) %>%
  # Recalculate after grouping
  group_by(severity, contrib) %>%
  summarise(composition = sum(composition), .groups = "drop") %>% 
ggplot(aes(fill = contrib, y = composition, x = severity)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_grid(. ~ factor(severity, levels = c("mild", "moderate", "severe")),
             scales = "free_x",
             space = "free_x") +
  theme_classic() +
  theme(
    panel.spacing = unit(0, "pt"),
    panel.border = element_rect(colour = "grey50", fill = NA)
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values= strat_pal) + 
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 14)
  ) +
  ylab("Tryptophanase (tnaA) [%]")
ggsave("func_figures_2025/tnaA_stratified_composition.pdf", device="pdf", height=5, width=9, dpi=300)
```

```{r}
# Set threshold for grouping low abundant species
abundance_threshold <- 1e-06  # Adjust this value as needed (absolute abundance units)

encoding_species_cpm <- ec_strat_meta %>%
  filter(grepl(pattern, contrib)) %>%
  group_by(severity, contrib) %>%
  summarise(Abundance = mean(Abundance), .groups = "drop") %>%
  # Calculate composition (convert to percentage)
  # Group low abundant species into "Other"
  mutate(
    contrib = ifelse(Abundance < abundance_threshold, "Other", contrib) 
  ) %>%
  # Recalculate after grouping
  group_by(severity, contrib) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  distinct(contrib) %>%
  pull(contrib)

strat_pal_cpm <- c("#F9A620", "#00A087FF","#054A29", "#E4E9B2","#7E6148FF","#91D1C2FF","#52414C", "#E64B35FF","#646464ff","#DCDCDC", "#DC0000FF", "#F39B7FFF", "#4DBBD5FF")
#c(, , )
names(strat_pal_cpm) <- encoding_species_cpm

ec_strat_meta %>% 
  filter(grepl(pattern, contrib)) %>%
  group_by(severity, contrib) %>% 
  summarise(Abundance = mean(Abundance), .groups = "drop") %>%
  # Group low abundant species into "Other"
  mutate(
    contrib = ifelse(Abundance < abundance_threshold, "Other", contrib)
  ) %>%
  # Recalculate after grouping
  group_by(severity, contrib) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>% 
ggplot(aes(fill = contrib, y = Abundance, x = severity)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_grid(. ~ factor(severity, levels = c("mild", "moderate", "severe")),
             scales = "free_x",
             space = "free_x") +
  theme_classic() +
  theme(
    panel.spacing = unit(0, "pt"),
    panel.border = element_rect(colour = "grey50", fill = NA)
  ) +
  scale_y_continuous(expand = c(0, 0)) +
    scale_fill_manual(values= strat_pal_cpm) + 
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14),
    strip.text = element_text(size = 14)
  ) +
  ylab("Tryptophanase (tnaA) [CPM]")
ggsave("func_figures_2025/tnaA_stratified_cpm.pdf", device="pdf", height=5, width=9, dpi=300)
```