---
title: "ahr_ileum_tax"
author: "Adam Sorbie"
date: "2023-06-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of taxonomic profiles from ileum content of sham/stroke mice - AHR project

This document contains code to analyse taxonomic profiles produced from 150bp PE
Illumina Nextseq sequencing of DNA extracted from the ileum content of sham
and stroke (fMCAO) mice. 

Sequencing Methods overview: 

Platform: Illumina Nextseq
Adapters: Nextera PE
Paired: TRUE
Read length: 150bp
Depth: 2Gbp (approximate target)
Avg. reads per sample:14078731.33

Note that 1 sample did not pass QC (M14). 


QC overview: 

Reads were merged (simple cat), trimmed and filtered and then aligned against
the mouse/phiX genomes to filter non-microbial reads 

Trimming: Atria 
Adapters: Nextera PE
Length filtering: 75-200bp 
Quality filtering: Q20
polyG trimming: TRUE
bowtie2: very-sensitive 


## Taxonomic profiling 

Post-QC reads were profiled using MetaPhlAn4 and mOTUs3. High sensitivity presets
were used in both cases, to increase the number of identified taxa. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
library(microViz)
library(ComplexHeatmap)
library(ggplotify)
library(tidytext)
source("../scripts/shotgun_analysis_utils.R")
source("../scripts/taxa_analysis.R")
metaphlan_path <- "../data/metaphlan4.1/merged_table/merged_table.txt"
conflicted::conflict_prefer_all("dplyr")
conflicted::conflicts_prefer(vegan::diversity)
conflicted::conflicts_prefer(purrr::map)
```

```{r}
cols <- c("sham"= "#5E5E5E","fMCAO" = "#D00000",mild = "#5E5E5E", "moderate" = "#FF7E79","severe" = "#D00000")
```
```{r}
metaphlan_species <- import_pseq_metag(merged_table_path = metaphlan_path,
                                           metapath = "../data/ahr_metadata-2025.txt",
                                           level = "species") %>% 
  transform()
```

Add sample IDs as column so we can annotate NMDS with names
```{r}
sample_ids <- rownames(meta_to_df(metaphlan_species))
metaphlan_species <- metaphlan_species %>% 
  ps_mutate(sampleID = sample_ids)
```

### Combined ileum and stool analysis 

```{r}
metaphlan_species <- metaphlan_species %>% 
  subset_samples(sampleID = sample_ids)
```

Relatively simple alpha-diversity metrics and beta-diversity based on bray-curtis 
and/or jaccard.

#### Analysis - alpha diversity 

```{r}

alpha_metaphlan <- calc_alpha(metaphlan_species) %>% 
  merge(meta_to_df(metaphlan_species), by=0)
comps <- list(c("sham", "fMCAO"))
sev_comps <- list(c("mild", "moderate"), c("mild", "severe"), c("moderate", "severe"))

capitalize <- function(string) {
  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
  string
  }
```

```{r}
alpha_metaphlan %>%
  ggplot(aes(x = fct_relevel(condition, "sham", "fMCAO"), y = Richness, color = condition)) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.1),
    size = 4,
    alpha = 0.4
  ) +
  theme_cowplot() +
  scale_color_manual(values = cols)+
  ylab("Species Richness") +
  xlab("") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "None"
  ) +
  stat_summary(
    fun.y = median,
    fun.ymin = median,
    fun.ymax = median,
    geom = "crossbar",
    width = 0.15,
    color = "black"
  ) +
  facet_grid(~sample_type, labeller = labeller(sample_type = capitalize)) + 
  stat_compare_means(
    comparisons = comps,
    method = "wilcox.test",
    label = "p.format"
  ) + 
  theme(strip.background =element_rect(fill="white", color="black")) + 
  rotate_x_text(45)
ggsave("tax_figures/richness_sham_vs_fMCAO.pdf", device=cairo_pdf,dpi=300, height=4.5, width=5)
```

```{r, fig.height=4, fig.width=4.5}
alpha_metaphlan %>%
  ggplot(aes(x = fct_relevel(condition, "sham", "fMCAO"), y = Richness)) +
  geom_jitter(
    aes(shape = severity, color = severity), # Move aesthetics here
    position = position_jitter(width = 0.1, height=0.1), # Use regular jitter, not jitterdodge
    size = 4,
    alpha = 0.5
  ) +
  theme_cowplot() +
  scale_color_manual(values = cols) +
  ylab("Species Richness") +
  xlab("") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "bottom",
    legend.justification = "centre"
  ) +
  stat_summary(
    fun = median, # Updated syntax (fun.y is deprecated)
    fun.min = median, # Updated syntax
    fun.max = median, # Updated syntax
    geom = "crossbar",
    width = 0.15,
    color = "black"
  ) +
  facet_grid(~sample_type, labeller = labeller(sample_type = capitalize)) +
  stat_compare_means(
    comparisons = comps,
    method = "wilcox.test",
    label = "p.format"
  ) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  rotate_x_text(45)

ggsave("tax_figures/richness_sham_vs_fMCAO_w_sev_colour.pdf", device=cairo_pdf,dpi=300, height=4.75, width=5)
```
```{r}
alpha_metaphlan %>%
  ggplot(aes(x = fct_relevel(severity, "mild", "moderate", "severe"), y = Richness, color = severity, shape=sample_type)) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.1),
    size = 4,
    alpha = 0.4
  ) +
  theme_cowplot() +
  scale_color_manual(values = cols)+
  ylab("Species Richness") +
  xlab("") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "None"
  ) +
  stat_summary(
    fun.y = median,
    fun.ymin = median,
    fun.ymax = median,
    geom = "crossbar",
    width = 0.15,
    color = "black"
  ) +
  facet_grid(~sample_type, labeller = labeller(sample_type = capitalize)) + 
  stat_compare_means(
    comparisons = sev_comps,
    method = "wilcox.test",
    label = "p.format"
  ) + 
  theme(strip.background =element_rect(fill="white", color="black")) + 
  rotate_x_text(45)
ggsave("tax_figures/richness_severity.pdf", device=cairo_pdf,dpi=300, height=4.5, width=5.25)
```


Beta-div and PCoA
```{r}
beta_metaphlan_il_stool <-
  calc_betadiv(
    transform(metaphlan_species, "arcsin"),
    dist = "bray",
    ord_method = "NMDS"
  )
```

PERMANOVA - sample_type
severity ~ sample_type


```{r}
severity_adonis <- phyloseq_adonis(metaphlan_species, beta_metaphlan_il_stool$Distance_Matrix,group_variable = "severity")
sample_type_adonis <- phyloseq_adonis(metaphlan_species, beta_metaphlan_il_stool$Distance_Matrix,group_variable = "sample_type")
interaction <- phyloseq_adonis(metaphlan_species, beta_metaphlan_il_stool$Distance_Matrix,group_variable = "severity*sample_type")
```

```{r}
plot_beta_div(metaphlan_species, beta_metaphlan_il_stool$Distance_Matrix, shape = "sample_type",beta_metaphlan_il_stool$Ordination, group_variable = "condition", cols = rev(cols[c(1,2)])) 

ggsave2("tax_figures/betadiv_il_stool_NMDS.pdf", device = "pdf", dpi = 300, height = 4, width = 5)
```

Stratified by severity 

```{r}
plot_beta_div(
  metaphlan_species,
  beta_metaphlan_il_stool$Distance_Matrix,
  shape = "sample_type",
  beta_metaphlan_il_stool$Ordination,
  group_variable = "severity",
  cols = rev(cols[c(3, 4, 5)])
) +
  labs(
    caption = bquote(atop(
      Severity ~ - ~ Adonis ~ R^2 ~ .(round(severity_adonis$R2[1], 2)) ~ ~ p - value ~ .(severity_adonis$`Pr(>F)`[1]),
      Sample ~ type ~ - ~ Adonis ~ R^2 ~ .(round(sample_type_adonis$R2[1], 2)) ~ ~ p - value ~ .(sample_type_adonis$`Pr(>F)`[1]),
      Interaction ~ - ~ Adonis ~ R^2 ~ .(round(interaction$R2[1], 2)) ~ ~ p - value ~ .(interaction$`Pr(>F)`[1])
    ))
  ) 
ggsave2(
  "tax_figures/betadiv_il_stool_NMDS_sev.pdf",
  device = "pdf",
  dpi = 300,
  height = 4.25,
  width = 5
)
```


### Analysis ileum

#### Analysis - bray-curtis 

```{r}
metaphlan_ileum <- metaphlan_species %>% 
  subset_samples(sample_type == "ileum")

```
```{r}
# metaphlan
beta_metaphlan <-
  calc_betadiv(transform(metaphlan_ileum, "arcsin"),
               dist = "bray",
               ord_method = "NMDS")


beta_metaphlan_exc_severe <-
  calc_betadiv(
    subset_samples(transform(metaphlan_ileum, "arcsin"), severity != "severe"),
    dist = "bray",
    ord_method = "NMDS"
  )
```

```{r}
plot_beta_div(metaphlan_ileum, beta_metaphlan$Distance_Matrix, beta_metaphlan$Ordination, group_variable = "condition", cols = rev(cols[c(1,2)])) 

#p +
 # geom_text(data = p$data, aes(x=NMDS1, y=NMDS2, label=sampleID), inherit.aes = F, nudge_y=0.05)

ggsave2("tax_figures/betadiv_NMDS_ileum.pdf", device = "pdf", dpi = 300, height = 4, width = 5)
```

```{r}
plot_beta_div(metaphlan_ileum, beta_metaphlan$Distance_Matrix, beta_metaphlan$Ordination, group_variable = "severity", cols = cols)
#p + geom_text(data = p$data, aes(x=NMDS1, y=NMDS2, label=sampleID), inherit.aes = F, nudge_y=0.05)
ggsave2("tax_figures/betadiv_severity_NMDS.pdf", device = "pdf", dpi = 300, height = 4, width = 7)
```


##### Coloured by tnaA abundance

```{r}
indole_genes_df <- read_tsv("../data/indole_metabolism_ko_ec.txt")
meta_path <- "../data/ahr_metadata-2025.txt"
ec_path <- "../data/humann_com/merged_table/unstrat/humann_genefamilies_ec_unstratified.tsv"
#un_path <- "../data/humann/merged_table/unstrat/humann_genefamilies_relab_unstratified.tsv"
ko_path <- "../data/humann_com/merged_table/unstrat/humann_genefamilies_ko_unstratified.tsv"
py_path <- "../data/humann_com/merged_table/unstrat/humann_pathabundance_relab_unstratified.tsv"
source("../../../Metagenomics/scripts/func_analysis.R")
func_profile <- import_func_profile(meta =meta_path,
                                    ec = ec_path,
                                    ko = ko_path,
                                    py = py_path, 
                                    filter_ungrouped = T, read = T, renorm=T) 
func_profile_ileum <- subset_samples_func(func_profile, "sample_type == 'ileum'") %>% 
   transform_func(transform = "relative") #%>% 
  #  transform_func(transform = "arcsin")
``` 
Filter for genes of interest and add to metadata 

```{r}
indole_genes <- paste0("EC:", indole_genes_df$EC, "_")
ec_goi <- func_profile_ileum$EC %>% 
  select(contains(indole_genes))
ko_goi <- func_profile_ileum$KO %>% 
  select(contains(indole_genes_df %>% drop_na(KO) %>% .$KO))
```

```{r}
meta_ec <- meta_to_df(metaphlan_ileum) %>% 
  merge(ec_goi, by=0) %>% 
  column_to_rownames("Row.names")
sample_data(metaphlan_ileum) <- meta_ec
```

NMDS plots colored by abundance of indole metabolism associated enzymes - EC

```{r}
ggsave3 <- function(...) {
  ggplot2::ggsave(...)
  invisible()
}
```

This currently doesn't work well because of the large gap between the 3 high samples (1 mild and 2 severe) and the other samples where it was detected - how to fix this? 

```{r}
enzymes <- colnames(meta_to_df(metaphlan_ileum) %>% select(contains("EC")))

map(enzymes, function(x) plot_beta_div(metaphlan_ileum, beta_metaphlan$Distance_Matrix, beta_metaphlan$Ordination, group_variable = x) +
  scale_color_viridis_c(alpha = 1, option = "viridis", oob=scales::squish, guide = guide_colorbar(frame.colour = "black", ticks.colour = "white", frame.linewidth = 0.75, ticks.linewidth = 0.75, draw.ulim = F)) + 
    ggtitle(x) + 
    ggsave3(paste0("indole_analysis/", x, "_NMDS.pdf"), device = "pdf", dpi = 300, height = 4, width = 6))

```

##### Coloured by stroke severity 

```{r}
cont_pal <- c("#FFFFFF","#FFC3C6FF","#FB535EFF", "#F94A56FF","#EA2439FF", "#D51F32FF", "#B01526FF", "#9C0F20FF", "#6C000EFF")
```


```{r}
plot_beta_div(metaphlan_ileum, beta_metaphlan$Distance_Matrix, beta_metaphlan$Ordination, group_variable = "severity_score") +
  scale_color_gradientn(colours = cont_pal, guide = guide_colorbar(frame.colour = "black", ticks.colour = "white", frame.linewidth = 0.75, ticks.linewidth = 0.75, draw.ulim = F)) +
  #paletteer::scale_color_paletteer_c("grDevices::Reds 3",direction = -1) +
    ggtitle("severity score") +
    ggsave3("indole_analysis/severity_score_NMDS_rw.pdf", device = "pdf", dpi = 300, height = 4, width = 5)
```
tnaA same colors as stroke severity 

```{r}
plot_beta_div(metaphlan_ileum, beta_metaphlan$Distance_Matrix, beta_metaphlan$Ordination, group_variable = "EC.4.1.99.1_Tryptophanase") +
  scale_color_gradientn(colours = cont_pal, guide = guide_colorbar(frame.colour = "black", ticks.colour = "white", frame.linewidth = 0.75, ticks.linewidth = 0.75, draw.ulim = F)) +
    ggtitle("EC.4.1.99.1_Tryptophanase \n (relative abundance)") +
    ggsave3("indole_analysis/EC.4.1.99.1_Tryptophanase_NMDS_alt_col.pdf", device = "pdf", dpi = 300, height = 4, width = 5.5)
```


### Taxonomic composition 

Mean by group

```{r, fig.height=8, fig.width=14}
metaphlan_ileum <- ps_arrange(metaphlan_ileum, "severity_score")
p <- plot_taxonomic_comp(metaphlan_ileum, "Species", "severity", n_taxa = 14, ord=c("mild", "moderate", "severe")) + 
  theme(strip.background = element_rect(fill="white", color = "black"),
        strip.text = element_text(size=18))
print(p)
```


```{r}
metaphlan_genus <- aggregate_taxa(metaphlan_ileum, level="Genus")

p <- plot_taxonomic_comp(metaphlan_genus, "Genus", "severity",merge=T, n_taxa = 10, ord=c("mild", "moderate", "severe")) + 
  theme(strip.background = element_rect(fill="white", color = "black"),
        strip.text = element_text(size=18))
print(p)
ggsave("tax_figures/taxonomic_comp_merged_by_group.pdf", device=cairo_pdf, dpi=300, height=5, width=5.25)
```



```{r, fig.height=8, fig.width=14}
metaphlan_species_stroke <- ps_arrange(metaphlan_species, "severity_score") %>% 
  subset_samples(condition == "fMCAO")
p <- plot_taxonomic_comp(metaphlan_species_stroke, "Species", "severity", n_taxa = 14, ord=c("moderate", "severe")) + 
  theme(strip.background = element_rect(fill="white", color = "black"),
        strip.text = element_text(size=18))
print(p)
```
```{r}
ggsave(filename = "tax_figures/taxonomic_comp_fMCAO_metaphlan.pdf", width = 14, height = 8, dpi=300, device = "pdf")
```


## Differential abundance 

### Filtering 

Select ileum data for FC analysis
```{r}
metaphlan_species_il_filt <- metaphlan_species %>% 
  subset_samples(sample_type == "ileum") %>% 
  microViz::tax_filter(min_prevalence = 0.33, prev_detection_threshold = 0, min_sample_abundance =1e-6, use_counts = F) 
```


Full dataset - not excluding the extreme samples with severe weight loss

consider adjusting for severity instead of separating 

#### Fold change 


```{r}
get_species <- function(df, col) {
  df_copy <- df
  df_copy[[col]] <- sub(".*\\|", "", df_copy[[col]])
  return(df_copy)
}

ggsave3 <- function(...) {
  ggplot2::ggsave(...)
  invisible()
}
```

Plot volcano from functional analysis script differs from taxonomic analysis one and this will fail with x matched by multiple actual arguments error if functional script sourced above after taxonomic. Either differentiate between the two or make them the same. 
```{r}
volcano_pal <- c("#E16036", "#466365", "#D6D6D6")
fc_res <- tax_fold_change(metaphlan_species_il_filt, var = "condition", case = "fMCAO", ctrl = "sham")
plot_volcano(results_df=fc_res, pthresh=0.1, FCthresh=0.58, x="log2FC", y="p.adj", groups=c("Sham", "fMCAO"), pal=volcano_pal) +
  geom_text_repel(
    inherit.aes = FALSE,
    data = filter(
      get_species(fc_res$res %>%
    rownames_to_column("taxonomy"), "taxonomy"),
      p.adj < 0.1 & abs(log2FC) > 1
    ),
    size = 3,
    max.overlaps = 10,
    aes(x = log2FC, y = -log10(p.adj), label = taxonomy)
  )
ggsave("tax_figures/volcano_shamvsfmcao.pdf", dpi = 300, height = 5, width = 6)
```

```{r, message=FALSE, echo=FALSE, warning=FALSE}
fc_res_moderate <- tax_fold_change(subset_samples(metaphlan_species_il_filt, severity != "severe"), var = "condition", case = "fMCAO", ctrl = "sham")
```


### Sham vs severe

#### Fold change 

```{r, message=FALSE, echo=FALSE, warning=FALSE}
fc_res_severe <- tax_fold_change(subset_samples(metaphlan_species_il_filt, severity != "moderate"), var = "condition", 
                 case = "fMCAO", ctrl = "sham")
```


Mild vs Moderate, Moderate vs Severe 
```{r}
fc_strat <- bind_rows(
  fc_res_moderate$res %>%
    rownames_to_column("taxonomy") %>%
    mutate(comparison = "Mild vs Moderate"),
  fc_res_severe$res %>%
    rownames_to_column("taxonomy") %>%
    mutate(comparison = "Mild vs Severe")
)

volcano_strat <- plot_volcano(
  results_df = fc_strat,
  pthresh = 0.1,
  FCthresh = 0.58,
  x = "log2FC",
  y = "p.adj",
  groups = c("Mild", "Moderate"),
  pal = volcano_pal
) +
  facet_grid(~comparison) +
  geom_text_repel(
    inherit.aes = FALSE,
    data = filter(
      get_species(fc_strat, "taxonomy"),
      p.adj < 0.1 & abs(log2FC) > 1
    ),
    size = 4,
    max.overlaps = 15,
    aes(x = log2FC, y = -log10(p.adj), label = taxonomy)
  )
volcano_strat
ggsave("tax_figures/volcano_strat.pdf", dpi = 300, height = 5, width = 10)
```

Diverging bar for Log2 FC

```{r}
# , , "#C2C5C5FF", "#AE123AFF", "#E4C7C1FF"
pval_colors <- c("#889296FF", "#A7ADAEFF", "#949C9FFF", "#F59885FF", "#ED5F55FF", "#DB3842FF", "#D13140FF")
pvals <- c(0.00, 0.30, 0.60, 1, 1.3,1.60, 1.9, 2.25)
```

```{r}
colpal2 <- c(cols[3:5], "lightgrey")
names(colpal2) <- c("mild", "moderate", "severe", "ns")
fc_strat %>%
  mutate(enrichment = case_when(log2FC < 1 ~ "depleted",
  log2FC > 1 ~ "enriched")) %>% 
  # filter to top 10+
  group_by(comparison, enrichment) %>%
    mutate(`-log10(qval)` = -log10(p.adj)) %>% 
  slice_max(abs(log2FC), n = 10) %>%
  ungroup() %>% 
  mutate(Species = sub(".*\\|", "", taxonomy)) %>%
  mutate(
    signif = case_when(
      p.adj < 0.1 & log2FC < 0 & comparison == "Mild vs Moderate" ~ "mild",
      p.adj < 0.1 & log2FC < 0 & comparison == "Mild vs Severe" ~ "mild",
      p.adj < 0.1 & log2FC > 0 & comparison == "Mild vs Moderate" ~ "moderate",
      p.adj < 0.1 & log2FC > 0 & comparison == "Mild vs Severe" ~ "severe",
      TRUE ~ "ns"
    )
  ) %>%
  ggplot(aes(x = log2FC, y = reorder_within(Species, log2FC, comparison), fill = `-log10(qval)`)) +
  geom_col() +
  scale_y_reordered() +
  theme_cowplot() +
 scale_fill_gradientn(
  colours = pval_colors,
  values = scales::rescale(pvals),
  limits = c(0, 2.25)
) + 
  #paletteer::scale_fill_paletteer_c("ggthemes::Red-Black Diverging", direction=-1) + 
  theme(panel.border = element_blank(), axis.line = element_line()) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18, face = "italic"),
    axis.title = element_text(size = 20),
    strip.background = element_rect(fill="white", color="black"),
    legend.text = element_text(size = 18)
  ) +
  xlab("Log2FC") +
  ylab("") +
  facet_wrap(vars(comparison), scales = "free")
ggsave("tax_figures/fc_sham_vs_moderate_vs_severe_barplot-new_alternate.pdf", device = "pdf", dpi = 300, height = 5.5, width = 16)
```

Diverging dotplot for Log2 FC

### Moderate vs severe

```{r}
fc_res_mod_severe <- tax_fold_change(subset_samples(metaphlan_species_il_filt, severity != "mild"), var = "severity", 
                 case = "moderate", ctrl = "severe")
```

### tnaA encoding 

```{r}
# list comes from here: http://eggnog5.embl.de/#/app/results#COG3033_datamenu
indole_species <- read_csv("tnaA_encoding_taxa.csv")
# remove those without species annotation - anything containing sp. or bacterium
drop <- c("sp.", "bacterium") 
indole_species <- indole_species %>% 
  filter(!grepl(paste(drop, collapse="|"), Organism)) 
# fix [] and add format to s__Genus_species
indole_species <- indole_species$Organism

indole_species <- gsub("\\[", "", indole_species)
indole_species <- gsub("\\]", "", indole_species)

indole_species <- gsub(" ", "_", indole_species)
indole_species <- paste0("s__", indole_species)
indole_species <- unique(indole_species)
```


```{r}
metaphlan_species_indole <- metaphlan_species %>%
  subset_samples(sample_type == "ileum") %>% 
  transform("arcsin") %>% 
  psmelt() %>% 
  filter(grepl(paste(indole_species, collapse = "|"), Species)) 

```

```{r}
metaphlan_species_indole %>% 
  filter(severity != "severe") %>% 
  filter(OTU == "k__Bacteria|p__Verrucomicrobia|c__Verrucomicrobiae|o__Verrucomicrobiales|f__Akkermansiaceae|g__Akkermansia|s__Akkermansia_muciniphila") %>% 
  pairwise_wilcox_test(Abundance~severity)
```

```{r}
metaphlan_species_indole %>% 
  filter(severity != "moderate") %>% 
  filter(OTU != "k__Bacteria|p__Firmicutes|c__Clostridia|o__Eubacteriales|f__Oscillospiraceae|g__Anaerotruncus|s__Anaerotruncus_colihominis") %>% 
  group_by(OTU) %>% 
  pairwise_wilcox_test(Abundance~severity)
```

```{r}
list_int_taxa <- unique(metaphlan_species_indole$OTU)
indole_producer_plots <- map(list_int_taxa, function(x) {
  
  dat <- metaphlan_species_indole %>%
    filter(OTU == x) 
  label <- gsub("s__","", unique(dat$Species))
  label <- gsub("_", " ", label)

  dat %>% 
    ggstripchart(
      x = "condition",
      y = "Abundance",
      ylab = paste0(label, "\n", "Asin(sqrt(rel. abundance))"),
      palette = cols,
      error.plot = "errorbar",
      color = "condition",
      add =c("mean_sd"),
      size=4,
      alpha=0.75,
      add.params = list(color = "black", size=0.5),
      order = c("sham", "fMCAO"),
      position = position_jitterdodge(0.1)
    ) +
      theme(axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      legend.title = element_blank()) + 
    stat_compare_means(comparisons = list(c("sham", "fMCAO")),
                       method = "wilcox.test",
                       label = "p.format") +
    stat_summary(fun = mean, geom = "errorbar",
                 aes(ymax = after_stat(y), ymin = after_stat(y)),
                 width = 0.2, 
                 color = "black",
                 size = 0.5) 
} ) 

 
indole_producer_plots
names(indole_producer_plots) <- unique(metaphlan_species_indole$Species)
imap(indole_producer_plots, function(.x, .y) ggsave(paste0("tax_figures/", .y, "_final_dotplot.pdf"), .x, device = cairo_pdf, dpi=300, height = 4, width = 3.75))
```
```{r}
indole_producer_plots_sev <- map(list_int_taxa, function(x) {
  
  dat <- metaphlan_species_indole %>%
    filter(OTU == x)
  label <- gsub("s__","", unique(dat$Species))
  label <- gsub("_", " ", label)

  dat %>%
    ggstripchart(
      x = "severity",
      y = "Abundance",
      ylab = paste0(label, "\n", "Asin(sqrt(rel. abundance))"),
      palette = cols,
      error.plot = "errorbar",
      color = "severity",
      add =c("mean_sd"),
      size=4,
      alpha=0.75,
      add.params = list(color = "black", size=0.5),
      order = c("mild", "moderate", "severe"),
      position = position_jitterdodge(0.1)
    ) +
      theme(axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      legend.title = element_blank()) + 
    stat_compare_means(comparisons = list(c("mild", "moderate"), c("mild", "severe"), c("moderate", "severe")),
                       method = "wilcox.test",
                       label = "p.format") +
    stat_summary(fun = mean, geom = "errorbar",
                 aes(ymax = after_stat(y), ymin = after_stat(y)),
                 width = 0.2, 
                 color = "black",
                 size = 0.5) 

} )

indole_producer_plots_sev
names(indole_producer_plots_sev) <- unique(metaphlan_species_indole$Species)
imap(indole_producer_plots_sev, function(.x, .y) ggsave(paste0("tax_figures/", .y, "_sev_dotplot.pdf"), .x, device = cairo_pdf, dpi=300, height = 4, width = 3.25))
```

Plot E. coli individually 

Use sample plot style etc for other indole producers below and facet by Species
```{r}
e_coli_dat <- metaphlan_species_indole %>%
    filter(OTU == "k__Bacteria|p__Proteobacteria|c__Gammaproteobacteria|o__Enterobacterales|f__Enterobacteriaceae|g__Escherichia|s__Escherichia_coli" )

e_coli_dat %>%
    ggstripchart(
      x = "severity",
      y = "Abundance",
      ylab = "Escherichia coli \n Asin(sqrt(rel. abundance))",
      palette = cols,
      error.plot = "errorbar",
      color = "severity",
      add =c("mean_sd"),
      size=4,
      alpha=0.75,
      add.params = list(color = "black", size=0.5),
      order = c("mild", "moderate", "severe"),
      position = position_jitterdodge(0.1)
    ) +
      theme(axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      legend.title = element_blank()) + 
    stat_compare_means(comparisons = list(c("mild", "severe"), c("moderate", "severe")),
                       method = "wilcox.test",
                       label = "p.format") +
    stat_summary(fun = mean, geom = "errorbar",
                 aes(ymax = after_stat(y), ymin = after_stat(y)),
                 width = 0.2, 
                 color = "black",
                 size = 0.5) 
ggsave(paste0("tax_figures/Escherichia_coli_sev_dotplot_final.pdf"), device = "pdf", dpi=300, height = 4, width = 3.75)
```
### ML regression 

```{r}
library(SIAMCAT)
library(ranger)
```
```{r}
label <- create.label("severity_score",meta = meta_to_df(metaphlan_ileum))
sc_ileum <- siamcat(phyloseq = metaphlan_ileum, label = label)
```

Filtering and normalisation 
```{r}
sc_ileum <- filter.features(sc_ileum, filter.method = "abundance", cutoff = 1e-6) %>% 
  filter.features(filter.method = "prevalence", cutoff = 0.1) %>%
  normalize.features(norm.method = "log.std", norm.param = list(log.n0 = 1e-06, sd.min.q = 0.1))
```

```{r}
sc_ileum <- create.data.split(sc_ileum, 
                              num.folds = nsamples(metaphlan_ileum), 
                              num.resample = 2) 

sc_ileum <- sc_ileum %>% 
  train.model(method="randomForest", measure="regr.rmse", verbose=3)
```

Model evaluation

```{r}
sc_ileum <- sc_ileum %>% 
  make.predictions() %>%
  evaluate.predictions()
```

```{r}
model.evaluation.plot('Severity score'=sc_ileum, fn.plot="tax_figures/roc_curve_severity.pdf",
    colours=c('orange'))
model.interpretation.plot(sc_ileum, fn.plot="tax_figures/feat_severity.pdf",consens.thres = 0.001, max.show=50)
```

### Analysis stool 

We do not see anything in terms of alpha-diversity in stool and shifts in composition are minimal. This fits with hypothesis that main changes are occuring in the ileum. However, we do want to show stool data in supplement. 

```{r}
metaphlan_stool <- metaphlan_species %>% 
  subset_samples(sample_type == "stool") %>% 
  transform()

```
```{r}
metaphlan_genus_st <- aggregate_taxa(metaphlan_stool, level="Genus")

p <- plot_taxonomic_comp(metaphlan_genus_st, "Genus", "severity",merge=T, n_taxa = 10, ord=c("mild", "moderate", "severe")) + 
  theme(strip.background = element_rect(fill="white", color = "black"),
        strip.text = element_text(size=18))
print(p)
ggsave("tax_figures/taxonomic_comp_merged_by_group_stool.pdf", device=cairo_pdf, dpi=300, height=5, width=5.25)
```

```{r}
metaphlan_species_st_filt <- metaphlan_stool %>% 
  microViz::tax_filter(min_prevalence = 0.33, prev_detection_threshold = 0, min_sample_abundance =1e-6, use_counts = F) 
```

```{r}
fc_res_moderate_st <- tax_fold_change(subset_samples(metaphlan_species_st_filt, severity != "severe"), var = "condition", case = "fMCAO", ctrl = "sham")
fc_res_severe_st <- tax_fold_change(subset_samples(metaphlan_species_st_filt, severity != "moderate"), var = "condition", 
                 case = "fMCAO", ctrl = "sham")
```

```{r}
fc_strat_st <- bind_rows(
  fc_res_moderate_st$res %>%
    rownames_to_column("taxonomy") %>%
    mutate(comparison = "Mild vs Moderate"),
  fc_res_severe_st$res %>%
    rownames_to_column("taxonomy") %>%
    mutate(comparison = "Mild vs Severe")
)
colpal2 <- c(cols[3:5], "lightgrey")
names(colpal2) <- c("mild", "moderate", "severe", "ns")
fc_strat_st %>%
  mutate(enrichment = case_when(log2FC < 1 ~ "depleted",
  log2FC > 1 ~ "enriched")) %>% 
  # filter to top 10+
  group_by(comparison, enrichment) %>%
    mutate(`-log10(qval)` = -log10(p.adj)) %>% 
  slice_max(abs(log2FC), n = 10) %>%
  ungroup() %>% 
  mutate(Species = sub(".*\\|", "", taxonomy)) %>%
  mutate(
    signif = case_when(
      p.adj < 0.1 & log2FC < 0 & comparison == "Mild vs Moderate" ~ "mild",
      p.adj < 0.1 & log2FC < 0 & comparison == "Mild vs Severe" ~ "mild",
      p.adj < 0.1 & log2FC > 0 & comparison == "Mild vs Moderate" ~ "moderate",
      p.adj < 0.1 & log2FC > 0 & comparison == "Mild vs Severe" ~ "severe",
      TRUE ~ "ns"
    )
  ) %>%
  ggplot(aes(x = log2FC, y = reorder_within(Species, log2FC, comparison), fill = `-log10(qval)`)) +
  geom_col() +
  scale_y_reordered() +
  theme_cowplot() +
 scale_fill_gradientn(
  colours = pval_colors,
  values = scales::rescale(c(0.28, 0.6, 0.95, 1.0, 1.4, 1.7, 2.01)),
  limits = c(0.28, 2.01)
) + 
  #paletteer::scale_fill_paletteer_c("ggthemes::Red-Black Diverging", direction=-1) + 
  theme(panel.border = element_blank(), axis.line = element_line()) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18, face = "italic"),
    axis.title = element_text(size = 20),
    strip.background = element_rect(fill="white", color="black"),
    legend.text = element_text(size = 18)
  ) +
  xlab("Log2FC") +
  ylab("") +
  facet_wrap(vars(comparison), scales = "free")
ggsave("tax_figures/fc_sham_vs_moderate_vs_severe_barplot-stool.pdf", device = "pdf", dpi = 300, height = 5.5, width = 16)
```

