---
title: "AhR_patient_analysis"
author: "Adam Sorbie"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# AhR at-risk and chronic analysis

```{r, include=FALSE}
library(gtsummary)
library(gt)
source("../../../Metagenomics/scripts/taxa_analysis.R")
source("../../../Metagenomics/scripts/shotgun_analysis_utils.R")
source("../../../Metagenomics/scripts/func_analysis.R")
conflicted::conflict_prefer_all("dplyr")
conflicted::conflict_prefer_all("purrr")
```

```{r}
# list comes from here: http://eggnog5.embl.de/#/app/results#COG3033_datamenu
indole_species <- read_csv("../../AhR_metagenomics/analysis/tnaA_encoding_taxa.csv")
# remove those without species annotation - anything containing sp. or bacterium
drop <- c("sp.", "bacterium") 
indole_species <- indole_species %>% 
  filter(!grepl(paste(drop, collapse="|"), Organism)) 
# fix [] and add format to s__Genus_species
indole_species <- indole_species$Organism

indole_species <- gsub("\\[", "", indole_species)
indole_species <- gsub("\\]", "", indole_species)

indole_species <- gsub(" ", "_", indole_species)
indole_species <- paste0("s__", indole_species)
indole_species <- unique(indole_species)
```

```{r}
load("../data/hammond_tax.RData")
load("../data/SMB_tax.RData")
load("../data/lledos_tax.RData")
# Functional profiles
load("../data/hammond_func.RData")
load("../data/lledos_func.RData")
cols <- c("healthy" = "#424143", "control" = "#424143", "At Risk" = "#EDEBD7", "stroke" = "#F7921E")
```

## Question 1. Do at-risk patients have higher numbers of indole-producers?

```{r, include=FALSE}
library(curatedMetagenomicData)
```

```{r}
conditions <- c("hypertension", "T1D", "T2D", "CAD", "hypercholesterolemia","control")
at_Risk_metadata <- sampleMetadata %>%
  # change to grepl
  filter(grepl(paste(conditions, collapse = "|"), study_condition)) %>%
  filter(body_site == "stool") %>%
  filter(age_category %in% c("senior", "adult")) %>%
  filter(is.na(days_from_first_collection)) %>%
  filter(antibiotics_current_use == "no") %>%
  # filter to central Europe - consider updating this to include southern europe
  filter(country %in% c("AUT","LUX","DEU","HUN", "SVK", "FRA"))
```

How many healthy/non-healthy samples do we have?

```{r}
table(at_Risk_metadata$disease)
```

Metacardis data

```{r}
# metacardis_meta <- at_Risk_metadata %>% 
#   filter(study_name == "MetaCardis_2020_a") 
#  
# table(metacardis_meta$disease)
```

Download in chunks and save objects before uploading to cluster with
high memory allowing us to merge

```{r}
# metacardis_diseases <- c("CAD", "CAD;T2D","healthy",  "T2D")
# for (i in metacardis_diseases){
#   uniref_dat <- metacardis_meta %>% 
#     filter(disease == i) %>%
#     returnSamples("gene_families")
# 
#   save(uniref_dat, file=paste0("F:/Metagenomics/", i, "_metacardis_uniref.RData"))
# }
```

Download mpa3 data

```{r}
metaphlan_cmd3 <- at_Risk_metadata %>%
  returnSamples("relative_abundance", rownames = "short")

```

```{r}
counts_cmd3 <- assay(metaphlan_cmd3)
tax_table <- rowData(metaphlan_cmd3)
meta_cmd3 <- colData(metaphlan_cmd3) %>% as.data.frame() %>% 
  mutate(CAD = case_when(grepl("CAD", disease) ~ 1,
                         TRUE ~ 0)) %>% 
   mutate(hypertension = case_when(grepl("hypertension", disease) ~ 1,
                         TRUE ~ 0)) %>% 
   mutate(diabetes = case_when(grepl("T1D", disease) ~ 1,
                               grepl("T2D", disease) ~ 1,
                         TRUE ~ 0)) %>% 
  mutate(obesity = case_when(BMI > 30 ~ 1,
                               TRUE ~ 0))


```

```{r}
ec_cmd3 <- read_tsv("../data/unstrat/humann_genefamilies_ec_unstratified.tsv") %>% 
  column_to_rownames("# Gene Family")

# check samples are consistent
colnames(ec_cmd3) %in% colnames(counts_cmd3)
```

Filter to only include samples also in EC data

```{r}
# keep_samples <- colnames(ec_cmd3)
# counts_cmd3 <- counts_cmd3[, colnames(counts_cmd3) %in% keep_samples]
# meta_cmd3 <- meta_cmd3[meta_cmd3$subject_id %in% keep_samples, ]
```

```{r}
# colnames(tax_table) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
# ps_cmd3 <- load_phylo(asvtab = assay(metaphlan_cmd3),taxa = tax_table, mapping =meta_cmd3)
# save(ps_cmd3, file="../data/cmd3_tax.RData")
```

```{r}
load("../data/cmd3_tax.RData")
```

```{r}
indole_species2 <- gsub("s__", "", indole_species)
indole_species2 <- gsub("_", " ", indole_species2)
df_indole_producers_cmd3 <- ps_cmd3 %>%
  transform("relative") %>% 
  transform("arcsin") %>% 
  psmelt() %>% 
  filter(grepl(paste(indole_species2, collapse = "|"), Species)) %>% 
  mutate(OTU = gsub("species:", "", OTU))

```

Add At Risk column

```{r}
df_indole_producers_cmd3 <- df_indole_producers_cmd3 %>% 
  mutate(group = case_when(disease != "healthy" ~ "At Risk",
                           TRUE ~ disease)) %>% 
  mutate(age_group = case_when(age < 65 & age >= 45 ~ "middle-aged",
                               age < 45 ~ "young",
                               age >= 65 ~ "elderly",
                               TRUE ~ NA))
```

Create new columns for each comparison - not possible to just plot e.g.
hypertension as here we would be plotting those with vs those without
(who may have CAD or T2D), rather than healthy controls vs hypertension

```{r}
df_indole_producers_cmd3 <- df_indole_producers_cmd3 %>% 
  mutate(hypertension2 = case_when(hypertension == 1 ~ "hypertension",
                                  disease == "healthy" ~ "healthy",
                                  TRUE ~ NA)) %>% 
  mutate(diabetes2 = case_when(diabetes == 1 ~ "diabetes",
                                  disease == "healthy" ~ "healthy",
                                  TRUE ~ NA)) %>% 
  mutate(CAD2 = case_when(CAD == 1 ~ "CAD",
                                  disease == "healthy" ~ "healthy",
                                  TRUE ~ NA)) %>% 
  mutate(obesity2 = case_when(obesity == 1 ~ "obesity",
                                  disease == "healthy" ~ "healthy",
                                  TRUE ~ NA)) 
```

Need to compare groups separately as some patients have multiple
comorbidities

can probably use map2 here to loop through conditions

```{r}
list_int_taxa_cmd3 <- unique(df_indole_producers_cmd3$OTU)
risk_factors <- c("hypertension2", "diabetes2", "CAD2", "obesity2")
# needs to be same length as above vector

list_int_taxa_cmd3_in <- rep(list_int_taxa_cmd3, times=length(risk_factors))
risk_factors <- rep(c("CAD2", "hypertension2", "diabetes2", "obesity2"), each=length(list_int_taxa_cmd3))

indole_producer_plots_cmd3 <- map2(list_int_taxa_cmd3_in, risk_factors, function(.x, .y){
  
  
  df_indole_producers_cmd3 %>%
    filter(!is.na( .data[[ .y ]])) %>% 
    filter(OTU == .x) %>%
    #ggplot(aes(x = fct_relevel(group, "healthy", "At Risk"), y = Abundance, color=group)) + 
    ggplot(aes(x = .data [[ .y ]], y = Abundance, color= .data[[ .y]])) + 
    geom_jitter(position = position_jitterdodge(jitter.width = 0.1), size=3.75, alpha=0.4) + 
    theme_cowplot() + 
    #scale_color_manual(values = cols)+
    scale_color_aaas() +
    ylab(paste0(sub(".*\\|", "", .x), "\n", "Asin(sqrt(rel. abundance))")) +
    xlab("") +
    theme(axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
    stat_compare_means(comparisons = list(c("healthy", gsub("2","", .y,))),
                       method = "wilcox.test",
                       label = "p.format")
}
)


indole_producer_plots_cmd3

names(indole_producer_plots_cmd3) <- unique(df_indole_producers_cmd3$Species)
#imap(indole_producer_plots_cmd3, function(.x, .y) ggsave(paste0("figures/", .y, "_cmd3__dotplot.pdf"), .x, device = "pdf", dpi=300, height = 4.5, width = 8))
```

How best to present this data - either we make one column for at Risk
and compare healthy to at Risk, or we calculate separately for each Risk
factor and

*E. coli* is main taxon of interest

Generate table with characteristics of CMD3 and plot as before with
diseases on x-axis

```{r}
df_indole_producers_cmd3_split <- df_indole_producers_cmd3 %>%
  # add obesity to disease column
  mutate(disease = case_when(obesity == 1 ~ paste0(disease, ";obesity"),
                             TRUE ~ disease)) %>% 
  mutate(disease = case_when(age_group == "elderly" ~ paste0(disease, ";age (>65)"),
                             TRUE ~ disease)) %>% 
  separate_rows(disease, sep = ";") %>%
  mutate(disease = trimws(disease)) %>% 
# remove fatty liver
  filter(disease != "fatty_liver") 
```

```{r}
disease_cols <- c("healthy" = "#424143", "age (>65)" = "#B38D97", "CAD" = "#EDEBD7", "hypertension" = "#9CC4B2", "obesity" = "#F9AFAF", "T2D" = "#F9D5A7")
comps_dis <- make_pairwise_comparisons(df_indole_producers_cmd3_split, "disease")
comps_dis <- comps_dis[grepl("healthy", comps_dis)]
# order alphabetically
comps_dis <- comps_dis[order(sapply(comps_dis, function(x) x[1]))]

# remove E. fergusonii and low prevalent taxa
list_int_taxa_cmd3 <- list_int_taxa_cmd3[!list_int_taxa_cmd3 %in% c("Escherichia fergusonii", "Fusobacterium nucleatum","Desulfovibrio vulgaris","Escherichia albertii", "Escherichia sp. ESNIH1", "Porphyromonas gingivalis")]
for (i in list_int_taxa_cmd3){
  
  stats_cmd3 <- df_indole_producers_cmd3_split %>%
  mutate(disease = factor(disease, levels=c("healthy","age (>65)", "CAD", "hypertension", "obesity", "T2D"))) %>% 
  filter(OTU == i) %>% 
  pairwise_wilcox_test(
          Abundance ~ disease,
          comparisons = comps_dis,
          p.adjust.method = "BH",
          paired = F
        ) %>%
        add_significance() %>% 
  add_xy_position(x="disease") 

df_indole_producers_cmd3_split %>% 
  filter(OTU == i) %>% 
ggplot(aes(x = fct_relevel(disease, "healthy","age (>65)", "CAD", "hypertension", "obesity", "T2D"), y = Abundance, color= disease)) + 
    geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
    theme_cowplot() + 
    #scale_color_manual(values = cols)+
    scale_color_manual(values = disease_cols) +
    ylab(paste0(i, "\n Asin(sqrt(rel. abundance)) \n curatedMetagenomicData3")) +
    xlab("") +
    theme(axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          legend.position = "None") +
  scale_y_continuous(limits = c(-0.1,1.1), breaks = seq(0,1, 0.2)) + 
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") +
        stat_pvalue_manual(
          stats_cmd3,
          label = "p.adj",
          hide.ns = T,
          inherit.aes = FALSE, bracket.nudge.y = 0.05, step.increase = 0.015) + 
  EnvStats::stat_n_text()
ggsave(paste0("figures/", i, "_cmd3.pdf"), device = "pdf", dpi=300, height = 4.5, width = 6)
}


```

E. coli stratified by sex 
```{r}
stats_e_coli_cmd3 <- df_indole_producers_cmd3_split %>%
  mutate(disease = factor(
    disease,
    levels = c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D")
  )) %>%
  drop_na(gender) %>%
  filter(OTU == "Escherichia coli") %>%
  group_by(gender) %>%
  pairwise_wilcox_test(
    Abundance ~ disease,
    comparisons = comps_dis,
    p.adjust.method = "BH",
    paired = F
  ) %>%
  add_significance() %>%
  add_xy_position(x = "disease")

df_indole_producers_cmd3_split %>%
  filter(OTU == "Escherichia coli") %>%
  drop_na(gender) %>%
  ggplot(aes(
    x = fct_relevel(
      disease,
      "healthy",
      "age (>65)",
      "CAD",
      "hypertension",
      "obesity",
      "T2D"
    ),
    y = Abundance,
    color = disease
  )) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.15),
    size = 3.75,
    alpha = 0.4
  ) +
  theme_cowplot() +
  #scale_color_manual(values = cols)+
  scale_color_manual(values = disease_cols) +
  ylab(paste0(i, "\n Asin(sqrt(rel. abundance)) \n curatedMetagenomicData3")) +
  xlab("") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "None",
    strip.background = element_rect(fill = "white", color = "black"),
  ) +
  scale_y_continuous(limits = c(-0.1, 1.25), breaks = seq(0, 1, 0.2)) +
  facet_wrap( ~ gender) +
  stat_summary(
    fun.y = median,
    fun.ymin = median,
    fun.ymax = median,
    geom = "crossbar",
    width = 0.15,
    color = "black"
  ) +
  stat_pvalue_manual(
    stats_e_coli_cmd3,
    label = "p.adj",
    hide.ns = T,
    inherit.aes = FALSE,
    bracket.nudge.y = -0.2,
    step.increase = 0.0015
  ) +
  EnvStats::stat_n_text() + 
  rotate_x_text(angle=45)
ggsave(
  "figures/e_coli_gender_strat_cmd3.pdf",
  device = "pdf",
  dpi = 300,
  height = 5,
  width = 8.5
)


```

##### Other indole producing taxa 

```{r}
stats_all_cmd3 <- df_indole_producers_cmd3_split %>%
  filter(OTU %in% list_int_taxa_cmd3) %>% 
  mutate(disease = factor(
    disease,
    levels = c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D")
  )) %>%
  group_by(OTU) %>%
  pairwise_wilcox_test(
    Abundance ~ disease,
    comparisons = comps_dis,
    p.adjust.method = "BH",
    paired = F
  ) %>%
  add_significance()
```




## Question 2. Do at-risk patients have higher levels of tryptophanase?

For paper analyses we focused on EC 4.1.99.1 - tryptophanase


Remove unmapped and ungrouped and convert to relative abundance then
arcsin transform

```{r}
df_ec_cmd3 <- ec_cmd3 %>%
        rownames_to_column("tmp") %>%
        filter(!tmp %in% c(
          "UNGROUPED", "UNMAPPED", "UNINTEGRATED"
        )) %>%
        column_to_rownames("tmp")
df_ec_cmd3 <- t_df(100 * t_df(df_ec_cmd3) / colSums(df_ec_cmd3)) 
df_ec_cmd3 <- asin_sqrt(df_ec_cmd3)
```

Merge with metdata and select tnaA column, renaming as Abundance

```{r}
df_tnaa_cmd3 <- df_ec_cmd3 %>%
  t_df() %>% 
  merge(meta_cmd3, by=0) %>%
  select(c(`4.1.99.1`, disease, age,gender, hypertension, diabetes, CAD, obesity, Row.names)) %>% 
  rename(Abundance = `4.1.99.1`)

```

Add At Risk column

```{r}
df_tnaa_cmd3 <- df_tnaa_cmd3 %>% 
  mutate(group = case_when(disease != "healthy" ~ "At Risk",
                           TRUE ~ disease)) %>% 
  mutate(age_group = case_when(age < 65 & age >= 45 ~ "middle-aged",
                               age < 45 ~ "young",
                               age >= 65 ~ "elderly",
                               TRUE ~ NA))
```

```{r}
df_tnaa_cmd3 <- df_tnaa_cmd3 %>% 
  mutate(hypertension2 = case_when(hypertension == 1 ~ "hypertension",
                                  disease == "healthy" ~ "healthy",
                                  TRUE ~ NA)) %>% 
  mutate(diabetes2 = case_when(diabetes == 1 ~ "diabetes",
                                  disease == "healthy" ~ "healthy",
                                  TRUE ~ NA)) %>% 
  mutate(CAD2 = case_when(CAD == 1 ~ "CAD",
                                  disease == "healthy" ~ "healthy",
                                  TRUE ~ NA)) %>% 
  mutate(obesity2 = case_when(obesity == 1 ~ "obesity",
                                  disease == "healthy" ~ "healthy",
                                  TRUE ~ NA)) 
```

```{r}
df_tnaa_cmd3_split <- df_tnaa_cmd3 %>%
  # add obesity to disease column
  mutate(disease = case_when(obesity == 1 ~ paste0(disease, ";obesity"),
                             TRUE ~ disease)) %>% 
    mutate(disease = case_when(age_group == "elderly" ~ paste0(disease, ";age (>65)"),
                             TRUE ~ disease)) %>% 
  separate_rows(disease, sep = ";") %>%
  mutate(disease = trimws(disease)) %>% 
# remove fatty liver
  filter(disease != "fatty_liver") 
```

```{r}
stats_tnaa_cmd3 <- df_tnaa_cmd3_split %>%
  mutate(disease = factor(disease, levels=c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D"))) %>% 
  pairwise_wilcox_test(
          Abundance ~ disease,
          comparisons = comps_dis,
          p.adjust.method = "BH",
          paired = F
        ) %>%
        add_significance() %>% 
  add_xy_position(x="disease") %>%
         filter(p.adj < 0.1)

df_tnaa_cmd3_split %>% 
ggplot(aes(x = fct_relevel(disease, "healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D"), y = Abundance, color= disease)) + 
    geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
    theme_cowplot() + 
    #scale_color_manual(values = cols)+
    scale_color_manual(values = disease_cols) +
  scale_y_continuous(limits = c(-0.00275, 0.034), breaks = seq(0, 0.03, 0.005)) + 
    ylab("tnaA \n Asin(sqrt(rel. abundance)) \n curatedMetagenomicData3") +
    xlab("") +
    theme(axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") +
        stat_pvalue_manual(
          stats_tnaa_cmd3,
          label = "p.adj",
          inherit.aes = FALSE, bracket.nudge.y = -0.002, step.increase = 0.0125) + 
  EnvStats::stat_n_text() 
  
ggsave("figures/tnaA_cmd3.pdf", device = "pdf", dpi=300, height = 4.5, width = 6)
```

Stratified by gender 
```{r}
stats_tnaa_cmd3 <- df_tnaa_cmd3_split %>%
  mutate(disease = factor(
    disease,
    levels = c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D")
  )) %>%
  drop_na(gender) %>%
  group_by(gender) %>%
  pairwise_wilcox_test(
    Abundance ~ disease,
    comparisons = comps_dis,
    p.adjust.method = "BH",
    paired = F
  ) %>%
  add_significance() %>%
  add_xy_position(x = "disease")

df_tnaa_cmd3_split %>%
  drop_na(gender) %>%
  ggplot(aes(
    x = fct_relevel(
      disease,
      "healthy",
      "age (>65)",
      "CAD",
      "hypertension",
      "obesity",
      "T2D"
    ),
    y = Abundance,
    color = disease
  )) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.15),
    size = 3.75,
    alpha = 0.4
  ) +
  theme_cowplot() +
  #scale_color_manual(values = cols)+
  scale_color_manual(values = disease_cols) +
  scale_y_continuous(limits = c(-0.00275, 0.034), breaks = seq(0, 0.03, 0.005)) + 
    ylab("tnaA \n Asin(sqrt(rel. abundance)) \n curatedMetagenomicData3") +
  xlab("") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "None",
    strip.background = element_rect(fill = "white", color = "black"),
  ) +
  facet_wrap( ~ gender) +
  stat_summary(
    fun.y = median,
    fun.ymin = median,
    fun.ymax = median,
    geom = "crossbar",
    width = 0.15,
    color = "black"
  ) +
  stat_pvalue_manual(
    stats_tnaa_cmd3,
    label = "p.adj",
    hide.ns = T,
    inherit.aes = FALSE,
    bracket.nudge.y = -0.2,
    step.increase = 0.0015
  ) +
  EnvStats::stat_n_text() + 
  rotate_x_text(angle=45)

ggsave(
  "figures/tnaa_gender_strat_cmd3.pdf",
  device = "pdf",
  dpi = 300,
  height = 5,
  width = 8.5
)


```


Need to compare groups separately again as some patients have multiple
comorbidities

```{r}
risk_factors <- c("hypertension2", "diabetes2", "CAD2", "obesity2")
# needs to be same length as above vector

tnaa_plots_cmd3 <- map(risk_factors, function(x){
  
  
  df_tnaa_cmd3 %>%
    filter(!is.na( .data[[ x ]])) %>% 
    #ggplot(aes(x = fct_relevel(group, "healthy", "At Risk"), y = Abundance, color=group)) + 
    ggplot(aes(x = .data [[ x ]], y = Abundance, color= .data[[ x ]])) + 
    geom_jitter(position = position_jitterdodge(jitter.width = 0.1), size=3.75, alpha=0.4) + 
    theme_cowplot() + 
    #scale_color_manual(values = cols)+
    scale_color_aaas() +
    ylab(paste0("EC 4.1.99.1 Tryptophanase", "\n", "Asin(sqrt(rel. abundance))")) +
    xlab("") +
    theme(axis.title = element_text(size=14),
          axis.text = element_text(size=12),
          legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
    stat_compare_means(comparisons = list(c("healthy", gsub("2","", x,))),
                       method = "wilcox.test",
                       label = "p.format")
}
)


tnaa_plots_cmd3

#imap(indole_producer_plots_cmd3, function(.x, .y) ggsave(paste0("figures/", .y, "_cmd3__dotplot.pdf"), .x, device = "pdf", dpi=300, height = 4.5, width = 8))
```

## Question 3. Are indole-producers/tryptophanase enzymes enriched in acute phase/associated with severity or outcome?

Steps

load phyloseq and convert to rel abund -\> then arcsin transform melt to
df and select feature (E. coli and any Lactobacilli) Encode binary
values as 0-1 and convert to factor

Melt phyloseq and select important variables, ensure binary are all 0-1

#### Patient characteristics

```{r}
# remove haemorrhagic stroke and TIA samples 
ps_lledos <- ps_lledos %>% 
  subset_samples(bleed != 1) %>% 
  subset_samples(class != "TIA") 
```


```{r}
meta_lledos <- meta_to_df(ps_lledos) %>% 
  mutate(outcome = case_when(mRS_90d %in% 0:2 ~ "mRS 0-2",
                             mRS_90d <= 2 ~ "mRS 0-2",
                             mRS_90d %in% 3:6 ~ "mRS 3-6",
                             mRS_90d > 2 ~ "mRS 3-6")) 

cat_vars <- c("hypertension", "diabetes", "dyslipidemia", "atrial_fibrillation")
meta_lledos <- meta_lledos %>% 
  select(class, sex, age, hypertension, diabetes, dyslipidemia, atrial_fibrillation, obesity, outcome, nihss_in, nihss_24, TOAST, ) %>% 
  mutate(sex = replace(sex, sex == '?', NA)) %>% 
  # fix differences in TOAST 
  mutate(TOAST = recode(TOAST, "LAA" = "Large artery atherosclerosis",
                        "Large Artery Atherosclerosis" = "Large artery atherosclerosis",
                        "Large arthery atherosclerosis" = "Large artery atherosclerosis",
                        "Athero" = "Large artery atherosclerosis",
                        "Cardiac" = "Cardioembolic",
                        "Cardiac (AF)" = "Cardioembolic",
                        "undertermined non-ESUS" = "Undetermined",
                        "Unknown" = "Undetermined",
                        "CE" = "Cardioembolic",
                        "Cardioembolism" = "Cardioembolic",
                        "Other ethiology" = "Other etiology")) %>% 
  mutate_at(cat_vars, factor) 

```

```{r}
table2 <-
  tbl_summary(
    meta_lledos,
    missing = "no", 
    by = "class",
    label = list(sex = "Sex", age = "Age", hypertension = "Hypertension", diabetes = "Diabetes",dyslipidemia = "Dyslipidemia",atrial_fibrillation = "Atrial fibrillation", nihss_in = "NIHSS score admission", nihss_24 = "NIHSS 24h") # don't list missing data separately
  ) %>%  
  add_n() %>%  # add column with total number of non-missing observations
  add_p() %>%  # test for a difference between groups
  bold_labels()
table2 %>%  
  as_gt() %>%  
  gt::gtsave(filename = "Baseline_characteristics.docx")
```

#### Differential abundance

Class

```{r}
da_res <- ps_lledos %>% 
  subset_samples(class %in% c("control", "stroke")) %>%
  filter_ps(abun = 1e-6, prev=0.33) %>% 
  transform("relative") %>%
  transform("arcsin") %>%
  maaslin2_tax(out="lledos_da_tax", fixed="class")
da_res_df <- da_res$results
da_res_df$Species <-  gsub(".*\\.(s__[^.]*).*", "\\1", da_res_df$feature)
```

```{r}
da_res_df %>%
  filter(qval < 0.05) %>% 
  slice_max(abs(coef), n=40) %>% 
  mutate(enriched_in = case_when(coef < 0 ~ "control", TRUE ~ "stroke")) %>%
  arrange(coef) %>%
  mutate(Species = factor(Species, levels=Species)) %>% 
  ggplot(aes(x = Species, y = coef, color = enriched_in)) +
  #scale_x_discrete(labels= da_res_df$Species) + 
  geom_point(size = 5) +
  labs(y = "Coefficient", x = "") +
  theme_bw() +
  theme(
    axis.text.x = element_text(color = "black", size = 14),
    axis.text.y = element_text(color = "black", size = 14),
    axis.title.y = element_text(size = 16),
    axis.title.x = element_text(size = 16),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.position = "none"
  ) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  scale_color_manual(values = cols)
ggsave("figures/DA_species_maaslin2_stroke_ctrl.pdf", device = "pdf", dpi=300, height = 7.25, width = 8)
```

### Bubble plot overview

How does this compare to stroke risk factors? 

Bubble plot control vs risk factors and stroke

prep stroke data 

Some of the disease groupings are not 1:1 between cmd3 and the stroke data, e.g.
T2D and Diabetes (unclear if T1D, T2D or both). To avoid making any comparison
which may not make sense, here we will only plot stroke samples, in a separate 
category and not distinguish between those with or without comorbidities for the
purpose of this plot. 

```{r}
df_indole_producers_stroke_split <- ps_lledos %>%
  transform("relative") %>% 
  transform("arcsin") %>% 
  speedyseq::psmelt() %>% 
  mutate(Species =  gsub("s__", "", Species)) %>% 
  mutate(Species = gsub("_", " ", Species)) %>% 
  filter(grepl(paste(indole_species2, collapse = "|"), Species)) %>% 
  filter(class == "stroke") %>%
  mutate(OTU = gsub("species:", "", OTU)) %>% 
  mutate(age_group = case_when(age < 65 & age >= 45 ~ "middle-aged",
                               age < 45 ~ "young",
                               age >= 65 ~ "elderly",
                               TRUE ~ NA)) %>% 
  mutate(disease = "stroke") %>% 
  mutate(OTU = Species)

  
```

rename gender column to sex so its the same as stroke dataset
```{r}
df_indole_producers_cmd3_split <- df_indole_producers_cmd3_split %>%
  rename(sex=gender)
```
```{r}
common_cols <- intersect(colnames(df_indole_producers_cmd3_split), colnames(df_indole_producers_stroke_split))
df_indole_producers_stroke_split <- df_indole_producers_stroke_split %>%
  select(all_of(common_cols))
df_indole_producers_cmd3_split <- df_indole_producers_cmd3_split %>%
  select(all_of(common_cols)) %>% 
  mutate(smoker = case_when(smoker == "yes" ~ 1,
                            smoker == "no" ~ 0, 
                            TRUE ~ NA)) %>% 
  mutate(sex = recode(sex, "male" = "M", "female" = "F"))
```

Find which species are present in cmd3 but not stroke and add dummy 0 values for
stroke dataset. 

```{r}
missing_species <- setdiff(list_int_taxa_cmd3, unique(df_indole_producers_stroke_split$Species))
df_indole_producers_stroke_split
```

```{r}
# Get taxonomic info for missing species, taking first occurrence of each
missing_species_info <- df_indole_producers_cmd3_split %>%
  filter(Species %in% missing_species) %>%
  group_by(Species) %>%
  slice_head(n = 1) %>%  # Take first occurrence of each species
  ungroup() %>%
  select(Species, Kingdom, Phylum, Class, Order, Family, Genus, OTU)

# Get sample metadata from df
sample_metadata <- df_indole_producers_stroke_split %>%
  select(Sample, disease, age, country, smoker, hypertension, sex,
         diabetes, obesity, age_group) %>%
  distinct()

# Create missing data
missing_data <- sample_metadata %>%
  crossing(missing_species_info) %>%
  mutate(Abundance = 0)

df_indole_producers_stroke_split <- bind_rows(df_indole_producers_stroke_split, missing_data)
```


```{r}
disease_cols_stroke <- c("healthy" = "#424143", "age (>65)" = "#B38D97", "CAD" = "#EDEBD7", "hypertension" = "#9CC4B2", "obesity" = "#F9AFAF", "T2D" = "#F9D5A7", "stroke" = "#F7921E")

df_indole_producers_cmd3_stroke_split <- df_indole_producers_cmd3_split %>%
  bind_rows(df_indole_producers_stroke_split)

# calc prevalence here too 
mean_abundance_data <- df_indole_producers_cmd3_stroke_split %>% 
  filter(OTU %in% list_int_taxa_cmd3) %>% 
  filter(disease %in% c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D", "stroke"))%>%
  filter(!is.na(Species), !is.na(Abundance)) %>%
  group_by(Species, disease) %>%
  summarise(
    mean_abundance = mean(Abundance, na.rm = TRUE),
    n_samples = n(),
     prevalence = sum(Abundance > 0, na.rm = TRUE) / n(),
    .groups = 'drop'
  )

species_order <- mean_abundance_data %>%
  group_by(Species) %>%
  summarise(overall_mean = mean(mean_abundance, na.rm = TRUE), .groups = 'drop') %>%
  arrange(overall_mean) %>%
  pull(Species)

# Step 2: Set factor levels for proper ordering
mean_abundance_data <- mean_abundance_data %>%
  mutate(
    disease = factor(disease, 
                    levels = c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D", "stroke")),
    Species = factor(Species, levels = species_order)
  ) 
```

```{r}
species_rem <- mean_abundance_data %>% 
  group_by(Species, disease) %>% 
  summarise(mean_prev = mean(prevalence)) %>% 
  filter(mean_prev < 0.33) %>% 
  distinct(Species) %>% 
  pull(Species)

```


```{r}
comps_dis_stroke <- make_pairwise_comparisons(df_indole_producers_cmd3_stroke_split, "disease")
comps_dis_stroke <- comps_dis_stroke[grepl("healthy", comps_dis_stroke)]
# order alphabetically
comps_dis_stroke <- comps_dis_stroke[order(sapply(comps_dis_stroke, function(x) x[1]))]

stats_all_cmd3_stroke <- df_indole_producers_cmd3_stroke_split %>%
  filter(OTU %in% list_int_taxa_cmd3) %>% 
  # no sample has abundance > 0
  filter(!OTU %in% species_rem) %>%
  mutate(disease = factor(
    disease,
    levels = c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D", "stroke")
  )) %>%
  group_by(OTU) %>%
  pairwise_wilcox_test(
    Abundance ~ disease,
    comparisons = comps_dis_stroke,
    p.adjust.method = "BH",
    paired = F
  ) %>%
  add_significance()

stats_all_cmd3_stroke_bubble <- stats_all_cmd3_stroke %>% 
  mutate(
    Species = OTU,
    disease = ifelse(group1 == "healthy", group2, group1),
    p.adj.signi = ifelse(p.adj.signif == "ns", "", p.adj.signif)
  ) %>%
  select(Species, disease, p.adj.signi) %>%
  # Add reference (healthy) with no stars
  add_row(Species = unique(.$Species), disease = "healthy", p.adj.signi = "") %>% 
  mutate(Species = factor(Species, levels = species_order)) %>%
  arrange(Species, disease)
# order according to mean abundance data

mean_abundance_data %>%
  # no sample has abundance > 0
  filter(!Species %in% species_rem) %>% 
  ggplot(aes(x = disease, y = Species)) +
  geom_point(aes(size = mean_abundance, color = disease), 
             alpha = 0.8) +
  # Custom color palette
  scale_color_manual(values = disease_cols_stroke) +
  
  # Scale for bubble size
  scale_size_continuous(
    name = "Mean\nAbundance",
    range = c(2, 20),
    guide = guide_legend(override.aes = list(color = "black"))
  ) +
  
  # Theme
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 11, face = "italic"),
    axis.title = element_blank(),
    legend.title = element_text(size = 12),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 16, hjust = 0.5)
   # aspect.ratio = 3/4
  ) +
geom_text(data = stats_all_cmd3_stroke_bubble,
          aes(x = disease, y = Species, label = ifelse(p.adj.signi != "", "*", "")),
          size = 6,
          color = "black",
          vjust = 0.5,
          inherit.aes = FALSE) +
  coord_fixed(ratio = 1.025)
ggsave("figures/indole_producing_taxa_stroke.pdf", device = "pdf", dpi=300, height = 6.5, width = 9)
```

```{r}
stats_e_coli_cmd3_stroke <- df_indole_producers_cmd3_stroke_split %>%
  mutate(disease = factor(
    disease,
    levels = c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D", "stroke")
  )) %>%
  drop_na(sex) %>%
  filter(OTU == "Escherichia coli") %>%
  group_by(sex) %>%
  pairwise_wilcox_test(
    Abundance ~ disease,
    comparisons = comps_dis_stroke,
    p.adjust.method = "BH",
    paired = F
  ) %>%
  add_significance() %>%
  add_xy_position(x = "disease")

df_indole_producers_cmd3_stroke_split %>%
  filter(OTU == "Escherichia coli") %>%
  drop_na(sex) %>%
  ggplot(aes(
    x = fct_relevel(
      disease,
      "healthy",
      "age (>65)",
      "CAD",
      "hypertension",
      "obesity",
      "T2D",
      "stroke"
    ),
    y = Abundance,
    color = disease
  )) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.15),
    size = 3.75,
    alpha = 0.4
  ) +
  theme_cowplot() +
  #scale_color_manual(values = cols)+
  scale_color_manual(values = disease_cols_stroke ) +
  ylab("Escherichia coli \n Asin(sqrt(rel. abundance))") +
  xlab("") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "None",
    strip.background = element_rect(fill = "white", color = "black"),
  ) +
  scale_y_continuous(limits = c(-0.1, 1.25), breaks = seq(0, 1, 0.2)) +
  facet_wrap( ~ sex) +
  stat_summary(
    fun.y = median,
    fun.ymin = median,
    fun.ymax = median,
    geom = "crossbar",
    width = 0.15,
    color = "black"
  ) +
  stat_pvalue_manual(
    stats_e_coli_cmd3_stroke,
    label = "p.adj",
    hide.ns = T,
    inherit.aes = FALSE,
    bracket.nudge.y = -0.2,
    step.increase = 0.0015
  ) +
  EnvStats::stat_n_text() + 
  rotate_x_text(angle=45)
ggsave(
  "figures/e_coli_gender_strat_cmd3_stroke.pdf",
  device = "pdf",
  dpi = 300,
  height = 5,
  width = 8.5
)


```


#### tnaA bubble plot 


This part is not really relevant but need to generate df_tnaa_stroke_split so it has the same/similar format
```{r}
df_tnaa_stroke_split <- lledos_func %>%
  transform_func("relative") %>% 
  transform_func("arcsin") %>% 
  subset_samples_func(class == "stroke" & class != "TIA" & bleed != 1) %>%
  pluck("EC") %>% 
  rownames_to_column("EC") %>%
  filter(grepl("EC:4.1.99.1_", EC)) %>% 
  pivot_longer(-EC, names_to = "Sample", values_to = "Abundance") %>% 
  merge(filter(meta_lledos %>% rownames_to_column("Sample"), class=="stroke"), by="Sample") %>% 
  mutate(age_group = case_when(age < 65 & age >= 45 ~ "middle-aged",
                               age < 45 ~ "young",
                               age >= 65 ~ "elderly",
                               TRUE ~ NA)) %>% 
  mutate(disease = "stroke") 
```

Need to repeat this for tnaA dfs (df_tnaa_cmd3_split + find or create df_tnaa_stroke_split)

```{r}
df_tnaa_cmd3_split <- df_tnaa_cmd3_split %>% 
  rename(sex = gender) %>% 
  mutate(sex = recode(sex, "male" = "M", "female" = "F"))
```

```{r}
common_cols <- intersect(colnames(df_tnaa_cmd3_split), colnames(df_tnaa_stroke_split))
df_tnaa_stroke_split <- df_tnaa_stroke_split %>%
  select(all_of(common_cols))
df_tnaa_cmd3_split <- df_tnaa_cmd3_split %>%
  select(all_of(common_cols)) 

# make sure column types are the same 

df_tnaa_stroke_split <- df_tnaa_stroke_split %>%
  mutate(
    hypertension = as.numeric(as.character(hypertension)),
    diabetes = as.numeric(as.character(diabetes)),
    obesity = as.integer(obesity)
  )
```

Combine dataframes and calculate mean abundance across disease 
```{r}
df_tnaa_cmd3_stroke_split <- df_tnaa_cmd3_split %>%
  bind_rows(df_tnaa_stroke_split)

mean_abundance_data_tnaa <- df_tnaa_cmd3_stroke_split %>% 
  filter(disease %in% c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D", "stroke"))%>%
  group_by(disease) %>%
  summarise(
    mean_abundance = mean(Abundance, na.rm = TRUE),
    n_samples = n(),
    .groups = 'drop'
  )



# Step 2: Set factor levels for proper ordering
mean_abundance_data_tnaa <- mean_abundance_data_tnaa %>%
  mutate(
    disease = factor(disease, 
                    levels = c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D", "stroke")),
  ) 
```

This performs stats and plotting 

Many minor things need edited here

```{r}
stats_all_cmd3_stroke_tnaa <- df_tnaa_cmd3_stroke_split %>%
  mutate(disease = factor(
    disease,
    levels = c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D", "stroke")
  )) %>%
  pairwise_wilcox_test(
    Abundance ~ disease,
    comparisons = comps_dis_stroke,
    p.adjust.method = "BH",
    paired = F
  ) %>%
  add_significance()

stats_all_cmd3_stroke_bubble_tnaa <- stats_all_cmd3_stroke_tnaa %>% 
  mutate(
    disease = ifelse(group1 == "healthy", group2, group1),
    p.adj.signif = ifelse(p.adj.signif == "ns", "", p.adj.signif)
  ) %>%
  mutate(EC = "tnaA (EC 4.1.99.1)") %>%
  select(EC, disease, p.adj.signif) %>%
  # Add reference (healthy) with no stars
  add_row(EC = unique(.$EC), disease = "healthy", p.adj.signif = "") 
  
  # %>% 
  # mutate(EC = factor(EC, levels = species_order)) %>%
  # arrange(EC, disease)
# order according to mean abundance data

mean_abundance_data_tnaa %>%
  mutate(EC = "tnaA (EC 4.1.99.1)") %>% 
  # no sample has abundance > 0
  ggplot(aes(x = disease, y = EC)) +
  geom_point(aes(size = mean_abundance, color = disease), 
             alpha = 0.8) +
  # Custom color palette
  scale_color_manual(values = disease_cols_stroke) +
  
  # Scale for bubble size
  scale_size_continuous(
    name = "Mean\nAbundance",
    range = c(2, 20),
    guide = guide_legend(override.aes = list(color = "black"))
  ) +
  
  # Theme
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 11, face = "italic"),
    axis.title = element_blank(),
    legend.title = element_text(size = 12),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 16, hjust = 0.5)
   # aspect.ratio = 3/4
  ) +
geom_text(data = stats_all_cmd3_stroke_bubble_tnaa,
          aes(x = disease, y = EC, label = ifelse(p.adj.signif != "", "*", "")),
          size = 6,
          color = "black",
          vjust = 0.5,
          inherit.aes = FALSE) +
  coord_fixed(ratio = 1.025)
ggsave("figures/tnaA_riskfactors_stroke.pdf", device = "pdf", dpi=300, height = 4, width = 9)
```


```{r}
stats_tnaa_cmd3_stroke <- df_tnaa_cmd3_stroke_split %>%
  mutate(disease = factor(
    disease,
    levels = c("healthy", "age (>65)", "CAD", "hypertension", "obesity", "T2D", "stroke")
  )) %>%
  drop_na(sex) %>%
  group_by(sex) %>%
  pairwise_wilcox_test(
    Abundance ~ disease,
    comparisons = comps_dis_stroke,
    p.adjust.method = "BH",
    paired = F
  ) %>%
  add_significance() %>%
  add_xy_position(x = "disease")

df_tnaa_cmd3_stroke_split %>%
  drop_na(sex) %>%
  ggplot(aes(
    x = fct_relevel(
      disease,
      "healthy",
      "age (>65)",
      "CAD",
      "hypertension",
      "obesity",
      "T2D",
      "stroke"
    ),
    y = Abundance,
    color = disease
  )) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.15),
    size = 3.75,
    alpha = 0.4
  ) +
  theme_cowplot() +
  #scale_color_manual(values = cols)+
  scale_color_manual(values = disease_cols_stroke ) +
  ylab("tnaA \n Asin(sqrt(rel. abundance))") +
  xlab("") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "None",
    strip.background = element_rect(fill = "white", color = "black"),
  ) +
  #scale_y_continuous(limits = c(-0.00275, 0.04), breaks = seq(0, 0.04, 0.005)) + 
  facet_wrap( ~ sex) +
  stat_summary(
    fun.y = median,
    fun.ymin = median,
    fun.ymax = median,
    geom = "crossbar",
    width = 0.15,
    color = "black"
  ) +
  stat_pvalue_manual(
    stats_tnaa_cmd3_stroke,
    label = "p.adj",
    hide.ns = T,
    inherit.aes = FALSE,
    step.increase = 0.0015
  ) +
  EnvStats::stat_n_text() + 
  rotate_x_text(angle=45)
ggsave(
  "figures/tnaa_gender_strat_cmd3_stroke.pdf",
  device = "pdf",
  dpi = 300,
  height = 5,
  width = 8.5
)


```

Severity

```{r}
da_res_sev_mild_sev <- ps_lledos %>% 
  ps_mutate(severity = case_when(nihss_in < 5 ~ "mild (NIHSS <5)",
                                              nihss_in >= 5 & nihss_in < 15 ~ "moderate (NIHSS >5 <15)",
                                              nihss_in >= 15 ~ "severe (NIHSS >15)")) %>%
  subset_samples(severity %in% c("mild (NIHSS <5)" , "severe (NIHSS >15)")) %>%
  filter_ps(abun = 1e-6, prev=0.33) %>% 
  transform("relative") %>%
  transform("arcsin") %>%
  maaslin2_tax(out="lledos_da_tax_sev_mild_sev", fixed="severity")
da_res_sev_mild_sev_df <- da_res_sev_mild_sev$results
```

```{r}
da_res_sev_mod_sev <- ps_lledos %>% 
  ps_mutate(severity = case_when(nihss_in < 5 ~ "mild (NIHSS <5)",
                                              nihss_in >= 5 & nihss_in < 15 ~ "moderate (NIHSS >5 <15)",
                                              nihss_in >= 15 ~ "severe (NIHSS >15)")) %>%
  subset_samples(severity %in% c("mild (NIHSS <5)","moderate (NIHSS >5 <15)" , "severe (NIHSS >15)")) %>%
  filter_ps(abun = 1e-6, prev=0.33) %>% 
  transform("relative") %>%
  transform("arcsin") %>%
  maaslin2_tax(out="lledos_da_tax_sev", fixed="severity", reference="severity,mild (NIHSS <5)")
da_res_sev_df <- da_res_sev$results
```

Outcome

Make sure this mutate also doesn't add 0 to missing samples

```{r}
da_res_outcome <- ps_lledos %>% 
  ps_mutate(outcome = case_when(mRS_90d == "" ~ NA, 
                                mRS_90d %in% 0:2 ~ "Favourable",
                             mRS_90d <= 2 ~ "Favourable",
                             mRS_90d %in% 3:6 ~ "Poor",
                             mRS_90d > 2 ~ "Poor")) %>%
  subset_samples(outcome %in% c("Poor", "Favourable")) %>%
  subset_samples(class == "stroke") %>%
  filter_ps(abun = 1e-6, prev=0.33) %>% 
  transform("relative") %>%
  transform("arcsin") %>%
  maaslin2_tax(out="lledos_da_tax_outcome", fixed="outcome")
da_res_outcome_df <- da_res_outcome$results
da_res_outcome_df$Species <-  gsub(".*\\.(s__[^.]*).*", "\\1", da_res_outcome_df$feature)
```

```{r}
outcome_cols <- c("Favourable"= "dodgerblue", "Poor" = "firebrick1")
da_res_outcome_df %>%
  filter(qval < 0.05) %>% 
  mutate(enriched_in = case_when(coef < 0 ~ "Favourable", TRUE ~ "Poor")) %>%
  arrange(coef) %>%
  mutate(Species = factor(Species, levels=Species)) %>% 
  ggplot(aes(x = Species, y = coef, color = enriched_in)) +
  #scale_x_discrete(labels= da_res_df$Species) + 
  geom_point(size = 5) +
  labs(y = "Coefficient", x = "") +
  theme_bw() +
  theme(
    axis.text.x = element_text(color = "black", size = 14),
    axis.text.y = element_text(color = "black", size = 14),
    axis.title.y = element_text(size = 16),
    axis.title.x = element_text(size = 16),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.position = "none"
  ) +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  scale_color_manual(values = outcome_cols)
ggsave("figures/DA_species_maaslin2_outcome.pdf", device = "pdf", dpi=300, height = 4.25, width = 6.5)
```


### Indole producers

```{r}
#save(species_outcome, file = "../data/species_outcome.RData")
#load("../data/species_outcome.RData")
conflicted::conflicts_prefer(base::as.factor)
species_outcome <- ps_lledos %>%
  speedyseq::psmelt() %>% 
  filter(class == "stroke") %>% 
  # remove samples missing from tnaA analysis
  filter(! Sample %in% c("s137", "s177")) %>% 
  mutate(infection_any = case_when(respiratory_infection == 1 ~ 1,
                                   other_infection == 1 ~ 1,
                                   TRUE ~ 0)) %>%
  mutate(outcome = case_when(mRS_90d == "" ~ NA_real_,
                             mRS_90d %in% 0:2 ~ 0,
                             mRS_90d <= 2 ~ 0,
                             mRS_90d %in% 3:6 ~ 1,
                             mRS_90d > 2 ~ 1,
                             TRUE ~ NA)) %>% 
  mutate(severity = case_when(nihss_in < 5 ~ "mild (NIHSS <5)",
                                              nihss_in >= 5 & nihss_in < 15 ~ "moderate (NIHSS >5 <15)",
                                              nihss_in >= 15 ~ "severe (NIHSS >15)")) %>% 
  select(Species, Sample, Abundance, age, sex, hypertension, diabetes, nihss_in, severity, dyslipidemia, infection_any, outcome) %>% 
  drop_na() %>% 
  mutate(across(c(sex, hypertension, diabetes, dyslipidemia, infection_any, outcome), as.factor))
```

```{r}
e_coli_outcome <- species_outcome %>% 
  filter(Species == "s__Escherichia_coli") %>% 
  rename(s__Escherichia_coli=Abundance)

a_muciniphila_outcome <- species_outcome %>% 
  filter(Species == "s__Akkermansia_muciniphila") %>% 
  rename(s__Akkermansia_muciniphila=Abundance)

e_eligens_outcome <- species_outcome %>% 
  filter(Species == "s__Eubacterium_hallii") %>% 
  rename(s__Eubacterium_hallii=Abundance)

  
```

```{r}
# Variables to test
variables <- c("s__Escherichia_coli", "age", "sex", "hypertension", "diabetes", 
               "nihss_in", "dyslipidemia", "infection_any")

# Run univariate logistic regression
e_coli_univariate <- lapply(variables, function(var) {
  formula <- as.formula(paste("outcome ~", var))
  glm(formula, data = e_coli_outcome, family = binomial)
})

# e_coli_univariate <- e_coli_outcome %>%
#   select(all_of(variables), outcome) %>% 
#   tbl_uvregression(                         ## produce univariate table
#     method = glm,                           ## define regression want to run (generalised linear model)
#     y = outcome,                            ## define outcome variable
#     method.args = list(family = binomial),  ## define what type of glm want to run (logistic)
#     exponentiate = TRUE                     ## exponentiate to produce odds ratios (rather than log odds)
# )

e_coli_univariate_summary <- lapply(e_coli_univariate, function(model) {
  tidy(model, conf.int = TRUE, exponentiate = TRUE)
})

# Combine results into a single dataframe
e_coli_univariate_df <- do.call(rbind, e_coli_univariate_summary)

```

Akkermansia muciniphila


```{r}
# Variables to test
variables <- c("s__Akkermansia_muciniphila", "age", "sex", "hypertension", "diabetes", 
               "nihss_in", "dyslipidemia", "infection_any")

# Run univariate logistic regression
a_muci_univariate <- lapply(variables, function(var) {
  formula <- as.formula(paste("outcome ~", var))
  glm(formula, data = a_muciniphila_outcome, family = binomial)
})

a_muci_univariate_summary <- lapply(a_muci_univariate, function(model) {
  tidy(model, conf.int = TRUE, exponentiate = TRUE)
})

# Combine results into a single dataframe
a_muci_univariate_df <- do.call(rbind, a_muci_univariate_summary)

```

Akkermansia not significant but similar estimate

### Other indoles
```{r}
other_indole_producers <- c("s__Bifidobacterium_longum", "s__Bifidobacterium_bifidum","s__Bifidobacterium_adolescentis","s__Clostridium_sporogenes","s__Clostridium_perfringens", "s__Lactobacillus_reuteri", "s__Lactobacillus_acidophilus", "s__Lactobacillus_plantarum", "s__Bacteroides_fragilis", "s__Bacteroides_thetaiotaomicron", "s__Enterococcus_faecalis")
other_indole_producers_det <- other_indole_producers[other_indole_producers %in% species_outcome$Species]

```

Lactobacillus reuteri

```{r}
other_indole_producers_outcome <- map(other_indole_producers_det, function(species) {
  outcome_data <- species_outcome %>% 
    filter(Species == species) 
  model <- glm(outcome ~ Abundance, data = outcome_data, family = binomial)
  tidy(model, conf.int = TRUE, exponentiate = TRUE)
})
names(other_indole_producers_outcome) <- other_indole_producers_det
```

```{r}
other_indole_producers_outcome_df <- imap(other_indole_producers_outcome, \(x, y) {
  x %>%
    filter(term != "(Intercept)") %>%
    mutate(Species = y)
}) %>%
  bind_rows() %>% 
  relocate(Species) %>% 
  adjust_pvalue(method = "BH", p.col = "p.value", output.col = "qval") %>%
  select(Species, estimate, conf.low, conf.high, p.value, qval) %>% 
  mutate(
    across(starts_with("estimate"), ~round(.x, 2)),
    across(starts_with("conf"), ~round(.x, 2)),
    across(starts_with("qval"), ~round(.x, 2)),
    across(starts_with("p.value"), ~case_when(
      .x < 0.001 ~ "< 0.001",
      TRUE ~ as.character(round(.x, 3))
    ))
  ) %>% 
    mutate(Species = gsub("s__", " ", Species)) %>% 
    mutate(Species= gsub("_", " ", Species))
```

```{r}
other_indole_producers_outcome_df %>% 
  unite("95% CI", conf.low, conf.high, sep = ", ") %>%
  rename(
    "p-value" = p.value,
    "q-value" = qval
  ) %>%
gt() %>%
   tab_style(
    style = list(
      cell_text(style = "italic")
      ),
    locations = cells_body(
      columns = Species
    )
  ) %>% 
  gt::gtsave(filename = "univariate_log_regression_other_indole_producers.docx")
```

E hallii and F. prausnitzii negatively associated with outcome

```{r}
# e_eligens_univariate <- glm(outcome ~ s__Eubacterium_hallii, data=e_eligens_outcome, family = binomial)
# tidy(e_eligens_univariate, conf.int = T, exponentiate = T)
```

Plot -

Remove y-axis line and title reading on best way to present this tidy up
names tidy up significance

to calculate odds ratios: exp(coef(model))

```{r}
# Extract clean variable names
e_coli_univariate_df <- e_coli_univariate_df %>%
    filter(term != "(Intercept)") %>% 
  mutate(clean_term = gsub("[0-9]+|[MF]$|TRUE|FALSE", "", term))  # Remove levels, M/F, TRUE/FALSE

# Further customization for specific variables if needed
clean_mapping <- c("s__Escherichia_coli" = "E. coli","sex" = "Sex", "age" = "Age", "hypertension" = "Hypertension", 
                   "diabetes" = "Diabetes", "nihss_in" = "NIHSS Score Admission", 
                   "dyslipidemia" = "Dyslipidemia", "infection_any" = "Infection (Any)")

e_coli_univariate_df <- e_coli_univariate_df %>%
  mutate(clean_term = recode(clean_term, !!!clean_mapping))
```

```{r}
e_coli_univariate_df %>% 
  mutate(
    significance = case_when(
      p.value < 0.001 ~ paste0(round(estimate, digits = 2), " ***"),
      p.value < 0.01 ~  paste0(round(estimate, digits = 2), " **"),
      p.value < 0.05 ~  paste0(round(estimate, digits = 2), " *"),
      TRUE ~ paste0(round(estimate, digits = 2), " "),
    ),
    term = factor(term, levels = term)  # Maintain the order of variables
  ) %>% 
  ggplot(aes(x = clean_term, y = estimate)) +
  geom_point(size = 3, shape=15, aes(color=estimate > 1)) +  # Points for odds ratios
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.15) +
  geom_text(aes(label = significance), nudge_x = 0.4, hjust=0.5, size = 4, color = "black") +# Error bars
  scale_y_log10() +  # Log scale for odds ratios
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Univariate regression (n = 148)",
    x ="",
    y = "Odds Ratio (log scale)"
  ) +
  theme_cowplot() + 
  scale_color_manual(values=c("dodgerblue", "firebrick1")) +
  theme(axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.text.y = element_text(
    face = c(rep("plain", 3), "italic", rep("plain", 4))),
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.position = "None") + 
  geom_hline(yintercept = 1, linetype="dashed",color="lightgrey")
ggsave("figures/e_coli_univariate.pdf", device = "pdf", dpi=300, height = 5, width = 6)
```

Plot model

```{r}
multivariate_model <- glm(outcome ~ s__Escherichia_coli + age + sex + hypertension + diabetes + 
             nihss_in + dyslipidemia  + infection_any, 
             data = e_coli_outcome, family = binomial)
multivariate_model %>% 
  tbl_regression(exponentiate = T, label = as.list(clean_mapping), show_single_row = c("sex", "hypertension", "diabetes", "dyslipidemia", "infection_any")) %>%
  bold_p(t = 0.05) %>% 
  as_gt() %>% 
  gt::gtsave(filename = "multivariate_log_regression_e_coli.docx")
```

##### Group comparisons

```{r}
e_coli_stroke <- ps_lledos %>% 
  transform("relative") %>%
  transform("arcsin") %>%
  speedyseq::psmelt() %>% 
  filter(class %in% c("control", "stroke")) %>% 
  filter(Species == "s__Escherichia_coli")
```

```{r}
e_coli_stroke %>% 
  ggplot(aes(x = class, y = Abundance, color=class)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
  scale_color_manual(values = cols) + 
  theme_cowplot() + 
  ylab(paste0("Escherichia coli\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  EnvStats::stat_n_text() +
  stat_compare_means(comparisons = list(c("control", "stroke")),
                     method = "wilcox.test",
                     label = "p.format") +
  coord_fixed(ratio = 2.5)
ggsave("figures/e_coli_class.pdf", device = "pdf", dpi=300, height = 4, width = 4.5)
```

Male stroke patients vs male controls

```{r}
e_coli_stroke %>% 
  filter(sex == "M") %>% 
  ggplot(aes(x = class, y = Abundance, color=class)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) +  
  scale_color_manual(values = cols) + 
  theme_cowplot() + 
  ylab(paste0("Escherichia coli\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  EnvStats::stat_n_text() +
  stat_compare_means(comparisons = list(c("control", "stroke")),
                     method = "wilcox.test",
                     label = "p.format") +
  coord_fixed(ratio = 2.5)
ggsave("figures/e_coli_class_male_only.pdf", device = "pdf", dpi=300, height = 4, width = 4.5)
```

Elderly patients 

```{r}
e_coli_stroke %>% 
  filter(age > 65) %>% 
  ggplot(aes(x = class, y = Abundance, color=class)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) +  
  scale_color_manual(values = cols) + 
  theme_cowplot() + 
  ylab(paste0("Escherichia coli\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  EnvStats::stat_n_text() +
  stat_compare_means(comparisons = list(c("control", "stroke")),
                     method = "wilcox.test",
                     label = "p.format") +
  coord_fixed(ratio = 2.5)
ggsave("figures/e_coli_class_elderly_only.pdf", device = "pdf", dpi=300, height = 4, width = 4.5)
```


Stroke severity and outcome

NIHSS score

```{r}
severity_pal <- c("#F7B267", "#CC3F0C", "#6F3A34")
names(severity_pal) <- c("mild (NIHSS <5)","moderate (NIHSS >5 <15)","severe (NIHSS >15)")
e_coli_sev_plot <- e_coli_stroke %>% 
  mutate(severity = case_when(nihss_in < 5 ~ "mild (NIHSS <5)",
                                              nihss_in >= 5 & nihss_in < 15 ~ "moderate (NIHSS >5 <15)",
                                              nihss_in >= 15 ~ "severe (NIHSS >15)")) %>% 
  mutate(outcome = case_when(mRS_90d == "" ~ NA_real_,
                             mRS_90d %in% 0:2 ~ 0,
                             mRS_90d <= 2 ~ 0,
                             mRS_90d %in% 3:6 ~ 1,
                             mRS_90d > 2 ~ 1,
                             TRUE ~ NA)) %>% 
  filter(class == "stroke")

severity_comps <- make_pairwise_comparisons(e_coli_sev_plot %>% 
  filter(!is.na(severity)), variable = "severity")

e_coli_sev_plot %>% 
  filter(!is.na(severity)) %>% 
  ggplot(aes(x =severity, class, y =Abundance, color=severity)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
  theme_cowplot() + 
  scale_color_manual(values = severity_pal) + 
  ylab(paste0("Escherichia coli\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  EnvStats::stat_n_text() +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  stat_compare_means(comparisons = severity_comps,
                     method = "wilcox.test",
                     label = "p.format") +
  coord_fixed(ratio = 2)
#ggsave("figures/e_coli_severity.pdf", device = "pdf", dpi=300, height = 4, width = 5)
```

```{r}
e_coli_sev_plot %>% 
  filter(!is.na(severity)) %>% 
  filter(sex == "M") %>% 
  ggplot(aes(x =severity, class, y = Abundance, color=severity)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
  theme_cowplot() + 
  scale_color_manual(values=severity_pal) + 
 ylab(paste0("Escherichia coli\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  stat_compare_means(comparisons = severity_comps,
                     method = "wilcox.test",
                     label = "p.format") +
  coord_fixed(ratio = 2)
ggsave("figures/e_coli_severity_male_only.pdf", device = "pdf", dpi=300, height = 4, width = 5)
```

```{r}
e_coli_sev_plot %>% 
  filter(!is.na(severity)) %>% 
  filter(age > 65) %>% 
  ggplot(aes(x =severity, class, y = Abundance, color=severity)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
  theme_cowplot() + 
  scale_color_manual(values=severity_pal) + 
 ylab(paste0("Escherichia coli\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  stat_compare_means(comparisons = severity_comps,
                     method = "wilcox.test",
                     label = "p.format") +
  coord_fixed(ratio = 2)
ggsave("figures/e_coli_severity_elderly_only.pdf", device = "pdf", dpi=300, height = 4, width = 5)
```

Outcome

```{r}
outcome_pal <- c("Favourable"="#104547", "Poor"="#D5573B")


e_coli_sev_plot %>% 
  filter(!is.na(outcome)) %>%
  mutate(outcome =  case_when(outcome == 0 ~ "Favourable",
                             outcome == 1 ~ "Poor")) %>%
  ggplot(aes(x = outcome, class, y = Abundance, color=outcome)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
  theme_cowplot() + 
  scale_color_manual(values = outcome_pal) + 
  ylab(paste0("Escherichia coli\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  stat_compare_means(comparisons = list(c("Favourable", "Poor")),
                     method = "wilcox.test",
                     label = "p.format") +
  EnvStats::stat_n_text() + 
  coord_fixed(ratio = 2.5)
ggsave("figures/e_coli_outcome.pdf", device = "pdf", dpi=300, height = 4, width = 4.5)
```

```{r}
e_coli_sev_plot %>%
  filter(!is.na(outcome)) %>%
  mutate(outcome =  case_when(outcome == 0 ~ "Favourable",
                             outcome == 1 ~ "Poor")) %>%
  ggplot(aes(x = outcome, class, y = Abundance, color=outcome)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) +
  theme_cowplot() + 
  scale_color_manual(values = outcome_pal) + 
  ylab(paste0("Escherichia coli\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None",
    strip.background = element_rect(fill = "white", color = "black")) +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  facet_wrap(~sex) +
  stat_compare_means(comparisons = list(c("Favourable", "Poor")),
                     method = "wilcox.test",
                     label = "p.format") + 
  coord_fixed(ratio = 2.5) + 
  EnvStats::stat_n_text()
ggsave("figures/e_coli_outcome_sex_strat_only.pdf", device = "pdf", dpi=300, height = 4, width = 4.5)
```

```{r}
e_coli_sev_plot %>%
  filter(!is.na(outcome)) %>%
  filter(age > 65) %>% 
  mutate(outcome =  case_when(outcome == 0 ~ "Favourable",
                             outcome == 1 ~ "Poor")) %>%
  ggplot(aes(x = outcome, class, y = Abundance, color=outcome)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) +
  theme_cowplot() + 
  scale_color_manual(values = outcome_pal) + 
  ylab(paste0("Escherichia coli\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  stat_compare_means(comparisons = list(c("Favourable", "Poor")),
                     method = "wilcox.test",
                     label = "p.format") + 
  EnvStats::stat_n_text() + 
  coord_fixed(ratio = 2.5)
ggsave("figures/e_coli_outcome_elderly_only.pdf", device = "pdf", dpi=300, height = 4, width = 4.5)
```

### Tryptophanase

Load data

```{r}
# meta_path <-"../../../Metagenomics/lledos/data/Lledos_metadata.tab"
# 
# humann_genes <- read_tsv("../../../Metagenomics/lledos/data/humann/merged_table/unstrat/humann_genefamilies_ec_unstratified.tsv")
# humann_py <- read_tsv("../../../Metagenomics/lledos/data/humann/merged_table/unstrat/humann_pathabundance_relab_unstratified.tsv")
# 
# current_samples_genes <- humann_genes %>% 
#   select(-`# Gene Family`) %>% 
#   colnames() 
# 
# current_samples_py <- humann_py %>% 
#   select(-`# Pathway`) %>% 
#   colnames() 
# 
# current_samples_genes <- gsub("_merged_clean_Abundance-RPKs", "", current_samples_genes)
# current_samples_py <- gsub("_merged_clean_Abundance*", "", current_samples_py)
# 
# 
# current_samples <- intersect(current_samples_genes, current_samples_py)
# meta_lledos <- read_tsv(meta_path)
# meta_subset <- meta_lledos %>% filter(sample_id %in% current_samples)
# write.table(meta_subset, file = "../../../Metagenomics/lledos/data/Lledos_metadata_subset.tab", sep = "\t", row.names = F, quote = F)
```

```{r}
# meta_subset_path <-  "../../../Metagenomics/lledos/data/Lledos_metadata_subset.tab"
# ec_path <- "../../../Metagenomics/lledos/data/humann/merged_table/unstrat/humann_genefamilies_ec_unstratified.tsv  "
# ko_path <- "../../../Metagenomics/lledos/data/humann/merged_table/unstrat/humann_genefamilies_ko_unstratified.tsv"
# py_path <- "../../../Metagenomics/lledos/data/humann/merged_table/unstrat/humann_pathabundance_relab_unstratified.tsv"
# lledos_func <- import_func_profile(ec = ec_path,ko =  ko_path, py = py_path, meta = meta_subset_path, read=T)
# save(lledos_func, file="../data/lledos_func.RData")
#load("../data/lledos_func.RData")
```

##### Differential abundance

```{r}
load("../data/lledos_func.RData")
# lledos_func_filt <- filter_func(lledos_func, prev = 0.33, abund = 1e-8, renorm = T) %>%
#   transform_func("relative") %>% 
#   transform_func("arcsin")

# # lledos_func_filt_py <- filter_func(lledos_func, prev = 0.05, abund = 1e-8, renorm = T) %>%
# #   transform_func("relative") %>% 
# #   transform_func("arcsin")

# # remove TIA and also create stroke only 
# lledos_func_filt <- subset_samples_func(lledos_func_filt, class %in% c('control', 'stroke'))
# # also add to metadata outcome and severity column
# lledos_func_filt_stroke <-  subset_samples_func(lledos_func_filt, class == 'stroke')
  
# lledos_func_filt_stroke$Metadata <- lledos_func_filt_stroke$Metadata %>% 
#   mutate(outcome = case_when(mRS_90d == "" ~ NA,
#                              mRS_90d %in% 0:2 ~ "Favourable",
#                              mRS_90d <= 2 ~ "Favourable",
#                              mRS_90d %in% 3:6 ~ "Poor",
#                              mRS_90d > 2 ~ "Poor")) %>% 
#   mutate(severity = case_when(nihss_in < 5 ~ "mild (NIHSS <5)",
#                                               nihss_in >= 5 & nihss_in < 15 ~ "moderate (NIHSS >5 <15)",
#                                               nihss_in >= 15 ~ "severe (NIHSS >15)"))



# # also add to metadata outcome and severity column
# # lledos_func_filt_stroke_py <-  subset_samples_func(lledos_func_filt_py, "class == 'stroke'")
  
# # lledos_func_filt_stroke_py$Metadata <- lledos_func_filt_stroke_py$Metadata %>% 
# #   mutate(outcome = case_when(mRS_90d %in% 0:2 ~ "Favourable",
# #                              mRS_90d <= 2 ~ "Favourable",
# #                              mRS_90d %in% 3:6 ~ "Poor",
# #                              mRS_90d > 2 ~ "Poor")) %>% 
# #   mutate(severity = case_when(nihss_in < 5 ~ "mild (NIHSS <5)",
# #                                               nihss_in >= 5 & nihss_in < 15 ~ "moderate (NIHSS >5 <15)",
# #                                               nihss_in >= 15 ~ "severe (NIHSS >15)"))

# severitys <- c("mild (NIHSS <5)","moderate (NIHSS >5 <15)","severe (NIHSS >15)")
# outcomes <- c("Favourable", "Poor")
```

Class - stroke vs control

```{r}
da_func_ec <- maaslin2_func(
  lledos_func_filt,
  prev_thresh = 0,
  abun_thresh = 0,
  fixed = "class",
  feattype = "EC",
  out = "lledos_da_ec",
  cores = 6
)
da_func_ec_df <- da_func_ec$results
```

```{r}
da_func_py <- maaslin2_func(
  lledos_func_filt,
  prev_thresh = 0,
  abun_thresh = 0,
  fixed = "class",
  feattype = "pathways",
  out = "lledos_func_filt",
  cores = 6
)
da_func_py_df <- da_func_py$results
```

```{r}
da_func_ec_sig <- da_func_ec_df %>% 
  filter(qval < 0.05) %>% 
  arrange(coef)
da_func_py_sig <- da_func_py_df %>% 
  filter(qval < 0.05) %>% 
  arrange(coef)
```

Severity

```{r}
da_func_ec_sev <- maaslin2_func(
  subset_samples_func(lledos_func_filt_stroke, "severity %in% c('mild (NIHSS <5)', 'moderate (NIHSS >5 <15)', 'severe (NIHSS >15)')"),
  prev_thresh = 0,
  abun_thresh = 0,
  fixed = "severity",
  feattype = "EC",
  out = "lledos_da_ec_sev",
  reference="severity,mild (NIHSS <5)",
  cores = 6
)
da_func_ec_sev_df <- da_func_ec_sev$results
```

```{r}
da_func_py_sev <- maaslin2_func(
  subset_samples_func(lledos_func_filt_stroke, severity %in% c('mild (NIHSS <5)', 'moderate (NIHSS >5 <15)', 'severe (NIHSS >15)')),
  prev_thresh = 0,
  abun_thresh = 0,
  fixed = "severity",
  feattype = "pathways",
  out = "lledos_da_py_sev",
  reference="severity,mild (NIHSS <5)",
  cores = 6
)
da_func_py_sev_df <- da_func_py_sev$results
```

```{r}
da_func_ec_sev_sig <- da_func_ec_sev_df %>% 
  filter(qval < 0.05) %>% 
  arrange(coef)
da_func_py_sev_sig <- da_func_py_sev_df %>% 
  filter(qval < 0.05) %>% 
  arrange(coef)
```

Outcome

```{r}
da_func_ec_outcome <- maaslin2_func(
  subset_samples_func(lledos_func_filt_stroke, "outcome %in% c('Favourable', 'Poor')"),
  prev_thresh = 0,
  abun_thresh = 0,
  fixed = "outcome",
  feattype = "EC",
  out = "lledos_da_ec_outcome",
  cores = 6
)
da_func_ec_outcome_df <- da_func_ec_outcome$results
```

```{r}
da_func_py_outcome <- maaslin2_func(
  subset_samples_func(lledos_func_filt_stroke_py, "outcome %in% c('Favourable', 'Poor')"),
  prev_thresh = 0,
  abun_thresh = 0,
  fixed = "outcome",
  feattype = "pathways",
  out = "lledos_da_py_outcome",
  cores = 6
)
da_func_py_outcome_df <- da_func_py_outcome$results
```

```{r}
da_func_ec_outcome_sig <- da_func_ec_outcome_df %>% 
  filter(qval < 0.25) %>% 
  arrange(coef)
da_func_py_outcome_sig <- da_func_py_outcome_df %>% 
  filter(qval < 0.25) %>% 
  arrange(coef)
```

##### tnaA outcome

Check why so many are missing outcome here

```{r}
tnaa_outcome <- lledos_func$EC %>% 
  t_df() %>% 
  merge(lledos_func$Metadata, by=0) %>% 
  filter(class != "TIA") %>% 
  filter(bleed != 1)

func_outcome_stroke <- tnaa_outcome %>% 
  filter(class == "stroke") %>% 
    # remove samples missing from tnaA analysis
  filter(Row.names != "s119") %>% 
  mutate(infection_any = case_when(respiratory_infection == 1 ~ 1,
                                   other_infection == 1 ~ 1,
                                   TRUE ~ 0)) %>%
  mutate(outcome = case_when(mRS_90d %in% 0:2 ~ 0,
                             mRS_90d <= 2 ~ 0,
                             mRS_90d %in% 3:6 ~ 1,
                             mRS_90d > 2 ~ 1)) %>% 
  mutate(severity = case_when(nihss_in < 5 ~ "mild (NIHSS <5)",
                                              nihss_in >= 5 & nihss_in < 15 ~ "moderate (NIHSS >5 <15)",
                                              nihss_in >= 15 ~ "severe (NIHSS >15)")) 

tnaa_outcome_stroke <- func_outcome_stroke %>% 
  select(Row.names, `EC:4.1.99.1_Tryptophanase`, age, sex, hypertension, diabetes, nihss_in, severity, dyslipidemia, infection_any, outcome) %>% 
  drop_na() %>% 
  mutate(across(c(sex, hypertension, diabetes, dyslipidemia, infection_any, outcome), as.factor))
```

#### Overview table

Bin tnaA into low and high

```{r}
tnaA_low_high <- quantile(tnaa_outcome_stroke$`EC:4.1.99.1_Tryptophanase`, probs = c(0.5))
tnaa_outcome_stroke <- tnaa_outcome_stroke %>% 
  mutate(tnaA_binned = case_when(`EC:4.1.99.1_Tryptophanase` < tnaA_low_high ~ "low",
                          `EC:4.1.99.1_Tryptophanase` >= tnaA_low_high ~ "high")) %>% 
  mutate(tnaA_binned = factor(tnaA_binned, levels = c("low", "high"))) 
```

```{r}
# Variables to test
variables <- c("tnaA_binned", "age", "sex", "hypertension", "diabetes", 
               "nihss_in", "dyslipidemia", "infection_any")

# Run univariate logistic regression
tnaA_univariate <- lapply(variables, function(var) {
  formula <- as.formula(paste("outcome ~", var))
  glm(formula, data = tnaa_outcome_stroke, family = binomial)
})

tnaA_univariate_summary <- lapply(tnaA_univariate, function(model) {
  tidy(model, conf.int = TRUE, exponentiate = TRUE)
})

# Combine results into a single dataframe
tnaA_univariate_df <- do.call(rbind, tnaA_univariate_summary)
```

```{r}
# Extract clean variable names
tnaA_univariate_df <- tnaA_univariate_df %>%
    filter(term != "(Intercept)") %>% 
  mutate(clean_term = gsub("[0-9]+|[MF]$|TRUE|FALSE", "", term))  # Remove levels, M/F, TRUE/FALSE

# Further customization for specific variables if needed
clean_mapping <- c("tnaA_binnedhigh" = "tnaA (high)","sex" = "Sex", "age" = "Age", "hypertension" = "Hypertension", 
                   "diabetes" = "Diabetes", "nihss_in" = "NIHSS Score Admission", 
                   "dyslipidemia" = "Dyslipidemia", "infection_any" = "Infection (Any)")

tnaA_univariate_df <- tnaA_univariate_df %>%
  mutate(clean_term = recode(clean_term, !!!clean_mapping))
```

```{r}
tnaA_univariate_df %>% 
  mutate(
    significance = case_when(
      p.value < 0.001 ~ paste0(round(estimate, digits = 2), " ***"),
      p.value < 0.01 ~  paste0(round(estimate, digits = 2), " **"),
      p.value < 0.05 ~  paste0(round(estimate, digits = 2), " *"),
      TRUE ~ paste0(round(estimate, digits = 2), " "),
    ),
    term = factor(term, levels = term)  # Maintain the order of variables
  ) %>% 
  ggplot(aes(x = clean_term, y = estimate)) +
  geom_point(size = 3, shape=15, aes(color=estimate > 1)) +  # Points for odds ratios
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.15) +
  geom_text(aes(label = significance), nudge_x = 0.4, hjust=0.5, size = 4, color = "black") +# Error bars
  scale_y_log10() +  # Log scale for odds ratios
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Univariate regression (n = 148)",
    x ="",
    y = "Odds Ratio (log scale)"
  ) +
  theme_cowplot() + 
  scale_color_manual(values=c("dodgerblue", "firebrick1")) +
  theme(axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.position = "None") + 
  geom_hline(yintercept = 1, linetype="dashed",color="lightgrey")
ggsave("figures/tnaA_univariate.pdf", device = "pdf", dpi=300, height = 5, width = 6)
```

```{r}
multivariate_model <- glm(outcome ~ tnaA_binned + age + sex + hypertension + diabetes + 
             nihss_in + dyslipidemia + infection_any, 
             data = tnaa_outcome_stroke, family = binomial)
#tidy(multivariate_model, conf.int = TRUE, exponentiate = TRUE)
 
multivariate_model %>% 
  tbl_regression(exponentiate = T, label = as.list(clean_mapping), show_single_row = c("sex", "hypertension", "diabetes", "dyslipidemia", "infection_any")) %>%
  bold_p(t = 0.05) %>% 
  as_gt() %>% 
  gt::gtsave(filename = "multivariate_log_regression_tnaA.docx")
```


#### E. coli and tnaA univariate

```{r}
tnaa_e_coli_outcome <-  e_coli_outcome %>%
  left_join(tnaa_outcome_stroke %>% select(Row.names, tnaA_binned), by = c("Sample" = "Row.names"))
```

```{r}
# Variables to test
variables <- c("s__Escherichia_coli", "tnaA_binned", "age", "sex", "hypertension", "diabetes", 
               "nihss_in", "dyslipidemia", "infection_any")

# Run univariate logistic regression
tnaa_e_coli_univariate <- lapply(variables, function(var) {
  formula <- as.formula(paste("outcome ~", var))
  glm(formula, data = tnaa_e_coli_outcome, family = binomial)
})


tnaa_e_coli_univariate_summary <- lapply(tnaa_e_coli_univariate, function(model) {
  tidy(model, conf.int = TRUE, exponentiate = TRUE)
})

# Combine results into a single dataframe
tnaa_e_coli_univariate_df <- do.call(rbind, tnaa_e_coli_univariate_summary)

```

```{r}
# Extract clean variable names
tnaa_e_coli_univariate_df <- tnaa_e_coli_univariate_df %>%
  filter(term != "(Intercept)") %>%
  mutate(clean_term = gsub("[0-9]+|[MF]$|TRUE|FALSE", "", term)) # Remove levels, M/F, TRUE/FALSE

# Further customization for specific variables if needed
clean_mapping <- c(
  "s__Escherichia_coli" = "Escherichia coli",
  "tnaA_binnedhigh" = "tnaA (high)",
  "sex" = "Sex (M)",
  "age" = "Age",
  "hypertension" = "Hypertension",
  "diabetes" = "Diabetes",
  "nihss_in" = "NIHSS Score Admission",
  "dyslipidemia" = "Dyslipidemia",
  "infection_any" = "Infection (Any)"
)

tnaa_e_coli_univariate_df <- tnaa_e_coli_univariate_df %>%
  mutate(clean_term = recode(clean_term, !!!clean_mapping))
```

```{r}
tnaa_e_coli_univariate_df %>% 
  mutate(
    significance = case_when(
      p.value < 0.001 ~ paste0(round(estimate, digits = 2), " ***"),
      p.value < 0.01 ~  paste0(round(estimate, digits = 2), " **"),
      p.value < 0.05 ~  paste0(round(estimate, digits = 2), " *"),
      TRUE ~ paste0(round(estimate, digits = 2), " "),
    )
  ) %>% 
  ggplot(aes(x = fct_relevel(clean_term, rev(c("Escherichia coli", "tnaA (high)", "Age", "Diabetes", "Dyslipidemia", "Hypertension", "Infection (Any)","NIHSS Score Admission", "Sex (M)"))), y = estimate)) +
  geom_point(size = 3, shape=15, aes(color=estimate > 1)) +  # Points for odds ratios
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.15) +
  geom_text(aes(label = significance), nudge_x = 0.4, hjust=0.5, size = 4, color = "black") +# Error bars
  scale_y_log10() +  # Log scale for odds ratios
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Univariate regression (n = 148)",
    x ="",
    y = "Odds Ratio (log scale)"
  ) +
  theme_cowplot() + 
  scale_color_manual(values=c("dodgerblue", "firebrick1")) +
  theme(axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.position = "None") + 
  geom_hline(yintercept = 1, linetype="dashed",color="lightgrey")
ggsave("figures/E_coli_tnaA_univariate.pdf", device = "pdf", dpi=300, height = 5, width = 6)
```

#### Mediation analysis 

Escherichia coli 
```{r}
library(mediation)


```


#### Interaction 

For each

Add elderly age variable to data
```{r}
tnaa_outcome_stroke <- tnaa_outcome_stroke %>% 
  mutate(elderly = case_when(age >= 65 ~ 1,
                            age < 65 ~ 0)) 
```
```{r}
# Variables to test
risk_factors <- c("elderly", "hypertension", "diabetes", "dyslipidemia")

# Run univariate logistic regression
tnaA_univariate_interaction <- lapply(risk_factors, function(var) {
  formula <- as.formula(paste("outcome ~ tnaA_binned * ", var))
  glm(formula, data = tnaa_outcome_stroke, family = binomial)
})

tnaA_univariate_interaction_summary <- lapply(tnaA_univariate_interaction, function(model) {
  tidy(model, conf.int = TRUE, exponentiate = TRUE)
})

# Combine results into a single dataframe
tnaA_univariate_interaction_df <- do.call(rbind, tnaA_univariate_interaction_summary) %>%
  mutate(
    p.adj = p.adjust(p.value, method = "BH")
  )
```

Forest plots
```{r}
tnaA_univariate_interaction_df %>%
    filter(term != "(Intercept)") %>% 
    filter(term != "tnaA_binnedhigh") %>%
    filter(grepl("tnaA_binnedhigh:", term)) %>%
  mutate(clean_term = recode(term, !!!clean_mapping)) %>% 
  mutate(
    significance = case_when(
      p.value < 0.001 ~ paste0(round(estimate, digits = 2), " ***"),
      p.value < 0.01 ~  paste0(round(estimate, digits = 2), " **"),
      p.value < 0.05 ~  paste0(round(estimate, digits = 2), " *"),
      TRUE ~ paste0(round(estimate, digits = 2), " "),
    ),
    term = factor(term, levels = term)  # Maintain the order of variables
  ) %>% 
  ggplot(aes(x = clean_term, y = estimate)) +
  geom_point(size = 3, shape=15, aes(color=estimate > 1)) +  # Points for odds ratios
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.15) +
  geom_text(aes(label = significance), nudge_x = 0.4, hjust=0.5, size = 4, color = "black") +# Error bars
  scale_y_log10() +  # Log scale for odds ratios
  coord_flip() +  # Flip coordinates for better readability
  labs(
    title = "Univariate regression (n = 148)",
    x ="",
    y = "Odds Ratio (log scale)"
  ) +
  theme_cowplot() + 
  scale_color_manual(values=c("dodgerblue", "firebrick1")) +
  theme(axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.position = "None") + 
  geom_hline(yintercept = 1, linetype="dashed",color="lightgrey")
```



Marginal effect plots 

```{r}
library(ggeffects)
library(patchwork)

marginal_plots <- lapply(risk_factors, function(rf) {
  
  # Refit the model
  formula_str <- paste0("outcome ~ tnaA_binned * ", rf)
  model <- glm(as.formula(formula_str), 
               family = binomial(),
               data = tnaa_outcome_stroke)
  
  # Generate marginal effects
  pred <- ggpredict(model, terms = c("tnaA_binned [all]", rf))
  
  # Plot
  plot(pred, log_y=T) +
    labs(
      title = tools::toTitleCase(rf),
      x = "tnaA_binned",
      y = "Predicted Probability",
      color = rf
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
})

# Combine
wrap_plots(marginal_plots, ncol = 2) 
```

Check additional indole genes for association

```{r}
indole_genes <- read_tsv("../../AhR_metagenomics/data/indole_metabolism_ko_ec.txt")
ec_ind_gene <- paste0("EC:", indole_genes$EC, "_")
```

```{r}
indole_outcome_stroke <-  func_outcome_stroke %>% 
  select(contains(ec_ind_gene), Row.names, age, sex, hypertension, diabetes, nihss_in, severity, dyslipidemia, infection_any, outcome) %>% 
  drop_na() %>% 
  mutate(across(c(sex, hypertension, diabetes, dyslipidemia, infection_any, outcome), as.factor))
```

```{r}
ec_ind_gene <- colnames(indole_outcome_stroke %>% select(contains("EC", ignore.case = F)))
# bin genes into low and high
indole_low_high <- map(ec_ind_gene, function(x) quantile(indole_outcome_stroke[[x]], probs = c(0.5))) 
names(indole_low_high) <- ec_ind_gene

indole_outcome_stroke <- indole_outcome_stroke %>% 
  mutate(across(all_of(ec_ind_gene), ~ case_when(. <= indole_low_high[[cur_column()]] ~ "low",
                                                  . > indole_low_high[[cur_column()]] ~ "high"))) %>% 
  mutate(across(all_of(ec_ind_gene), factor, levels = c("low", "high")))

# Remove columns with single value e.g. all 0 
indole_outcome_stroke <- indole_outcome_stroke %>% 
   select(where(~ n_distinct(.x) > 1)) 
  
ec_ind_gene_reg <- paste0("`", colnames(indole_outcome_stroke %>% select(contains("EC", ignore.case = F))), "`")

# Run univariate logistic regression
indole_univariate <- lapply(ec_ind_gene_reg, function(var) {
  formula <- as.formula(paste("outcome ~", var))
  glm(formula, data = indole_outcome_stroke, family = binomial)
})

indole_univariate_summary <- lapply(indole_univariate, function(model) {
  tidy(model, conf.int = TRUE, exponentiate = TRUE)
})
# Combine results into a single dataframe
indole_univariate_df <- do.call(rbind, indole_univariate_summary)
indole_univariate_df <- indole_univariate_df %>%
  filter(term != "(Intercept)") %>% 
  mutate(clean_term = gsub("[0-9]+|[MF]$|TRUE|FALSE", "", term))
```

##### Group comparisons

```{r}
tnaa_outcome %>% 
  ggplot(aes(x = class, y = `EC:4.1.99.1_Tryptophanase`, color=class)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) +  
  scale_color_manual(values = cols) + 
  theme_cowplot() + 
  ylab(paste0("tnaA\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  EnvStats::stat_n_text() +
  stat_compare_means(comparisons = list(c("control", "stroke")),
                     method = "wilcox.test",
                     label = "p.format") 
ggsave("figures/tnaa_class.pdf", device = "pdf", dpi=300, height = 4, width = 4.5)
```

Male stroke patients vs male controls

```{r}
tnaa_outcome %>% 
  filter(sex == "M") %>% 
  ggplot(aes(x = class, y = `EC:4.1.99.1_Tryptophanase`, color=class)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
  scale_color_manual(values = cols) + 
  theme_cowplot() + 
  ylab(paste0("tnaA\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") 
  stat_compare_means(comparisons = list(c("control", "stroke")),
                     method = "wilcox.test",
                     label = "p.format")
ggsave("figures/tnaa_class_male_only.pdf", device = "pdf", dpi=300, height = 5, width = 6)
```

Stroke severity and outcome

NIHSS score

```{r}
severity_comps <- make_pairwise_comparisons(tnaa_outcome_stroke, variable = "severity")

tnaa_outcome_stroke %>% 
  ggplot(aes(x =severity, class, y = `EC:4.1.99.1_Tryptophanase`, color=severity)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) +  
  theme_cowplot() + 
  scale_color_manual(values = severity_pal) + 
  ylab(paste0("tnaA\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
    EnvStats::stat_n_text() +
  stat_compare_means(comparisons = severity_comps,
                     method = "wilcox.test",
                     label = "p.format") 
ggsave("figures/tnaa_severity.pdf", device = "pdf", dpi=300, height = 4, width = 5)
```

```{r}
tnaa_outcome_stroke %>% 
  filter(sex == "M") %>% 
  ggplot(aes(x =severity, class, y = `EC:4.1.99.1_Tryptophanase`, color=severity)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) +  
  theme_cowplot() + 
  scale_color_manual(values = severity_pal) +  
  ylab(paste0("tnaA\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
    EnvStats::stat_n_text() +
  stat_compare_means(comparisons = severity_comps,
                     method = "wilcox.test",
                     label = "p.format") 
ggsave("figures/tnaa_severity_male_only.pdf", device = "pdf", dpi=300, height = 4, width = 5)
```

Outcome

```{r}
tnaa_outcome_stroke %>% 
  mutate(outcome =  case_when(outcome == 0 ~ "Favourable",
                             outcome == 1 ~ "Poor")) %>%
  ggplot(aes(x = outcome, class, y = `EC:4.1.99.1_Tryptophanase`, color=outcome)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
  theme_cowplot() + 
  scale_color_manual(values = outcome_pal) +
  ylab(paste0("tnaA\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
      EnvStats::stat_n_text() +
  stat_compare_means(comparisons = list(c("Favourable", "Poor")),
                     method = "wilcox.test",
                     label = "p.format")
ggsave("figures/tnaa_outcome.pdf", device = "pdf", dpi=300, height = 4, width = 4.5)
```

```{r}
tnaa_outcome_stroke %>%
  mutate(outcome =  case_when(outcome == 0 ~ "Favourable",
                             outcome == 1 ~ "Poor")) %>%
  ggplot(aes(x = outcome, class, y = `EC:4.1.99.1_Tryptophanase`, color=outcome)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
  theme_cowplot() + 
  scale_color_manual(values = outcome_pal) + 
  ylab(paste0("tnaA\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None",
    strip.background = element_rect(fill = "white", color = "black")) +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  facet_wrap(~sex) + 
  stat_compare_means(comparisons = list(c("Favourable", "Poor")),
                     method = "wilcox.test",
                     label = "p.format")
ggsave("figures/tnaa_outcome_sex_strat.pdf", device = "pdf", dpi=300, height = 5, width = 6)
```

## Question 4. Are indole-producers/tryptophanase enzymes still altered in chronic phase?

Load and combine data from SMB and PRISE 

```{r}
load("../data/SMB_tax.RData")
load("../data/PRISE_tax.RData")
```

```{r}
ps_smb_norm <- ps_smb %>% transform()
ps_prise_norm <- ps_prise %>% transform()
```

```{r}
ps_chronic <- merge_phyloseq(ps_smb_norm, ps_prise_norm) %>% 
  subset_samples(timepoint == "chronic" | class == "control") 
```

##### E. coli

```{r}
df_indole_producers_chronic <- ps_chronic %>%
  transform("arcsin") %>% 
  psmelt() %>% 
  filter(grepl(paste(indole_species, collapse = "|"), Species)) %>% 
  mutate(OTU = gsub("species:", "", OTU)) %>% 
  filter(class != "TIA") 

e_coli_stroke_chronic <- df_indole_producers_chronic %>%
  filter(Species == "s__Escherichia_coli") %>% 
  mutate(group2 = case_when(timepoint == "chronic" ~ "Chronic stroke",
                            class == "control" ~ "Control"))
```



```{r}
chronic_stroke_cols <- c("Chronic stroke" = "#74A57F", "Control" = "#424143")
e_coli_stroke_chronic %>% 
  ggplot(aes(x = fct_relevel(group2, "Control", "Chronic stroke"), y = Abundance, color=group2)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) + 
  scale_color_manual(values = chronic_stroke_cols) + 
  theme_cowplot() + 
  ylab(paste0("Escherichia coli\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  EnvStats::stat_n_text() +
  stat_compare_means(comparisons = list(c("Control", "Chronic stroke")),
                     method = "wilcox.test",
                     label = "p.format") 
ggsave("figures/e_coli_chronic_stroke.pdf", device = "pdf", dpi=300, height = 4, width = 4.5)
```

Is E. coli still associated with stroke outcome in chronic phase?

```{r}
e_coli_stroke_chronic_outcome <- e_coli_stroke_chronic %>% 
  mutate(outcome = case_when(mRS_90d %in% 0:2 ~ "Favourable",
                             mRS_90d <= 2 ~ "Favourable",
                             mRS_90d %in% 3:6 ~ "Poor",
                             mRS_90d > 2 ~ "Poor")) %>% 
  filter(!is.na(outcome))
glm(Abundance ~ outcome, data = e_coli_stroke_chronic_outcome, family = binomial) %>% 
  tidy(exponentiate = T, conf.int = T) 
```
##### Tryptophanase

```{r}
load("../data/smb_func.RData")
load("../data/prise_func.RData")

smb_func$Metadata <- smb_func$Metadata %>% 
  mutate(antibiotics = case_when(antibiotics == "Yes" ~ 1,
                             antibiotics == "No" ~ 0))
```

```{r}
tmp1 <- smb_func$EC %>% 
  merge(smb_func$Metadata, by=0) %>% 
  filter(class != "TIA") %>% 
  filter(timepoint == "chronic" | class == "control")

tmp2 <- prise_func$EC %>% 
  merge(prise_func$Metadata, by=0)

tnaa_outcome_chronic <- bind_rows(tmp1, tmp2) %>%
  mutate(outcome = case_when(class == "control" ~ NA,
                             mRS_90d %in% 0:2 ~ "Favourable",
                             mRS_90d <= 2 ~ "Favourable",
                             mRS_90d %in% 3:6 ~ "Poor",
                             mRS_90d > 2 ~ "Poor")) %>% 
  select(Row.names, `EC:4.1.99.1_Tryptophanase`, age, sex,timepoint, hypertension, nihss_in, dyslipidemia, outcome) %>% 
  mutate(across(c(sex, hypertension, dyslipidemia, outcome), as.factor)) %>% 
  mutate(group2 = case_when(timepoint == "chronic" ~ "Chronic stroke",
                            TRUE ~ "Control"))

```

```{r}
tnaa_outcome_chronic %>% 
  ggplot(aes(x = fct_relevel(group2, "Control", "Chronic stroke"), y = `EC:4.1.99.1_Tryptophanase`, color=group2)) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15), size=3.75, alpha=0.4) +  
  scale_color_manual(values = chronic_stroke_cols) + 
  theme_cowplot() + 
  ylab(paste0("tnaA\n", "Asin(sqrt(rel. abundance))")) +
  xlab("") +
  theme(axis.title = element_text(size=14),
        axis.text = element_text(size=12),
        legend.position = "None") +
  stat_summary(fun.y = median, 
               fun.ymin = median, 
               fun.ymax = median, 
               geom = "crossbar", 
               width = 0.15, color="black") + 
  EnvStats::stat_n_text() +
  stat_compare_means(comparisons = list(c("Control", "Chronic stroke")),
                     method = "wilcox.test",
                     label = "p.format") 
ggsave("figures/tnaA_chronic_stroke.pdf", device = "pdf", dpi=300, height = 4, width = 4.5)
```

`
```{r}
glm(`EC:4.1.99.1_Tryptophanase` ~ outcome, data = tnaa_outcome_chronic, family = binomial) %>% 
  tidy(exponentiate = T, conf.int = T)
```

